### 2024-01-11 abstract.qmd
add frontmatter

---
title: A mixture of hidden Markov models to predict the lymphatic spread on head and neck cancer
author:
  - name: Roman Ludwig
    orcid: 0000-0001-9434-328X
    email: roman.ludwig@usz.ch
    affiliations:
      - ref: 1
      - ref: 2
  - name: Julian Brönnimann
    affiliations:
      - ref: 1
      - ref: 2
  - name: Jan Unkelbach
    orcid: 0000-0002-4275-990X
    affiliations:
      - ref: 1
      - ref: 2
affiliation:
  - id: 1
    name: University of Zurich
    departement: Departement of Physics
    city: Zurich
    country: Switzerland
  - id: 2
    name: University Hospital Zurich
    departement: Departement of Radiation Oncology
    city: Zurich
    country: Switzerland
bibliography: references.bib
abstract: |
  We previously developed a mechanistic hidden Markov model (HMM) to predict the
  lymphatic tumor progression in oropharyngeal squamous cell carcinomas (OPSCCs). To
  extend the model to other tumor subsites, defined by ICD-10 codes, in the head and
  neck, we employ a mixture of these HMMs and learn the cluster assignments and model
  parameters in an iterative, EM-like algorithm from multi-centric data. The mixture
  model manages to cluster anatomically close subsites and correctly infers the
  clusters' parameters. Using this mixture model allows the prediction of individual
  risks of occult nodal disease, given a diagnosis that includes tumor subsite.
---

# Introduction

# Materials and Methods

# Results

# Discussion

### 2024-01-11 abstract.qmd
write large part of introduction

@@
 ---
-title: A mixture of hidden Markov models to predict the lymphatic spread on head and neck cancer
+title: A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer
 author:
@@
       - ref: 2
-affiliation:
+affiliations:
   - id: 1
@@
   risks of occult nodal disease, given a diagnosis that includes tumor subsite.
+format:
+  iccr-pdf: default
 ---
@@
 
-# Materials and Methods
+Current diagnostic imaging modalities are unable to detect microscopic nodal metastases, which requires pathological examination of extracted tissue [@snyder_petct_2021;@strohl_petct_2021].
+However, the presence of nodal disease is detrimental to a patient's outcome [@jones_level_1994;@lim_distributions_2006;@takes_staging_2004].
+Thus, large volumes in the head and neck region are often irradiated electively to minimize the risk of missing occult disease.
+However, the decision guidelines [@vorwerk_guidelines_2011;@gregoire_delineation_2018;@biau_selection_2019] which nodal regions -- i.e., anatomically defined lymph node levels (LNLs) -- to irradiate are not based on a patient's individual risk, but only on the overall prevalence of nodal disease as reported in the literature [@woolgar_topography_2007;@candela_patterns_1990;@chao_determination_2002;@shah_patterns_1990].
 
-# Results
+To personalize this prediction of the risk for occult diease, given a patient's individual diagnosis, we published
+
+1. large, multi-centric data that details per patient which LNLs where clinically and/or pathologically (following neck dissection) involved [@ludwig_dataset_2022;@ludwig_multi-centric_2023].
+
+And, building on this work,
+
+2. an interpretable hidden Markov model (HMM), trained with this data, to predict the risk for occult nodal disease [@ludwig_hidden_2021;@ludwig_modelling_2023].
+
+Such a personalized risk prediction may allow clinitians to safely reduce the elective clinical target volume (CTV-N) and thus reduce side-effects that degrade the patient's quality of life [@batth_practical_2014].
+
+HNSCC patients with primary tumors at different subsites, e.g.
+in the oral cavity and in the oropharynx, also show different patterns of lymphatic spread.
+To capture this in a risk prediction using our previously developed model, one would need to train entirely separate models for each subsite, based on separated datasets.
+We consider this a limitation for two reasons:
+
+1.
+While the spread from the tumor to the LNLs may differ for different primary tumor subsites, we would assume the spread from involved LNLs further down the lymphatic network to still be similar.
+Thus, a dataset of oral cavity patients may contain information that is relevant to the oropharynx model and vice versa.
+2.
+A tumor may lie anatomically between two subsites.
+Then, it makes sense to assume that this tumor spreads in a way that can be described as a combination of the typical spread patterns of these two subsites, given they are close enough to each other.
+This presents an opportunity to reduce the model's complexity: If a subsite's lymphatic spread can be described as a mixture of two others' spread patterns, we only require two distinct HMMs to describe three tumor subsites.
+
+In the work presented here, we implement a mixture of HMMs that aims to address these limitations.
+In @sec-materials-and-methods we first introduce the basic concept of the HMM, followed by a general description of mixture models and how we implemented them in our particular case.
+
+# Materials and Methods {#sec-materials-and-methods}
+
+# Results {#sec-results}
 

### 2024-01-15 content/manuscript.qmd
write about mixture model

---
title: A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer
author:
  - name: Roman Ludwig
    orcid: 0000-0001-9434-328X
    email: roman.ludwig@usz.ch
    affiliations:
      - ref: 1
      - ref: 2
  - name: Julian Brönnimann
    affiliations:
      - ref: 1
      - ref: 2
  - name: Jan Unkelbach
    orcid: 0000-0002-4275-990X
    affiliations:
      - ref: 1
      - ref: 2
affiliations:
  - id: 1
    name: University of Zurich
    departement: Departement of Physics
    city: Zurich
    country: Switzerland
  - id: 2
    name: University Hospital Zurich
    departement: Departement of Radiation Oncology
    city: Zurich
    country: Switzerland
bibliography: ../references.bib
abstract: |
  We previously developed a mechanistic hidden Markov model (HMM) to predict the
  lymphatic tumor progression in oropharyngeal squamous cell carcinomas (OPSCCs). To
  extend the model to other tumor subsites, defined by ICD-10 codes, in the head and
  neck, we employ a mixture of these HMMs and learn the cluster assignments and model
  parameters in an iterative, EM-like algorithm from multi-centric data. The mixture
  model manages to cluster anatomically close subsites and correctly infers the
  clusters' parameters. Using this mixture model allows the prediction of individual
  risks of occult nodal disease, given a diagnosis that includes tumor subsite.
format:
  iccr-pdf:
    keep-tex: true
    include-in-header:
      text: "\\usepackage{newtxtext,newtxmath}"
crossref:
  fig-prefix: "figure"
  tbl-prefix: "table"
  eq-prefix: "equation"
  sec-prefix: "section"
---

# Introduction

Head and neck squamous cell carcinomas (HNSCC) frequently spread through the lymphatic system [@lindberg_distribution_1972;@candela_patterns_1990;@woolgar_histological_1999]. Current diagnostic imaging modalities are unable to detect microscopic nodal metastases, which requires pathological examination of extracted tissue [@snyder_petct_2021;@strohl_petct_2021]. Since the recurrence of nodal disease is detrimental to a patient's outcome [@ho_decision_2014], large volumes in the head and neck region are often irradiated electively to minimize the risk of missing occult disease. Decision guidelines about which nodal regions -- i.e., anatomically defined lymph node levels (LNLs) -- to irradiate [@vorwerk_guidelines_2011;@gregoire_delineation_2018;@biau_selection_2019] are currently not based on a patient's individual risk, but only on the overall prevalence of nodal disease as reported in the literature [@lindberg_distribution_1972;@candela_patterns_1990;@woolgar_histological_1999;@chao_determination_2002].

To personalize this prediction of the risk for occult diease, given a patient's individual diagnosis, we published

1. large, multi-centric data that reports per patient which LNLs where clinically and/or pathologically involved [@ludwig_dataset_2022;@ludwig_multi-centric_2023].

And, building on this work,

2. an interpretable hidden Markov model (HMM), trained with this data, to predict the risk for occult nodal disease [@ludwig_hidden_2021;@ludwig_modelling_2023], given an individual patient's diagnosis.

Such a personalized risk prediction may allow clinitians to safely reduce the elective clinical target volume (CTV-N) and thus reduce side-effects that degrade the patient's quality of life [@batth_practical_2014].

HNSCC patients with primary tumors at different subsites, e.g. in the oral cavity and in the oropharynx, also show different patterns of lymphatic spread [@lindberg_distribution_1972;@candela_patterns_1990;@woolgar_histological_1999]. To naively capture this in a risk prediction using our previously developed model, one would need to train entirely separate models for each subsite, based on separated datasets. We consider this a limitation for two reasons:

1. While the spread from the tumor to the LNLs may differ for different primary tumor subsites, we would assume the spread from involved LNLs further down the lymphatic network to still be similar. Thus, a dataset of oral cavity patients may contain information that is relevant to the oropharynx model and vice versa.
2. A tumor may lie anatomically between two subsites. Then, it makes sense to assume that this tumor spreads in a way that can be described as a combination of the typical spread patterns of these two subsites, given they are close enough to each other. This presents an opportunity to reduce the model's complexity: If a subsite's lymphatic spread can be described as a mixture of two others' spread patterns, we only require two distinct HMMs to describe three tumor subsites.

In the work presented here, we implement a mixture of HMMs that aims to address these limitations. In @sec-materials-and-methods we first introduce the basic concept of the HMM (@sec-hmm), followed by a general description of mixture models (@sec-mixture-of-hmms) and how we implemented them in our particular case.

# Materials and Methods {#sec-materials-and-methods}

To compute the personalized risk of occult disease $\mathbf{X}$, given a diagnosis $\mathbf{Y}$, we can begin by stating Bayes' law:

$$
P \left( \mathbf{X} \mid \mathbf{Y} \right) = \frac{P \left( \mathbf{Y} \mid \mathbf{X} \right) P \left( \mathbf{X} \right)}{\sum_{\mathbf{X}^\star} P \left( \mathbf{Y} \mid \mathbf{X}^\star \right) P \left( \mathbf{X}^\star \right)}
$$ {#eq-bayes-law}

In the above equation, the term $P \left( \mathbf{Y} \mid \mathbf{X} \right)$ is given by the sensitivity and specificity of the diagnosis. The crucial term that our model attempts to compute, is the prior probability of involvement $P \left( \mathbf{X} \right)$ (or rather of the hidden state of involvement, see the [next section](#sec-hmm)).

## Hidden Markov Model for Lymphatic Progression {#sec-hmm}

Each LNL $v \in \{ 1, 2, \ldots, V\}$ considered in our model is represented by a hidden binary random variable (RV) $X_v[t]$ representing the true -- but not directly observable -- state of that level at the abstract time-step $t$. It may take on the values "healthy" (0) or "involved" (1). Collected in a random vector $\mathbf{X}[t] = \left( X_1[t], X_2[t], \ldots, X_V[t] \right)$ they form the patient's hidden state w.r.t. their lymphatic involvement at time $t$.

When a patient gets diagnosed, their LNLs' involvement status is assessed using a clinical modality $\mathcal{O}$ with its repsective sensitivity $s_N^\mathcal{O}$ and specificity $s_P^\mathcal{O}$. This involvement status, too, is modelled as an RV $Y_v^\mathcal{O}$. Hence, a diagnosis obtained with diagnostic modality $\mathcal{O}$ can be collected in the random vector $\mathbf{Y}^\mathcal{O}$.

![Example of a parametrized graph representation of a small lymphatic network consisting of only three LNLs. Blue, round nodes represent the hidden RVs, while the orange square nodes show the observed RVs. Every arc represents a conditional probability parametrized with the quantity noted next to it.](../static/graph_with_obs.png){#fig-graph-with-obs width="25%"}

We model the process of tumor progression via lymphatic drainage by connecting the hidden RVs in a graph, as shown in @fig-graph-with-obs. The arcs in [this graph](#fig-graph-with-obs) represent conditional probabilities. The orange arcs, as mentioned, correspond to observing a diagnosis $Z_v^\mathcal{O}$, given a hidden state $X_v$. The values this conditional probability $P(Y_v \mid X_v)$ can take on are tabulated in @tbl-sens-spec.

+---------+---------+---------+
|         | $X_v=0$ | $X_v=1$ |
+=========+:=======:+:=======:+
| $Y_v=0$ | $s_P$   | $1-s_N$ |
+---------+---------+---------+
| $Y_v=1$ | $1-s_P$ | $s_N$   |
+---------+---------+---------+

: Table encoding the conditional probability $P(Y_v \mid X_v)$ to correctly (or wrongly) diagnose an LNL's true hidden state. This table is associated with the orange arcs in @fig-graph-with-obs. {#tbl-sens-spec}

Similarily, the red arcs in @fig-graph-with-obs depict the probability that the primary tumor spreads within one abstract time-step. Lastly, the blue arcs symbolize the spread from an upstream LNL -- given it is already metastatic -- to a downstream level. For example, the edge from $X_2$ to $X_3$ encodes the probability $P\left( X_3[t+1] \mid X_2[t] \right)$, which is parametrized with $b_3$, and $t_{2 \rightarrow 3}$ and tabulated in @tbl-spread-2-to-3. There is an additional restriction on any LNL's state $X_v[t+1]=0$ to be healthy: It requires that the level was also healthy in the time-step before, meaning $X_v[t]=0$. This is because we assume no spontaneous self-healing of metastatic levels.

+--------------+------------+-----------------------------------------------------+
|              | $X_2[t]=0$ | $X_2[t]=1$                                          |
+==============+:==========:+:===================================================:+
| $X_3[t+1]=0$ | $1-b_3$    | $(1-b_3)(1-t_{2 \rightarrow 3})$                    |
+--------------+------------+-----------------------------------------------------+
| $X_3[t+1]=1$ | $b_3$      | $1-b_3-t_{2 \rightarrow 3}+b_3 t_{2 \rightarrow 3}$ |
+--------------+------------+-----------------------------------------------------+

: Conditional probability $P\left( X_3[t+1] \mid X_2[t] \right)$ for a spread from LNL II to III during the transition from $t$ to $t+1$. This corresponds to one of the blue arcs in @fig-graph-with-obs. {#tbl-spread-2-to-3 tbl-colwidths="[25,18,40]"}

With the introduced conditional probabilities, we can now compute the joint probability of any complete hidden state $\mathbf{X}[t]=\boldsymbol{\xi}_i$ transitioning to any other possible state $\mathbf{X}[t+1]=\boldsymbol{\xi}_j$. Here, the $i$ and $j$ enumerate all $2^V$ combinations of the $V$ binary RVs. In the graph shown in @fig-graph-with-obs, this amounts to $2^V=8$ distinct $\boldsymbol{\xi}_i$. Because these terms are essentially products of terms like those in @tbl-spread-2-to-3. We can then collect these terms in a _transition matrix_ $\mathbf{A}$:

$$
\mathbf{A} = \begin{pmatrix} A_{ij} \end{pmatrix} = P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right)
$$ {#eq-transition-matrix}

Note that this matrix still depends on the $b_v$ and $t_{r \rightarrow v}$ parameters, although we have dropped the explicit dependcy to keep the equations brief. Now, assuming that every patient started their disease with all LNLs being healthy, we can define the _starting distribution_ $\boldsymbol{\pi}$:

$$
\boldsymbol{\pi} = \begin{pmatrix} \pi_i \end{pmatrix} = P \left( \mathbf{X}[0] = \boldsymbol{\xi}_i \right)
$$ {#eq-starting-distribution}

And set every entry of this starting distribution to zero, except the first entry, which we set to one. This means at $t=0$ there is a probability of one to be in the completly healthy state $\boldsymbol{\xi}_0 = \begin{pmatrix} 0, 0, \ldots, 0 \end{pmatrix}$.

Using the quantities introduced so far, the probability distribution vector with elements $P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right)$ after $t$ time-steps can now be conveniently expressed as a matrix product:

$$
P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right) = \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
$$ {#eq-evolution}

This evolution implicitly marginalizes over all possible paths to arrive at state $\boldsymbol{\xi}_i$ after $t$ time-steps. Additionally, we also need to marginalize over the time of diagnosis -- which is unknown -- using a time-prior $P_T(t)$. Fortunately, the exact length and shape of this distribution on its own has little impact. But because we assume that early and advanced T-category patients are fundamentally the same, just on average diagnosed at different times $t$, we use the time-prior only to separate the respective patient's evolutions:

$$
P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid T \right) = \sum_{t=0}^{t_\text{max}} P_T(t) \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
$$ {#eq-marginalized-evolution}

where $T \in \{ \text{early}, \text{advanced} \}$ denotes the T-category. We fix $P_\text{early}(t)$ to a binomial distribution with parameters $n=t_\text{max}=$ {{< var t_max >}} and $p_\text{early}=$ {{< var time-prior.early.p >}}, while the advanced T-category's time-prior is also a binomial distribution where the $p_\text{advanced}$ parameter is learned.

Note that [the above equation](#eq-marginalized-evolution) still depends on the parametrization of the transition matrix. We collect these parameters in a vector $\boldsymbol{\theta}=\{ b_v, t_{r \rightarrow v}, p_\text{advanced} \}$.

To train our model, we need to compute the likelihood of a dataset $\mathcal{D}=\left( \mathbf{Y}_1, \mathbf{Y}_2, \ldots, \mathbf{Y}_N \right)$ of $N$ previously observed patient diagnoses, given the model's parameters $\boldsymbol{\theta}$. This is given by a product of the sums in the denominator of @eq-bayes-law:

$$
\mathcal{L}(\mathcal{D} \mid \boldsymbol{\theta}) = \prod_{i=1}^N \sum_{j=1}^{2^V} P \left( \mathbf{Y}_i \mid \mathbf{X} = \boldsymbol{\xi}_j \right) P \left( \mathbf{X} = \boldsymbol{\xi}_j \mid T_i \right)
$$

Typically, one would compute the logarithm of this quantity for computational reasons. There is a wide array of available methods to obtain maximum likelihood estimates from this function or sample from the posterior over $\boldsymbol{\theta}$. In this work, we use Markov chain Monte Carlo sampling (MCMC) via the `emcee` Python package [@foreman-mackey_emcee_2013].

## Mixture of HMMs {#sec-mixture-of-hmms}

The just introduced model is capable of learning one set of or distribution over parameters $\boldsymbol{\theta}_s$ from a cohort of patients with a primary tumor in a given subsite $s$. If we tried to train it with a cohort consisting of patients with tumors in two very different subsites, the model would likely learn parameters that represent a compromise between the two subcohort's true parameters. This compromise might describe neither of the subcohorts' lymphatic spread patterns sufficiently well.

In such cases, mixture models are often considered. They assume the data to come from a finite mixture distribution, which -- in our particular case -- can be written following @mclachlan_finite_2019:

$$
P \left( \mathbf{X} \mid \boldsymbol{\Psi} \right) = \sum_{i=0}^g \pi_i \cdot P \left( \mathbf{X} \mid \boldsymbol{\theta}_i \right)
$$ {#eq-mixture-distribution}

Here, the $\pi_i \in [0,1]$ are the mixing proportions with $\sum_{i=0}^g \pi_i = 1$, while the $\boldsymbol{\Psi} = \left( \boldsymbol{\theta}_1, \ldots, \boldsymbol{\theta}_g, \pi_1, \ldots, \pi_g \right)$ is the entire vector of unknown parameters in this mixture of $g$ components. Note that we will implement our model such that some of the parameters in each $\boldsymbol{\theta}_i$ are shared across the $g$ components -- namely the $t_{r \rightarrow v}$ corresponding to the blue arcs in @fig-graph-with-obs.

Let now $\mathcal{D} = \left( \mathcal{D}_1, \ldots, \mathcal{D}_s \right)$ be a dataset consisting of $s$ subcohorts each corresponding to patients with tumor in the same subsite. Also, let subcohort $i$ have $N_i$ patients in it.

We can now think of a way of generating this dataset $\mathcal{D}$ under this mixture model:

1. We draw a categorical RV $C_j$ that may take on the values $1, \ldots, g$ with corresponding probabilities $\pi_1, \ldots, \pi_g$:
  $$
  P(C_j=c_j) = \pi_{c_j}
  $$
  For convenience, we also transform this into a one-hot encoding $\mathbf{z}_j$ that is zero everywhere, except $z_{ij}=1$ when $C_j=i$:
  $$
  P \left( \mathbf{Z}_j=\mathbf{z}_j \right) = \prod_{i=0}^g \pi_i^{z_{ij}}
  $$

2. Then, we draw $\mathbf{y}_j$ from the conditional probability of $P \left( \mathbf{Y}_j \mid \mathbf{Z}_j, \boldsymbol{\Psi} \right)$, which is now given by a slightly modified version of @eq-mixture-distribution:
  $$
  P \left( \mathbf{Y}_j \mid \mathbf{Z}_j, \boldsymbol{\Psi} \right) = \frac{\sum_{i=0}^g z_{ij} \cdot \pi_i \cdot P \left( \mathbf{Y}_j \mid \boldsymbol{\theta}_i \right)}{\sum_{i=0}^g z_{ij} \cdot \pi_i} = P \left( \mathbf{Y}_j \mid \boldsymbol{\theta}_\ell \right)
  $$
  where $\ell$ is just so that $z_{\ell j}=1$, and all other entries of $\mathbf{z}_j$ are zero.

# Results {#sec-results}

# Discussion

### 2024-01-16 content/manuscript.qmd
correct/expand already written stuff

@@
     include-in-header:
-      text: "\\usepackage{newtxtext,newtxmath}"
+      - text: "\\usepackage{newtxtext,newtxmath}"
+  html:
+    code-line-numbers: true
+    code-tools: true
 crossref:
@@
 
-Head and neck squamous cell carcinomas (HNSCC) frequently spread through the lymphatic system [@lindberg_distribution_1972;@candela_patterns_1990;@woolgar_histological_1999].
+Head and neck squamous cell carcinomas (HNSCC) frequently spread through the lymphatic system [@lindberg_distribution_1972;@woolgar_histological_1999].
 Current diagnostic imaging modalities are unable to detect microscopic nodal metastases, which requires pathological examination of extracted tissue [@snyder_petct_2021;@strohl_petct_2021].
 Since the recurrence of nodal disease is detrimental to a patient's outcome [@ho_decision_2014], large volumes in the head and neck region are often irradiated electively to minimize the risk of missing occult disease.
-Decision guidelines about which nodal regions -- i.e., anatomically defined lymph node levels (LNLs) -- to irradiate [@vorwerk_guidelines_2011;@gregoire_delineation_2018;@biau_selection_2019] are currently not based on a patient's individual risk, but only on the overall prevalence of nodal disease as reported in the literature [@lindberg_distribution_1972;@candela_patterns_1990;@woolgar_histological_1999;@chao_determination_2002].
+Decision guidelines about which nodal regions -- i.e., anatomically defined lymph node levels (LNLs) -- to irradiate [@vorwerk_guidelines_2011;@biau_selection_2019] are currently not based on a patient's individual risk, but only on the overall prevalence of nodal disease as reported in the literature [@lindberg_distribution_1972;@woolgar_histological_1999].
 
@@
 
-2.
-an interpretable hidden Markov model (HMM), trained with this data, to predict the risk for occult nodal disease [@ludwig_hidden_2021;@ludwig_modelling_2023], given an individual patient's diagnosis.
+2. an interpretable hidden Markov model (HMM), trained with this data, to predict the risk for occult nodal disease [@ludwig_hidden_2021], given an individual patient's diagnosis.
 
@@
 HNSCC patients with primary tumors at different subsites, e.g.
-in the oral cavity and in the oropharynx, also show different patterns of lymphatic spread [@lindberg_distribution_1972;@candela_patterns_1990;@woolgar_histological_1999].
+in the oral cavity and in the oropharynx, also show different patterns of lymphatic spread [@lindberg_distribution_1972;@woolgar_histological_1999].
 To naively capture this in a risk prediction using our previously developed model, one would need to train entirely separate models for each subsite, based on separated datasets.
@@
 
-Each LNL $v \in \{ 1, 2, \ldots, V\}$ considered in our model is represented by a hidden binary random variable (RV) $X_v[t]$ representing the true -- but not directly observable -- state of that level at the abstract time-step $t$.
-It may take on the values "healthy" (0) or "involved" (1).
+Each LNL $v \in \{ 1, 2, \ldots, V\}$ considered in our model is represented by a hidden binary random variable (RV) $X_v[t]$ taking on the true -- but not directly observable -- state of that level at the abstract time-step $t$ (0 for "healthy" and 1 for "involved").
 Collected in a random vector $\mathbf{X}[t] = \left( X_1[t], X_2[t], \ldots, X_V[t] \right)$ they form the patient's hidden state w.r.t.
@@
 
-![Example of a parametrized graph representation of a small lymphatic network consisting of only three LNLs.
-Blue, round nodes represent the hidden RVs, while the orange square nodes show the observed RVs.
-Every arc represents a conditional probability parametrized with the quantity noted next to it.](../static/graph_with_obs.png){#fig-graph-with-obs width="25%"}
+![Parametrized graph representation of the lymphatic network considering four LNLs.
+Blue, round nodes represent the hidden RVs, while orange square nodes show the observed RVs.
+Arcs represent a conditional probability parametrized with the quantity noted next to it.](../static/graph_with_obs.svg){#fig-graph-with-obs width="25%"}
 
@@
 This corresponds to one of the blue arcs in @fig-graph-with-obs.
+Note that the values in the row with $X_3[t+1]=0$ is all zeros, and the row with $X_3[t+1]=1$ all ones if $X_3[t]=1$.
 {#tbl-spread-2-to-3 tbl-colwidths="[25,18,40]"}
@@
 
-And set every entry of this starting distribution to zero, except the first entry, which we set to one.
+And set every entry of this starting distribution to zero, except the first one, which we set to one.
 This means at $t=0$ there is a probability of one to be in the completly healthy state $\boldsymbol{\xi}_0 = \begin{pmatrix} 0, 0, \ldots, 0 \end{pmatrix}$.
@@
 $$
-P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid T \right) = \sum_{t=0}^{t_\text{max}} P_T(t) \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
+P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid T, \boldsymbol{\theta} \right) = \sum_{t=0}^{t_\text{max}} P_T(t) \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
 $$ {#eq-marginalized-evolution}
@@
 
-Note that [the above equation](#eq-marginalized-evolution) still depends on the parametrization of the transition matrix.
+Note that @eq-marginalized-evolution still depends on the parametrization of the transition matrix.
 We collect these parameters in a vector $\boldsymbol{\theta}=\{ b_v, t_{r \rightarrow v}, p_\text{advanced} \}$.
 
-To train our model, we need to compute the likelihood of a dataset $\mathcal{D}=\left( \mathbf{Y}_1, \mathbf{Y}_2, \ldots, \mathbf{Y}_N \right)$ of $N$ previously observed patient diagnoses, given the model's parameters $\boldsymbol{\theta}$.
+To train our model, we need to compute the likelihood of a dataset $\mathcal{D}=\left( \mathbf{y}_1, \mathbf{y}_2, \ldots, \mathbf{y}_N \right)$ of $N$ previously observed patient diagnoses, given the model's parameters $\boldsymbol{\theta}$.
 This is given by a product of the sums in the denominator of @eq-bayes-law:
@@
 $$
-\mathcal{L}(\mathcal{D} \mid \boldsymbol{\theta}) = \prod_{i=1}^N \sum_{j=1}^{2^V} P \left( \mathbf{Y}_i \mid \mathbf{X} = \boldsymbol{\xi}_j \right) P \left( \mathbf{X} = \boldsymbol{\xi}_j \mid T_i \right)
+\mathcal{L}(\mathcal{D} \mid \boldsymbol{\theta}) = \prod_{i=1}^N \underbrace{\sum_{j=1}^{2^V} P \left( \mathbf{Y} = \mathbf{y}_i \mid \mathbf{X} = \boldsymbol{\xi}_j \right) P \left( \mathbf{X} = \boldsymbol{\xi}_j \mid T_i, \boldsymbol{\theta} \right)}_{P \left( \mathbf{Y} = \mathbf{y}_i \mid T_i, \boldsymbol{\theta} \right)}
 $$
@@
 
-The just introduced model is capable of learning one set of or distribution over parameters $\boldsymbol{\theta}_s$ from a cohort of patients with a primary tumor in a given subsite $s$.
+The just introduced model is capable of learning one set of or distribution over parameters $\boldsymbol{\theta}$ from a cohort of patients with a primary tumor in a given subsite.
 If we tried to train it with a cohort consisting of patients with tumors in two very different subsites, the model would likely learn parameters that represent a compromise between the two subcohort's true parameters.
@@
 $$
-P \left( \mathbf{X} \mid \boldsymbol{\Psi} \right) = \sum_{i=0}^g \pi_i \cdot P \left( \mathbf{X} \mid \boldsymbol{\theta}_i \right)
+P \left( \mathbf{Y} \mid \boldsymbol{\Psi} \right) = \sum_{i=0}^g \pi_i \cdot P \left( \mathbf{Y} \mid \boldsymbol{\theta}_i \right)
 $$ {#eq-mixture-distribution}
@@
 
-1. We draw a categorical RV $C_j$ that may take on the values $1, \ldots, g$ with corresponding probabilities $\pi_1, \ldots, \pi_g$:
-  $$
-  P(C_j=c_j) = \pi_{c_j}
-  $$
-  For convenience, we also transform this into a one-hot encoding $\mathbf{z}_j$ that is zero everywhere, except $z_{ij}=1$ when $C_j=i$:
+1.
+We draw a categorical RV $C_j$ that may take on the values $1, \ldots, g$ with corresponding probabilities $\pi_1, \ldots, \pi_g$.
+Hence, we have $P(C_j=c_j) = \pi_{c_j}$.
+For convenience, we also transform this into a one-hot encoding $\mathbf{z}_j$ that is zero everywhere, except $z_{\ell j}=1$ when $C_j=\ell$:
   $$
@@
   $$
-  P \left( \mathbf{Y}_j \mid \mathbf{Z}_j, \boldsymbol{\Psi} \right) = \frac{\sum_{i=0}^g z_{ij} \cdot \pi_i \cdot P \left( \mathbf{Y}_j \mid \boldsymbol{\theta}_i \right)}{\sum_{i=0}^g z_{ij} \cdot \pi_i} = P \left( \mathbf{Y}_j \mid \boldsymbol{\theta}_\ell \right)
+  P \left( \mathbf{Y}_j \mid \mathbf{Z}_j, \boldsymbol{\Psi} \right) = \frac{\sum_{i=0}^g z_{ij} \pi_i P \left( \mathbf{Y}_j \mid \boldsymbol{\theta}_i \right)}{\sum_{i=0}^g z_{ij} \pi_i} = P \left( \mathbf{Y}_j \mid \boldsymbol{\theta}_\ell \right)
   $$

### 2024-01-16 content/manuscript.qmd
improve mixture model formalism

@@
 Since the recurrence of nodal disease is detrimental to a patient's outcome [@ho_decision_2014], large volumes in the head and neck region are often irradiated electively to minimize the risk of missing occult disease.
-Decision guidelines about which nodal regions -- i.e., anatomically defined lymph node levels (LNLs) -- to irradiate [@vorwerk_guidelines_2011;@biau_selection_2019] are currently not based on a patient's individual risk, but only on the overall prevalence of nodal disease as reported in the literature [@lindberg_distribution_1972;@woolgar_histological_1999].
+Decision guidelines about which nodal regions -- i.e., anatomically defined lymph node levels (LNLs) -- to irradiate [@biau_selection_2019] are currently not based on a patient's individual risk, but only on the overall prevalence of nodal disease as reported in the literature [@lindberg_distribution_1972;@woolgar_histological_1999].
 
@@
 
+![Parametrized graph representation of the lymphatic network considering four LNLs.
+Blue, round nodes represent the hidden RVs, while orange square nodes show the observed RVs.
+Arcs represent a conditional probability parametrized with the quantity noted next to it.](../static/graph_with_obs.svg){#fig-graph-with-obs width="22%"}
+
 To compute the personalized risk of occult disease $\mathbf{X}$, given a diagnosis $\mathbf{Y}$, we can begin by stating Bayes' law:
@@
 Hence, a diagnosis obtained with diagnostic modality $\mathcal{O}$ can be collected in the random vector $\mathbf{Y}^\mathcal{O}$.
-
-![Parametrized graph representation of the lymphatic network considering four LNLs.
-Blue, round nodes represent the hidden RVs, while orange square nodes show the observed RVs.
-Arcs represent a conditional probability parametrized with the quantity noted next to it.](../static/graph_with_obs.svg){#fig-graph-with-obs width="25%"}
 
@@
 
-To train our model, we need to compute the likelihood of a dataset $\mathcal{D}=\left( \mathbf{y}_1, \mathbf{y}_2, \ldots, \mathbf{y}_N \right)$ of $N$ previously observed patient diagnoses, given the model's parameters $\boldsymbol{\theta}$.
+To train our model, we need to compute the likelihood of a dataset $\mathbf{D}=\left( \mathbf{y}_1, \mathbf{y}_2, \ldots, \mathbf{y}_N \right)$ of $N$ patients, given the model's parameters $\boldsymbol{\theta}$.
 This is given by a product of the sums in the denominator of @eq-bayes-law:
@@
 $$
-\mathcal{L}(\mathcal{D} \mid \boldsymbol{\theta}) = \prod_{i=1}^N \underbrace{\sum_{j=1}^{2^V} P \left( \mathbf{Y} = \mathbf{y}_i \mid \mathbf{X} = \boldsymbol{\xi}_j \right) P \left( \mathbf{X} = \boldsymbol{\xi}_j \mid T_i, \boldsymbol{\theta} \right)}_{P \left( \mathbf{Y} = \mathbf{y}_i \mid T_i, \boldsymbol{\theta} \right)}
-$$
+P \left(\mathbf{D} \mid \boldsymbol{\theta} \right) = \prod_{i=1}^N \underbrace{\sum_{j=1}^{2^V} P \left( \mathbf{Y} = \mathbf{y}_i \mid \mathbf{X} = \boldsymbol{\xi}_j \right) P \left( \mathbf{X} = \boldsymbol{\xi}_j \mid T_i, \boldsymbol{\theta} \right)}_{P \left( \mathbf{Y} = \mathbf{y}_i \mid T_i, \boldsymbol{\theta} \right)}
+$$ {#eq-likelihood}
 
@@
 
-In such cases, mixture models are often considered. They assume the data to come from a finite mixture distribution, which -- in our particular case -- can be written following @mclachlan_finite_2019:
+In such cases, mixture models are often considered. They assume the data to come from a finite mixture distribution, which -- in our particular case -- can be written as follows:
 
 $$
-P \left( \mathbf{Y} \mid \boldsymbol{\Psi} \right) = \sum_{i=0}^g \pi_i \cdot P \left( \mathbf{Y} \mid \boldsymbol{\theta}_i \right)
+P \left( \mathbf{D} \mid \boldsymbol{\Psi} \right) = \sum_{j=0}^g c_j P \left( \mathbf{D} \mid \boldsymbol{\theta}_j \right)
 $$ {#eq-mixture-distribution}
 
-Here, the $\pi_i \in [0,1]$ are the mixing proportions with $\sum_{i=0}^g \pi_i = 1$, while the $\boldsymbol{\Psi} = \left( \boldsymbol{\theta}_1, \ldots, \boldsymbol{\theta}_g, \pi_1, \ldots, \pi_g \right)$ is the entire vector of unknown parameters in this mixture of $g$ components.
-Note that we will implement our model such that some of the parameters in each $\boldsymbol{\theta}_i$ are shared across the $g$ components -- namely the $t_{r \rightarrow v}$ corresponding to the blue arcs in @fig-graph-with-obs.
+Here, the $\mathbf{c} \in [0,1]^g$ is the vector of mixing proportions with $\sum_{j=0}^g c_j = 1$, while the $\boldsymbol{\Psi} = \left( \boldsymbol{\theta}_1, \ldots, \boldsymbol{\theta}_g\right)$ is the vector of all $g$ models' parameters.
+Note that we will implement our model such that some of the parameters in each $\boldsymbol{\theta}_j$ are shared across the $g$ components -- namely the $t_{r \rightarrow v}$ corresponding to the blue arcs in @fig-graph-with-obs.
 
-Let now $\mathcal{D} = \left( \mathcal{D}_1, \ldots, \mathcal{D}_s \right)$ be a dataset consisting of $s$ subcohorts each corresponding to patients with tumor in the same subsite.
-Also, let subcohort $i$ have $N_i$ patients in it.
+Let now $\mathfrak{D} = \left( \mathbf{D}_1, \ldots, \mathbf{D}_s \right)$ be a dataset consisting of $s$ subcohorts of patients.
+Within a subcohort $i$ we find $N_i$ patients with tumors in the same subsite.
+We can then introduce a _latent variable_ $\mathbf{Z}$ with a one-hot-encoding: Basically, it can take on values $\mathbf{z}_i \in \{ 0,1 \}^g$ with $z_{ij}=1$ if subcohort $i$ belongs to component $j$ and $z_{ij}=0$ else.
 
-We can now think of a way of generating this dataset $\mathcal{D}$ under this mixture model:
+Introducing this latent variable is helpful, because the mixture model likelihood is invariant w.r.t.
+permutations of the component labels.
+This introduces a number of problems, e.g.
+for common MCMC sampling methods.
+The latent variable $\mathbf{Z}$ allows us to derive two sets of interdependent equations that we may solve in an iterative fashion (see e.g.
+@bishop_pattern_2006 for a detailed derivation) that is commonly referred to as _expectation-maximization (EM)_ algorithm:
 
-1.
-We draw a categorical RV $C_j$ that may take on the values $1, \ldots, g$ with corresponding probabilities $\pi_1, \ldots, \pi_g$.
-Hence, we have $P(C_j=c_j) = \pi_{c_j}$.
-For convenience, we also transform this into a one-hot encoding $\mathbf{z}_j$ that is zero everywhere, except $z_{\ell j}=1$ when $C_j=\ell$:
+The first are the probabilities of subcohort $i$ to belong to component $j$, given a set of parameters $\boldsymbol{\Psi}^\star$. These are often called the _responsibilities_:
 $$
-  P \left( \mathbf{Z}_j=\mathbf{z}_j \right) = \prod_{i=0}^g \pi_i^{z_{ij}}
+\gamma (z_{ij}) = P \left( z_{ij}=1 \mid \mathbf{D}_i , \boldsymbol{\Psi}^\star, \mathbf{c} \right) = \frac{ c_j P \left( \mathbf{D}_i \mid \boldsymbol{\theta}_j^\star \right) }{\sum_{k=0}^g c_k P \left( \mathbf{D}_i \mid \boldsymbol{\theta}_k^\star \right)}
 $$
 
-2.
-Then, we draw $\mathbf{y}_j$ from the conditional probability of $P \left( \mathbf{Y}_j \mid \mathbf{Z}_j, \boldsymbol{\Psi} \right)$, which is now given by a slightly modified version of @eq-mixture-distribution:
+From this, we can compute new mixing proportions $c_j^\star=\sum_{i=1}^s \gamma (z_{ij}) / s$ and then infer new parameters $\boldsymbol{\Psi}^\star$ -- e.g.
+via MCMC sampling -- from the resulting likelihood:
+
 $$
-  P \left( \mathbf{Y}_j \mid \mathbf{Z}_j, \boldsymbol{\Psi} \right) = \frac{\sum_{i=0}^g z_{ij} \pi_i P \left( \mathbf{Y}_j \mid \boldsymbol{\theta}_i \right)}{\sum_{i=0}^g z_{ij} \pi_i} = P \left( \mathbf{Y}_j \mid \boldsymbol{\theta}_\ell \right)
+P \left( \mathbf{D} \mid \boldsymbol{\Psi}, \mathbf{c}^\star \right) = \sum_{j=0}^g c_j^\star P \left( \mathbf{D} \mid \boldsymbol{\theta}_j \right)
 $$
-  where $\ell$ is just so that $z_{\ell j}=1$, and all other entries of $\mathbf{z}_j$ are zero.
+
+## Implementation {#sec-implementation}
+
+We did this in this and that fashion...
 
@@
 
-# Discussion
+Works super well, of course!
+
+# Discussion {#sec-discussion}
+
+All that's left is for other people to bring this into clinical practice.

### 2024-01-17 content/manuscript.qmd
add more co-authors

@@
       - ref: 2
+  - name: Yoel Perez Haas
+    affiliations:
+      - ref: 1
+      - ref: 2
+  - name: Esmée Lauren Looman
+    affiliations:
+      - ref: 1
+      - ref: 2
+  - name: Adrian Schubert
+    affiliations:
+      - ref: 3
+      - ref: 4
+      - ref: 7
+  - name: Dorothea Barbatei
+    affiliations:
+      - ref: 8
+  - name: Laurence Bauwens
+    affiliations:
+      - ref: 8
+  - name: Jean-Marc Hoffmann
+    orcid: 0000-0003-1193-0863
+    affiliations:
+      - ref: 2
+  - name: Sandrine Werlen
+    affiliations:
+      - ref: 4
+      - ref: 5
+  - name: Olgun Elicin
+    orcid: 0000-0002-6996-0646
+    affiliations:
+      - ref: 3
+  - name: Matthias Dettmer
+    orcid: 0000-0003-0948-1392
+    affiliations:
+      - ref: 6
+      - ref: 10
+  - name: Philippe Zrounba
+    orcid: 0000-0001-6223-6113
+    affiliations:
+      - ref: 9
+  - name: Bertrand Poumayou
+    orcid: 0000-0002-8060-7827
+    affiliations:
+      - ref: 2
+  - name: Panagiotis Balermpas
+    orcid: 0000-0001-5261-6446
+    affiliations:
+      - ref: 2
+  - name: Vincent Grégoire
+    affiliations:
+      - ref: 8
+  - name: Roland Giger
+    orcid: 0000-0002-5574-3210
+    affiliations:
+      - ref: 4
+      - ref: 5
   - name: Jan Unkelbach
@@
     name: University of Zurich
-    departement: Departement of Physics
+    department: Department of Physics
     city: Zurich
@@
     name: University Hospital Zurich
-    departement: Departement of Radiation Oncology
+    department: Department of Radiation Oncology
     city: Zurich
     country: Switzerland
+  - id: 3
+    name: Bern University Hospital
+    department: Department of Radiation Oncology
+    city: Bern
+    country: Switzerland
+  - id: 4
+    name: Bern University Hospital
+    department: Department of ENT, Head & Neck Surgery
+    city: Bern
+    country: Switzerland
+  - id: 5
+    name: Bern University Hospital
+    department: Head and Neck Anticancer Center
+    city: Bern
+    country: Switzerland
+  - id: 6
+    name: Bern University Hospital
+    department: Institute of Tissue Medicine and Pathology
+    city: Bern
+    country: Switzerland
+  - id: 7
+    name: Réseau Hospitalier Neuchâtelois
+    department: Department of ENT, Head & Neck Surgery
+    city: Neuchâtelois
+    country: Switzerland
+  - id: 8
+    name: Centre Léon Bérard
+    department: Department of Radiation Oncology
+    city: Lyon
+    country: France
+  - id: 9
+    name: Centre Léon Bérard
+    department: Department of Head and Neck Surgery
+    city: Lyon
+    country: France
+  - id: 10
+    name: Klinikum Stuttgart
+    department: Institute of Pathology
+    city: Stuttgart
+    country: Germany
 bibliography: ../references.bib
@@
 Blue, round nodes represent the hidden RVs, while orange square nodes show the observed RVs.
-Arcs represent a conditional probability parametrized with the quantity noted next to it.](../static/graph_with_obs.svg){#fig-graph-with-obs width="22%"}
+Arcs represent a conditional probability parametrized with the quantity noted next to it.](../static/graph_with_obs.svg){#fig-graph-with-obs width="20%"}
 
@@
 
-When a patient gets diagnosed, their LNLs' involvement status is assessed using a clinical modality $\mathcal{O}$ with its repsective sensitivity $s_N^\mathcal{O}$ and specificity $s_P^\mathcal{O}$.
-This involvement status, too, is modelled as an RV $Y_v^\mathcal{O}$.
-Hence, a diagnosis obtained with diagnostic modality $\mathcal{O}$ can be collected in the random vector $\mathbf{Y}^\mathcal{O}$.
+When a patient gets diagnosed, their LNLs' involvement status is assessed using a clinical modality (e.g.
+computed tomography (CT) or magnetic resonance imaging (MRI)) with its repsective sensitivity $s_N$ and specificity $s_P$.
+This involvement status, too, is modelled as an RV $Y_v$.
+Hence, a diagnosis obtained with this diagnostic modality can be collected in the random vector $\mathbf{Y}$.
 
@@
 The arcs in [this graph](#fig-graph-with-obs) represent conditional probabilities.
-The orange arcs, as mentioned, correspond to observing a diagnosis $Z_v^\mathcal{O}$, given a hidden state $X_v$.
+The orange arcs, as mentioned, correspond to observing a diagnosis $Y_v$, given a hidden state $X_v$.
 The values this conditional probability $P(Y_v \mid X_v)$ can take on are tabulated in @tbl-sens-spec.

### 2024-01-18 content/manuscript.qmd
add info about data & edits (see below)
I have removed the formalism related to sensitivity and specificity.
Also, the mixture model is now not motivated as a lift of the old
model's limitations anymore.

--- 2024-01-16 content/manuscript.qmd
@@
       - ref: 2
+  - name: Yoel Perez Haas
+    affiliations:
+      - ref: 1
+      - ref: 2
+  - name: Esmée Lauren Looman
+    affiliations:
+      - ref: 1
+      - ref: 2
+  - name: Sergi Benavente
+    orcid: 0000-0002-2700-3522
+    affiliations:
+      - ref: 11
+  - name: Adrian Schubert
+    affiliations:
+      - ref: 3
+      - ref: 4
+      - ref: 7
+  - name: Dorothea Barbatei
+    affiliations:
+      - ref: 8
+  - name: Laurence Bauwens
+    affiliations:
+      - ref: 8
+  - name: Jean-Marc Hoffmann
+    orcid: 0000-0003-1193-0863
+    affiliations:
+      - ref: 2
+  - name: Sandrine Werlen
+    affiliations:
+      - ref: 4
+      - ref: 5
+  - name: Olgun Elicin
+    orcid: 0000-0002-6996-0646
+    affiliations:
+      - ref: 3
+  - name: Matthias Dettmer
+    orcid: 0000-0003-0948-1392
+    affiliations:
+      - ref: 6
+      - ref: 10
+  - name: Philippe Zrounba
+    orcid: 0000-0001-6223-6113
+    affiliations:
+      - ref: 9
+  - name: Bertrand Poumayou
+    orcid: 0000-0002-8060-7827
+    affiliations:
+      - ref: 2
+  - name: Panagiotis Balermpas
+    orcid: 0000-0001-5261-6446
+    affiliations:
+      - ref: 2
+  - name: Vincent Grégoire
+    affiliations:
+      - ref: 8
+  - name: Roland Giger
+    orcid: 0000-0002-5574-3210
+    affiliations:
+      - ref: 4
+      - ref: 5
   - name: Jan Unkelbach
@@
     name: University Hospital Zurich
-    departement: Departement of Radiation Oncology
+    department: Department of Radiation Oncology
     city: Zurich
     country: Switzerland
+  - id: 3
+    name: Bern University Hospital
+    department: Department of Radiation Oncology
+    city: Bern
+    country: Switzerland
+  - id: 4
+    name: Bern University Hospital
+    department: Department of ENT, Head & Neck Surgery
+    city: Bern
+    country: Switzerland
+  - id: 5
+    name: Bern University Hospital
+    department: Head and Neck Anticancer Center
+    city: Bern
+    country: Switzerland
+  - id: 6
+    name: Bern University Hospital
+    department: Institute of Tissue Medicine and Pathology
+    city: Bern
+    country: Switzerland
+  - id: 7
+    name: Réseau Hospitalier Neuchâtelois
+    department: Department of ENT, Head & Neck Surgery
+    city: Neuchâtelois
+    country: Switzerland
+  - id: 8
+    name: Centre Léon Bérard
+    department: Department of Radiation Oncology
+    city: Lyon
+    country: France
+  - id: 9
+    name: Centre Léon Bérard
+    department: Department of Head and Neck Surgery
+    city: Lyon
+    country: France
+  - id: 10
+    name: Klinikum Stuttgart
+    department: Institute of Pathology
+    city: Stuttgart
+    country: Germany
+  - id: 11
+    name: Hospital Vall d'Hebron
+    department: Departement of Radiation Oncology
+    city: Barcelona
+    country: Spain
 bibliography: ../references.bib
@@
 in the oral cavity and in the oropharynx, also show different patterns of lymphatic spread [@lindberg_distribution_1972;@woolgar_histological_1999].
-To naively capture this in a risk prediction using our previously developed model, one would need to train entirely separate models for each subsite, based on separated datasets.
-We consider this a limitation for two reasons:
-
-1.
-While the spread from the tumor to the LNLs may differ for different primary tumor subsites, we would assume the spread from involved LNLs further down the lymphatic network to still be similar.
-Thus, a dataset of oral cavity patients may contain information that is relevant to the oropharynx model and vice versa.
-2.
-A tumor may lie anatomically between two subsites.
-Then, it makes sense to assume that this tumor spreads in a way that can be described as a combination of the typical spread patterns of these two subsites, given they are close enough to each other.
-This presents an opportunity to reduce the model's complexity: If a subsite's lymphatic spread can be described as a mixture of two others' spread patterns, we only require two distinct HMMs to describe three tumor subsites.
-
-In the work presented here, we implement a mixture of HMMs that aims to address these limitations.
-In @sec-materials-and-methods we first introduce the basic concept of the HMM (@sec-hmm), followed by a general description of mixture models (@sec-mixture-of-hmms) and how we implemented them in our particular case.
+Our model does so far not have the capability to naturally describe different tumor subsites.
+To that end, we present an approach using mixtures of these HMMs in this work.
+This makes intuitive sense, because if a tumor lies anatomically between two anatomically close subsites with slightly different spread patterns, we may be able to describe its lymphatic progression as a mixture of these two spread patterns.
 
 # Materials and Methods {#sec-materials-and-methods}
-
-![Parametrized graph representation of the lymphatic network considering four LNLs.
-Blue, round nodes represent the hidden RVs, while orange square nodes show the observed RVs.
-Arcs represent a conditional probability parametrized with the quantity noted next to it.](../static/graph_with_obs.svg){#fig-graph-with-obs width="22%"}
 
@@
 
-Each LNL $v \in \{ 1, 2, \ldots, V\}$ considered in our model is represented by a hidden binary random variable (RV) $X_v[t]$ taking on the true -- but not directly observable -- state of that level at the abstract time-step $t$ (0 for "healthy" and 1 for "involved").
-Collected in a random vector $\mathbf{X}[t] = \left( X_1[t], X_2[t], \ldots, X_V[t] \right)$ they form the patient's hidden state w.r.t.
+Each LNL $v \in \{ 1, 2, \ldots, V\}$ considered in our model is represented by a binary random variable (RV) $X_v[t]$ taking on the true state of that level at the abstract time-step $t$ (0 for "healthy" and 1 for "involved").
+Collected in a random vector $\mathbf{X}[t] = \left( X_1[t], X_2[t], \ldots, X_V[t] \right)$ they form the patient's state w.r.t.
 their lymphatic involvement at time $t$.
 
-When a patient gets diagnosed, their LNLs' involvement status is assessed using a clinical modality $\mathcal{O}$ with its repsective sensitivity $s_N^\mathcal{O}$ and specificity $s_P^\mathcal{O}$.
-This involvement status, too, is modelled as an RV $Y_v^\mathcal{O}$.
-Hence, a diagnosis obtained with diagnostic modality $\mathcal{O}$ can be collected in the random vector $\mathbf{Y}^\mathcal{O}$.
+We model the process of tumor progression via lymphatic drainage by connecting the RVs in a graph, as shown in @fig-graph-with-obs.
+The arcs in [this graph](#fig-graph-with-obs) represent conditional probabilities.
+The orange arcs correspond to observing a diagnosis $Y_v$, given the true state $X_v$.
+For the sake of brevity, we will not go into the details of describing how to infer the true -- but technically hidden -- state of LNL involvement from diagnoses with lower-than-one sensitivity and specificity.
+This description can be found in @zora231470.
+Throughout this work, we instead combine diagnostic and pathologic involvement information from the data into a "maximum likelihood" diagnosis and assume its sensitivity and specificity to be one, meaning the normally hidden state $X_v$ becomes the observed state.
+This simplification is reasonable for pathologic involvement and because we are at this stage more interested in testing whether our model is able to describe a realistic probability distribution over lymphatic involvement.
 
-We model the process of tumor progression via lymphatic drainage by connecting the hidden RVs in a graph, as shown in @fig-graph-with-obs.
-The arcs in [this graph](#fig-graph-with-obs) represent conditional probabilities.
-The orange arcs, as mentioned, correspond to observing a diagnosis $Z_v^\mathcal{O}$, given a hidden state $X_v$.
-The values this conditional probability $P(Y_v \mid X_v)$ can take on are tabulated in @tbl-sens-spec.
-
-+---------+---------+---------+
-|         | $X_v=0$ | $X_v=1$ |
-+=========+:=======:+:=======:+
-| $Y_v=0$ | $s_P$   | $1-s_N$ |
-+---------+---------+---------+
-| $Y_v=1$ | $1-s_P$ | $s_N$   |
-+---------+---------+---------+
-
-: Table encoding the conditional probability $P(Y_v \mid X_v)$ to correctly (or wrongly) diagnose an LNL's true hidden state.
-This table is associated with the orange arcs in @fig-graph-with-obs.
-{#tbl-sens-spec}
-
-Similarily, the red arcs in @fig-graph-with-obs depict the probability that the primary tumor spreads within one abstract time-step.
-Lastly, the blue arcs symbolize the spread from an upstream LNL -- given it is already metastatic -- to a downstream level.
+The red arcs in the graph of @fig-graph-with-obs depict the probability that the primary tumor spreads within one abstract time-step.
+While the blue arcs symbolize the spread from an upstream LNL -- given it is already metastatic -- to a downstream level.
 For example, the edge from $X_2$ to $X_3$ encodes the probability $P\left( X_3[t+1] \mid X_2[t] \right)$, which is parametrized with $b_3$, and $t_{2 \rightarrow 3}$ and tabulated in @tbl-spread-2-to-3.
@@
 
-With the introduced conditional probabilities, we can now compute the joint probability of any complete hidden state $\mathbf{X}[t]=\boldsymbol{\xi}_i$ transitioning to any other possible state $\mathbf{X}[t+1]=\boldsymbol{\xi}_j$.
-Here, the $i$ and $j$ enumerate all $2^V$ combinations of the $V$ binary RVs.
+With the introduced conditional probabilities, we can now compute the joint probability of any complete state $\mathbf{X}[t]=\boldsymbol{\xi}_i$ transitioning to any other possible state $\mathbf{X}[t+1]=\boldsymbol{\xi}_j$.
+Here, when we use $\boldsymbol{\xi}$ instead of $\mathbf{x}$ for the values the random vector $\mathbf{X}$ can take on, the $i$ and $j$ enumerate all $2^V$ combinations of the $V$ binary RVs.
 In the graph shown in @fig-graph-with-obs, this amounts to $2^V=8$ distinct $\boldsymbol{\xi}_i$.
@@
 
-To train our model, we need to compute the likelihood of a dataset $\mathbf{D}=\left( \mathbf{y}_1, \mathbf{y}_2, \ldots, \mathbf{y}_N \right)$ of $N$ patients, given the model's parameters $\boldsymbol{\theta}$.
-This is given by a product of the sums in the denominator of @eq-bayes-law:
+To train our model, we need to compute the likelihood of a dataset $\mathbf{D}=\left( \mathbf{x}_1, \mathbf{x}_2, \ldots, \mathbf{x}_N \right)$ of $N$ patients, given the model's parameters $\boldsymbol{\theta}$.
+This is given by a product of the terms in @eq-marginalized-evolution:
 
 $$
-P \left(\mathbf{D} \mid \boldsymbol{\theta} \right) = \prod_{i=1}^N \underbrace{\sum_{j=1}^{2^V} P \left( \mathbf{Y} = \mathbf{y}_i \mid \mathbf{X} = \boldsymbol{\xi}_j \right) P \left( \mathbf{X} = \boldsymbol{\xi}_j \mid T_i, \boldsymbol{\theta} \right)}_{P \left( \mathbf{Y} = \mathbf{y}_i \mid T_i, \boldsymbol{\theta} \right)}
+P \left(\mathbf{D} \mid \boldsymbol{\theta} \right) = \prod_{i=1}^N P \left( \mathbf{X} = \mathbf{x}_i \mid T_i, \boldsymbol{\theta} \right)
 $$ {#eq-likelihood}
@@
 In this work, we use Markov chain Monte Carlo sampling (MCMC) via the `emcee` Python package [@foreman-mackey_emcee_2013].
+
+![Parametrized graph representation of the lymphatic network considering four LNLs.
+Blue, round nodes represent the hidden RVs, while orange square nodes show the observed RVs.
+Arcs represent a conditional probability parametrized with the quantity noted next to it.](../static/graph_with_obs.svg){#fig-graph-with-obs width="25%" fig-pos="th"}
 
@@
 
-Introducing this latent variable is helpful, because the mixture model likelihood is invariant w.r.t.
-permutations of the component labels.
-This introduces a number of problems, e.g.
+The latent variables are helpful in resolving the invariance of the likelihood w.r.t.
+permutations of the component labels, which may introduce problems, e.g.
 for common MCMC sampling methods.
-The latent variable $\mathbf{Z}$ allows us to derive two sets of interdependent equations that we may solve in an iterative fashion (see e.g.
+The $\mathbf{Z}$ allows us to derive two sets of interdependent equations that we may solve in an iterative fashion (see e.g.
 @bishop_pattern_2006 for a detailed derivation) that is commonly referred to as _expectation-maximization (EM)_ algorithm:
 
-The first are the probabilities of subcohort $i$ to belong to component $j$, given a set of parameters $\boldsymbol{\Psi}^\star$. These are often called the _responsibilities_:
+The first set are the probabilities of subcohort $i$ to belong to component $j$, given a set of parameters $\boldsymbol{\Psi}^\star$. These are often called the _responsibilities_:
 $$
@@
 From this, we can compute new mixing proportions $c_j^\star=\sum_{i=1}^s \gamma (z_{ij}) / s$ and then infer new parameters $\boldsymbol{\Psi}^\star$ -- e.g.
-via MCMC sampling -- from the resulting likelihood:
+via MCMC sampling -- from the resulting likelihood, which is the second set:
 
@@
 
+## Multi-Centric Data
+
+For the analyses in this work, we used five datasets from four different institutions:
+
+1. 287 oropharyngeal patients from the University of Zurich in Swizerland
+2. 263 oropharyngeal patients from the Centre Léon Bérard in Fance
+3. 289 oropharyngeal and oral cavity patients from the Inselspital Bern in Swizerland
+4. 239 oropharyngeal and oral cavity patients from the Centre Léon Bérard in Fance
+5. 162 oropharyngeal patients from the Hospital Vall d'Hebron in Spain
+
+The data comes in the form of CSV tables and are -- except for the last and most recent addition -- publicly available [@ludwig_detailed_2022;@ludwig_multi-centric_2023] and may be interactively explored in our **Ly**mphatic **Pro**gression e**X**plorer [LyProX](https://lyprox.org).
+Each row of these tables corresponds to one patient and details in which LNL metastatic involvement was found or not, according to different diagnostic and pathologic modalities.
+
 # Results {#sec-results}

### 2024-01-19 content/manuscript.qmd
update data, add preval plot, and add it to text

@@
 
+![Prevalence of LNL involvement stratified by subsite.
+The subsites are sorted in ascending order by their prevalence of involvement in LNL II.
+Oral cavity subsites are plotted in shaed of blue, oropharynx subsites in shades of orange.](../figures/prevalence_by_subsite.svg){#fig-prevalence-by-subsite fig-pos="b" width="49%"}
+
 The data comes in the form of CSV tables and are -- except for the last and most recent addition -- publicly available [@ludwig_detailed_2022;@ludwig_multi-centric_2023] and may be interactively explored in our **Ly**mphatic **Pro**gression e**X**plorer [LyProX](https://lyprox.org).

### 2024-01-19 scripts/prevalence_by_subsite.py

@@ -1,0 +1,126 @@
+"""
+Make a bar plot to visualize the prevalence of involvement in each of the four
+considered lymph node levels (LNLs), stratified by subsite (different bras) and by
+location (different colors).
+"""
+import argparse
+from pathlib import Path
+from typing import Generator
+import yaml
+
+import matplotlib.pyplot as plt
+from matplotlib.colors import to_rgba
+from tueplots import figsizes, fontsizes
+from lyscripts.utils import load_patient_data
+from lyscripts.plot.utils import COLORS as USZ
+
+
+SUBSITE = ("tumor", "1", "subsite")
+LOCATION = ("tumor", "1", "location")
+LNL_I = ("max_llh", "ipsi", "I")
+LNL_II = ("max_llh", "ipsi", "II")
+LNL_III = ("max_llh", "ipsi", "III")
+LNL_IV = ("max_llh", "ipsi", "IV")
+ICD_CODE_MAP = {
+    "C00": "lip",
+    "C01": "base of tongue",
+    "C02": "other/unspec. pt. of tongue",
+    "C03": "gum",
+    "C04": "floor of mouth",
+    "C05": "palate",
+    "C06": "other/unspec. pt. of mouth",
+    "C09": "tonsil",
+    "C10": "oropharynx",
+}
+OROPHARYNX_ICDS = ["C01", "C09", "C10"]
+
+
+def create_parser() -> argparse.ArgumentParser:
+    """Assemble the parser for the command line arguments."""
+    parser = argparse.ArgumentParser(description=__doc__)
+    parser.add_argument(
+        "-i", "--input", type=Path, default="data/enhanced.csv",
+        help="Path to the patient data file.",
+    )
+    parser.add_argument(
+        "-o", "--output", type=Path, default="figures/prevalence_by_subsite.png",
+        help="Path to the output file.",
+    )
+    parser.add_argument(
+        "-p", "--params", type=Path, default="params.yaml",
+        help="Path to the parameter file. Looks for a key `barplot_kwargs`.",
+    )
+    return parser
+
+
+def simplify_subsite(icd_code: str) -> str:
+    """Only use the part of the ICD code before the decimal point."""
+    return icd_code.split(".")[0]
+
+
+def make_readable(icd_code: str) -> str:
+    """Make the ICD code human-readable."""
+    return ICD_CODE_MAP[icd_code]
+
+
+def generate_location_colors(
[…]

### 2024-01-22 scripts/prevalence_by_subsite.py
add independent model sampling to pipeline

@@ -114,7 +114,7 @@ def main():
     ax.grid(axis="y")
     ax.set_xlabel("Lymph node level")
     ax.set_xticklabels(["I", "II", "III", "IV"], rotation=0)
-    ax.set_ylabel("Prevalence (%)")
+    ax.set_ylabel("Prevalence [%]")
     ax.legend(fontsize="x-small", labelspacing=0.3)
     ax.set_xlim(-0.46, 3.46)
 

### 2024-01-22 scripts/write_num_patients.py

@@ -1,0 +1,39 @@
+"""
+Write the number of patients in the data CSV file into the `_variables.yml` file.
+"""
+import argparse
+from pathlib import Path
+import yaml
+
+from lyscripts.utils import load_patient_data
+
+
+def create_parser() -> argparse.ArgumentParser:
+    """Assemble the parser for the command line arguments."""
+    parser = argparse.ArgumentParser(description=__doc__)
+    parser.add_argument(
+        "-i", "--input", type=Path, default="data/enhanced.csv",
+        help="Path to the patient data file.",
+    )
+    parser.add_argument(
+        "-o", "--output", type=Path, default="_variables.yml",
+        help="Path to the parameter file.",
+    )
+    return parser
+
+
+def main():
+    """Load data and create bar plot from it."""
+    args = create_parser().parse_args()
+    with open(args.output, "r") as file:
+        variables = yaml.safe_load(file)
+
+    patient_data = load_patient_data(args.input)
+    variables["data"]["num_patients"] = len(patient_data)
+
+    with open(args.output, "w") as file:
+        yaml.dump(variables, file)
+
+
+if __name__ == "__main__":
+    main()

### 2024-01-22 content/manuscript.qmd
add & correct variable access in manuscript

@@
 where $T \in \{ \text{early}, \text{advanced} \}$ denotes the T-category.
-We fix $P_\text{early}(t)$ to a binomial distribution with parameters $n=t_\text{max}=$ {{< var t_max >}} and $p_\text{early}=$ {{< var time-prior.early.p >}}, while the advanced T-category's time-prior is also a binomial distribution where the $p_\text{advanced}$ parameter is learned.
+We fix $P_\text{early}(t)$ to a binomial distribution with parameters $n=t_\text{max}=$ {{< var model.max_t >}} and $p_\text{early}=$ {{< var model.first_binom_prob >}}, while the advanced T-category's time-prior is also a binomial distribution where the $p_\text{advanced}$ parameter is learned.
 
@@
 
+In @fig-prevalence-by-subsite, we have plotted the prevalence of involvement in the four LNLs I, II, III, and IV stratified by the primary tumor's subsite.
+We only included patients with tumors in the gum, floor of mouth, other/unspecified parts of the mouth, palate, oropharynx, tosil, base of tongue, and other/unspecified parts of tongue, resulting in {{< var data.num_patients >}} patients.
+
 # Results {#sec-results}

### 2024-01-22 content/manuscript.qmd
add quarto render to pipeline

@@
 In @fig-prevalence-by-subsite, we have plotted the prevalence of involvement in the four LNLs I, II, III, and IV stratified by the primary tumor's subsite.
-We only included patients with tumors in the gum, floor of mouth, other/unspecified parts of the mouth, palate, oropharynx, tosil, base of tongue, and other/unspecified parts of tongue, resulting in {{< var data.num_patients >}} patients.
+We only included patients with tumors in the gum, floor of mouth, other/unspecified parts of the mouth, palate, oropharynx, tosil, base of tongue, and other/unspecified parts of tongue, resulting in {{< var num_patients >}} patients.
 

### 2024-01-22 scripts/write_num_patients.py

@@ -29,7 +29,7 @@ def main():
         variables = yaml.safe_load(file)
 
     patient_data = load_patient_data(args.input)
-    variables["data"]["num_patients"] = len(patient_data)
+    variables["num_patients"] = len(patient_data)
 
     with open(args.output, "w") as file:
         yaml.dump(variables, file)

### 2024-01-22 content/manuscript.qmd
rename some var keys & rerun pipeline

@@
 In @fig-prevalence-by-subsite, we have plotted the prevalence of involvement in the four LNLs I, II, III, and IV stratified by the primary tumor's subsite.
-We only included patients with tumors in the gum, floor of mouth, other/unspecified parts of the mouth, palate, oropharynx, tosil, base of tongue, and other/unspecified parts of tongue, resulting in {{< var num_patients >}} patients.
+We only included patients with tumors in the gum, floor of mouth, other/unspecified parts of the mouth, palate, oropharynx, tosil, base of tongue, and other/unspecified parts of tongue, resulting in {{< var num_patients.total >}} patients.
 

### 2024-01-22 scripts/prevalence_by_subsite.py

@@ -21,17 +21,6 @@ LNL_I = ("max_llh", "ipsi", "I")
 LNL_II = ("max_llh", "ipsi", "II")
 LNL_III = ("max_llh", "ipsi", "III")
 LNL_IV = ("max_llh", "ipsi", "IV")
-ICD_CODE_MAP = {
-    "C00": "lip",
-    "C01": "base of tongue",
-    "C02": "other/unspec. pt. of tongue",
-    "C03": "gum",
-    "C04": "floor of mouth",
-    "C05": "palate",
-    "C06": "other/unspec. pt. of mouth",
-    "C09": "tonsil",
-    "C10": "oropharynx",
-}
 OROPHARYNX_ICDS = ["C01", "C09", "C10"]
 
 
@@ -56,11 +45,6 @@ def create_parser() -> argparse.ArgumentParser:
 def simplify_subsite(icd_code: str) -> str:
     """Only use the part of the ICD code before the decimal point."""
     return icd_code.split(".")[0]
-
-
-def make_readable(icd_code: str) -> str:
-    """Make the ICD code human-readable."""
-    return ICD_CODE_MAP[icd_code]
 
 
 def generate_location_colors(
@@ -92,6 +76,7 @@ def main():
         with open(args.params) as f:
             params = yaml.safe_load(f)
         barplot_kwargs.update(params.get("barplot_kwargs", {}))
+        icd_code_map = params.get("icd_code_map", {})
 
     patient_data = load_patient_data(args.input)
     patient_data[SUBSITE] = patient_data[SUBSITE].apply(simplify_subsite)
@@ -100,7 +85,7 @@ def main():
         values=[LNL_I, LNL_II, LNL_III, LNL_IV]
     ).sort_values(by=LNL_II)
     colors = list(generate_location_colors(pivot_table.index))
-    pivot_table.index = pivot_table.index.map(make_readable)
+    pivot_table.index = pivot_table.index.map(icd_code_map)
 
     plt.rcParams.update(figsizes.icml2022_half())
     plt.rcParams.update(fontsizes.icml2022())

### 2024-01-22 scripts/write_num_patients.py

@@ -8,6 +8,9 @@ import yaml
 from lyscripts.utils import load_patient_data
 
 
+SUBSITE = ("tumor", "1", "subsite")
+
+
 def create_parser() -> argparse.ArgumentParser:
     """Assemble the parser for the command line arguments."""
     parser = argparse.ArgumentParser(description=__doc__)
@@ -29,7 +32,15 @@ def main():
         variables = yaml.safe_load(file)
 
     patient_data = load_patient_data(args.input)
-    variables["num_patients"] = len(patient_data)
+    icd_code_map = variables.get("icd_code_map", {})
+    patient_info = variables.get("num_patients", {})
+    patient_info["total"] = len(patient_data)
+
+    for icd_code in icd_code_map.keys():
+        num_icd_code = int(patient_data[SUBSITE].str.contains(icd_code).sum())
+        patient_info[icd_code] = num_icd_code
+
+    variables.update({"num_patients": patient_info})
 
     with open(args.output, "w") as file:
         yaml.dump(variables, file)

### 2024-01-22 scripts/lymixture/__init__.py
add lymixture code & start refactoring

@@ -1,0 +1,4 @@
+"""
+In this subpackage, I put only those functions and things that are necessary for
+Julian's EM implementation.
+"""

### 2024-01-22 scripts/lymixture/em_sampling.py

@@ -1,0 +1,693 @@
+import emcee
+from tqdm import tqdm
+
+from typing import Optional, List
+import logging
+import numpy as np
+from scipy.optimize import minimize
+import matplotlib.pyplot as plt
+from matplotlib.ticker import MaxNLocator
+
+from lyscripts.plot.utils import save_figure, get_size
+from scripts.lymixture.types import EMConfigType
+from lyscripts.sample import sample_from_global_model_and_configs
+
+global MODELS, N_CLUSTERS
+
+logger = logging.getLogger(__name__)
+
+
+def emcee_simple_sampler(
+    log_prob_fn: callable,
+    ndim: int,
+    sampling_params: dict,
+    hdf5_backend: Optional[emcee.backends.HDFBackend] = None,
+    starting_point: Optional[np.ndarray] = None,
+    llh_args: Optional[List] = None,
+    show_progress=True,
+):
+    """A simple sampler."""
+    nwalkers = sampling_params["walkers_per_dim"] * ndim
+    burnin = sampling_params["nburnin"]
+    nstep = sampling_params["nsteps"]
+
+    if starting_point is None:
+        starting_point = np.random.uniform(size=(nwalkers, ndim)) / 10
+    else:
+        if np.shape(starting_point) != np.shape(
+            np.random.uniform(size=(nwalkers, ndim))
+        ):
+            starting_point = np.tile(starting_point, (nwalkers, 1))
+
+    original_sampler_mp = emcee.EnsembleSampler(
+        nwalkers,
+        ndim,
+        log_prob_fn,
+        args=llh_args,
+        backend=hdf5_backend,
+    )
+    sampling_results = original_sampler_mp.run_mcmc(
+        initial_state=starting_point,
+        nsteps=nstep + burnin,
+        progress=show_progress,
+    )
+
+    ar = np.mean(original_sampler_mp.acceptance_fraction)
+    print(f"Accepted {ar * 100 :.2f} % of samples.")
+    samples = original_sampler_mp.get_chain(flat=True, discard=burnin)
+    log_probs = original_sampler_mp.get_log_prob(flat=True)
+    end_point = original_sampler_mp.get_last_sample()[0]
+
+    return samples, end_point, log_probs
+
+
+def exceed_param_bound(p: np.ndarray) -> bool:
+    """Checks if any element in p exceeds the range [0,1]"""
+    for ps in p:
+        if ps <= 0 or 1 <= ps:
+            return True
+    return False
+
+
[…]

### 2024-01-22 scripts/lymixture/mixture_model.py

@@ -1,0 +1,686 @@
+from functools import cached_property
+import logging
+import os
+import random
+from typing import Any, Dict, List, Optional
+from pathlib import Path
+
+import numpy as np
+import pandas as pd
+import emcee
+import lymph
+from lyscripts.sample import sample_from_global_model_and_configs
+
+from scripts.lymixture.em_sampling import (
+    ExpectationMaximization,
+    History,
+    emcee_simple_sampler,
+    exceed_param_bound,
+)
+from scripts.lymixture.model_functions import (
+    compute_state_probability_matrices,
+    compute_cluster_assignment_matrix,
+    compute_cluster_state_probabilty_matrices,
+    create_diagnose_matrices,
+)
+from scripts.lymixture.mm_predict import (
+    _create_obs_pred_df_single,
+    create_obs_pred_df_single,
+    mm_generate_predicted_prevalences,
+    mm_predicted_risk,
+)
+from mm_plotting import plot_cluster_assignments, plot_cluster_parameters
+
+
+# Define a global variable which can be used within this module.
+global LMM_GLOBAL
+
+
+def get_param_labels_temp(model):
+    """
+    Temporary method to get parameter labels from a model.
+
+    Args:
+        model: Lymph model instance.
+
+    Returns:
+        list: List of parameter labels.
+    """
+    return [
+        t.replace("primary", "T").replace("_spread", "")
+        for t in model.get_params(as_dict=True).keys()
+    ]
+
+
+def as_list(val):
+    """Return val as list if it is not already a list"""
+    if not isinstance(val, list):
+        return [val]
+    return val
+
+
+def log_ll_cl_parameters(cluster_parameters):
+    if exceed_param_bound(cluster_parameters):
+        return -np.inf
+    llh = LMM_GLOBAL.mm_hmm_likelihood(cluster_parameters=cluster_parameters)
+    if np.isinf(llh):
+        return -np.inf
+    return llh
+
+
+class LymphMixtureModel:
+    """
+    Wrapper model which handles the mixture components (clusters) in the lymph models.
+
+    Args:
+        lymph_models (list): List of lymph models.
+        n_clusters (int): Number of clusters.
+        base_dir (Path): Base directory for saving outputs.
+        name (str, optional): Name for the model. Defaults to 'LMM'.
+    """
+    def __init__(
+        self,
+        lymph_model: lymph.models.Unilateral,
+        n_clusters: int,
+        n_subpopulation: int,
[…]

### 2024-01-22 scripts/lymixture/mm_predict.py

@@ -1,0 +1,350 @@
+import random
+from typing import Dict, Generator, List, Optional
+
+import pandas as pd
+import numpy as np
+from lyscripts.predict.risks import predicted_risk
+from lyscripts.predict.utils import complete_pattern
+import lymph
+from lyscripts.predict.prevalences import (
+    create_patient_row,
+    compute_observed_prevalence,
+    generate_predicted_prevalences,
+    compute_predicted_prevalence_for_mixture,
+)
+
+from scripts.lymixture.mixture_model import LymphMixtureModel
+
+## Prevalences
+
+def mm_generate_predicted_prevalences(
+    cluster_assignments: np.ndarray,
+    cluster_parameters: np.ndarray,
+    pattern: Dict[str, Dict[str, bool]],
+    model: lymph.models.Unilateral,
+    t_stage: str = "early",
+    modality_spsn: Optional[List[float]] = None,
+    invert: bool = False,
+    **_kwargs,
+) -> Generator[float, None, None]:
+    """Wrapper for the `lyscript predict` function."""
+
+    # The number of clusters are extracted from the length of cluster assignments given:
+    n_clusters = len(cluster_assignments)
+    n_p = len(model.get_params())
+
+    cluster_predictions = np.zeros(shape=(n_clusters))
+    for k in range(n_clusters):
+        k_cluster_parameters = cluster_parameters[:, k * n_p : (k + 1) * n_p]
+        cluster_predictions[k] = next(
+            generate_predicted_prevalences(
+                pattern,
+                model,
+                k_cluster_parameters,
+                t_stage=t_stage,
+                modality_spsn=modality_spsn,
+                invert=invert,
+                **_kwargs,
+            )
+        )
+    return cluster_assignments @ cluster_predictions
+
+
+## Risks
+
+
+def mm_predicted_risk(
+    involvement: Dict[str, Dict[str, bool]],
+    model: lymph.models.Unilateral,
+    cluster_assignment: np.ndarray,
+    cluster_parameters: np.ndarray,
+    t_stage: str,
+    midline_ext: bool = False,
+    given_diagnosis: Optional[Dict[str, Dict[str, bool]]] = None,
+    given_diagnosis_spsn: Optional[List[float]] = None,
+    invert: bool = False,
+    **_kwargs,
+) -> Generator[float, None, None]:
+    """Wrapper for the `lyscript predicted_risk` function."""
+
+    if midline_ext is not False:
+        raise NotImplementedError
+    # The number of clusters are extracted from the length of cluster assignments given:
[…]

### 2024-01-22 scripts/lymixture/types.py

@@ -1,0 +1,29 @@
+from typing import TypedDict, Callable, Dict
+
+
+class EStep(TypedDict):
+    method: str
+    walkers_per_dim: int
+    nsteps: int
+    nburnin: int
+    sampler: str
+    show_progress: bool
+
+
+class MStep(TypedDict):
+    minimize_method: str
+    imputation_function: Callable[[int], int]  # Function type
+
+
+class Convergence(TypedDict):
+    criterion: str
+    default: Dict[int, float]
+
+
+class EMConfigType(TypedDict):
+    max_steps: int  # Max steps until force exit.
+    method: str  # DEFAULT or INVERTED method.
+    verbose: bool
+    e_step: EStep
+    m_step: MStep
+    convergence: Convergence

### 2024-01-22 scripts/lymixture/utils.py

@@ -1,0 +1,667 @@
+from cycler import cycler
+import pandas as pd
+import lymph
+import numpy as np
+import emcee
+import multiprocess as mp
+import os
+from lyscripts.predict.prevalences import (
+    compute_observed_prevalence,
+)
+import matplotlib.colors as mcolors
+from pathlib import Path
+from scipy.special import factorial
+import itertools, sys
+import scipy as sp
+import matplotlib.pyplot as plt
+
+# Plotting styles
+usz_blue = "#005ea8"
+usz_blue_border = "#2387D5"
+usz_green = "#00afa5"
+usz_green_border = "#30DFD5"
+usz_red = "#ae0060"
+usz_red_border = "#FB4EAE"
+usz_orange = "#f17900"
+usz_orange_border = "#F8AB5C"
+usz_gray = "#c5d5db"
+usz_gray_border = "#DFDFDF"
+usz_colors = [usz_blue, usz_green, usz_red, usz_orange, usz_gray]
+edge_colors = [
+    usz_blue_border,
+    usz_green_border,
+    usz_red_border,
+    usz_orange_border,
+    usz_gray_border,
+]
+COLORS = {
+    "blue": "#005ea8",
+    "orange": "#f17900",
+    "green": "#00afa5",
+    "red": "#ae0060",
+    "gray": "#c5d5db",
+}
+
+PLOT_PATH = Path("./figures/")
+
+
+# Define a function to set plot size based on desired width, unit, and ratio
+def set_size(width="single", unit="cm", ratio="golden"):
+    """
+    Set the size of the plot for Matplotlib.
+
+    Parameters:
+    - width (int or str): The width of the plot.
+    - unit (str): The unit of width ('cm', 'inch', etc.).
+    - ratio (float or str): The aspect ratio of the plot.
+
+    Returns:
+    - tuple: A tuple (width, height) specifying the plot size.
+    """
+    if width == "single":
+        width = 10
+    elif width == "full":
+        width = 16
+    else:
+        try:
+            width = width
+        except:
+            width = 10
+
+    if unit == "cm":
+        width = width / 2.54
+
+    if ratio == "golden":
+        ratio = 1.618
+    else:
+        ratio = ratio
+
+    try:
+        height = width / ratio
+    except:
+        height = width / 1.618
+
+    return (width, height)
+
+
+def binom_pmf(k: np.ndarray, n: int, p: float):
+    """Binomial PMF"""
[…]

### 2024-01-23 scripts/helpers.py
set up lymixture

@@ -1,0 +1,7 @@
+"""
+Helper functions for the scripts.
+"""
+
+def simplify_subsite(icd_code: str) -> str:
+    """Only use the part of the ICD code before the decimal point."""
+    return icd_code.split(".")[0]

### 2024-01-23 scripts/lymixture/__init__.py

@@ -2,3 +2,7 @@
 In this subpackage, I put only those functions and things that are necessary for
 Julian's EM implementation.
 """
+import logging
+
+logger = logging.getLogger(__name__)
+logger.addHandler(logging.NullHandler())

### 2024-01-23 scripts/lymixture/em_sampling.py

@@ -9,8 +9,8 @@ import matplotlib.pyplot as plt
 from matplotlib.ticker import MaxNLocator
 
 from lyscripts.plot.utils import save_figure, get_size
-from scripts.lymixture.types import EMConfigType
-from lyscripts.sample import sample_from_global_model_and_configs
+from lymixture.types import EMConfigType
+# from lyscripts.sample import sample_from_global_model_and_configs
 
 global MODELS, N_CLUSTERS
 
@@ -53,7 +53,7 @@ def emcee_simple_sampler(
     )
 
     ar = np.mean(original_sampler_mp.acceptance_fraction)
-    print(f"Accepted {ar * 100 :.2f} % of samples.")
+    logger.info(f"Accepted {ar * 100 :.2f} % of samples.")
     samples = original_sampler_mp.get_chain(flat=True, discard=burnin)
     log_probs = original_sampler_mp.get_log_prob(flat=True)
     end_point = original_sampler_mp.get_last_sample()[0]
@@ -284,7 +284,7 @@ class ExpectationMaximization:
 
 
     def run_em(self):
-        print(f"Run EM with method {self.em_config['method']} for max {self.max_steps} steps.")
+        logger.info(f"Run EM with method {self.em_config['method']} for max {self.max_steps} steps.")
         self.init_run_em()
 
         # Run the EM algorithm.
@@ -313,19 +313,19 @@ class ExpectationMaximization:
             self.history.save(save_dir=self.em_dir)
 
             if converged:
-                print(f"Step {self.current_step} / {self.max_steps}: Converged!")
+                logger.info(f"Step {self.current_step} / {self.max_steps}: Converged!")
                 break
 
             self.current_step = iteration
 
         if not converged:
-            print("Max steps reached, no convergence, return current approximation.")
+            logger.info("Max steps reached, no convergence, return current approximation.")
 
         return self.current_cluster_assignments, self.history
 
 
     def e_step(self):
-        print(f"Step {self.current_step}: Perform expectation.")
+        logger.info(f"Step {self.current_step}: Perform expectation.")
         # Set the cluster parameters to the current estimates.
         self.lmm.cluster_parameters = self.current_cluster_parameters
         # self.lmm.diagnose_matrices
@@ -340,11 +340,11 @@ class ExpectationMaximization:
             self.lmm.n_cluster_assignments,
             initial_guess=self.emcee_end_point,
         )
-        # print(cluster_assignments_posterior)
+        # logger.info(cluster_assignments_posterior)
         # raise
[…]

### 2024-01-23 scripts/lymixture/mixture_model.py

@@ -9,27 +9,34 @@ import numpy as np
 import pandas as pd
 import emcee
 import lymph
-from lyscripts.sample import sample_from_global_model_and_configs
+# from lyscripts.sample import sample_from_global_model_and_configs
 
-from scripts.lymixture.em_sampling import (
+from lymixture.em_sampling import (
     ExpectationMaximization,
     History,
     emcee_simple_sampler,
     exceed_param_bound,
 )
-from scripts.lymixture.model_functions import (
+from lymixture.model_functions import (
     compute_state_probability_matrices,
     compute_cluster_assignment_matrix,
     compute_cluster_state_probabilty_matrices,
     create_diagnose_matrices,
 )
-from scripts.lymixture.mm_predict import (
+from lymixture.mm_predict import (
     _create_obs_pred_df_single,
     create_obs_pred_df_single,
     mm_generate_predicted_prevalences,
     mm_predicted_risk,
 )
-from mm_plotting import plot_cluster_assignments, plot_cluster_parameters
+from lymixture.mm_plotting import (
+    plot_cluster_assignments,
+    plot_cluster_parameters,
+)
+# pylint: disable=logging-fstring-interpolation
+
+
+logger = logging.getLogger(__name__)
 
 
 # Define a global variable which can be used within this module.
@@ -87,13 +94,10 @@ class LymphMixtureModel:
         name: Optional[str] = None,
         **_kwargs,
     ):
-        # Initialize logger
-        self._setup_logger()
-
         # Initialize model configurations
         self.lymph_model = lymph_model
         if not isinstance(lymph_model, lymph.models.Unilateral):
-            self.logger.error(
+            logger.error(
                 "Mixture model is only implemented for Unilateral cluster model."
             )
             raise NotImplementedError
@@ -110,8 +114,9 @@ class LymphMixtureModel:
         self.base_dir = base_dir.joinpath(self.name)
         self._create_directories()
 
-        self.logger.info(
-            f"Create LymphMixtureModel of type {type(self.lymph_model)} with {self.n_clusters} clusters in {self.base_dir}"
+        logger.info(
+            f"Create LymphMixtureModel of type {type(self.lymph_model)} with "
+            f"{self.n_clusters} clusters in {self.base_dir}"
         )
 
     def _create_directories(self):
@@ -126,16 +131,11 @@ class LymphMixtureModel:
         self.figures_dir.mkdir(parents=True, exist_ok=True)
[…]

### 2024-01-23 scripts/lymixture/mm_predict.py

@@ -10,10 +10,10 @@ from lyscripts.predict.prevalences import (
     create_patient_row,
     compute_observed_prevalence,
     generate_predicted_prevalences,
-    compute_predicted_prevalence_for_mixture,
+    # compute_predicted_prevalence_for_mixture,
 )
 
-from scripts.lymixture.mixture_model import LymphMixtureModel
+# from lymixture.mixture_model import LymphMixtureModel
 
 ## Prevalences
 
@@ -122,7 +122,7 @@ def _create_obs_pred_df_single(
                 t_stage=t_stage,
                 lnls=lnls,
             )
-            # print(f"{loc}: Observed P({lnl}) = {round(op[0]/op[1], 2)} ({op})")
+            # logger.info(f"{loc}: Observed P({lnl}) = {round(op[0]/op[1], 2)} ({op})")
             pp_list = list(
                 generate_predicted_prevalences(
                     pattern={"ipsi": pattern},
@@ -146,7 +146,7 @@ def _create_obs_pred_df_single(
                 "t": op[1],
             }
 
-            # print(f"{loc}: Predicted P({lnl}) = {round(pp_list.mean(), 2)}")
+            # logger.info(f"{loc}: Predicted P({lnl}) = {round(pp_list.mean(), 2)}")
         obs_prev[t_stage] = obs_prev_t
         pred_prev[t_stage] = pred_prev_t
         pred_prev_std[t_stage] = pred_prev_std_t
@@ -190,63 +190,63 @@ def _create_obs_pred_df_single(
     return df, [obs_prev, pred_prev, pred_prev_std, pred_prev_list]
 
 
-def generate_predicted_prevalences_for_mixture(
-    pattern: Dict[str, Dict[str, bool]],
-    model: lymph.models.Unilateral,
-    n_clusters: int,
-    cluster_assignment: np.ndarray,
-    samples: np.ndarray,
-    t_stage: str = "early",
-    modality_spsn: Optional[List[float]] = None,
-    invert: bool = False,
-    **_kwargs,
-) -> Generator[float, None, None]:
-    """Compute the prevalence of a given `pattern` of lymphatic progression using a
-    `model` and trained `samples`.
+# def generate_predicted_prevalences_for_mixture(
+#     pattern: Dict[str, Dict[str, bool]],
+#     model: lymph.models.Unilateral,
+#     n_clusters: int,
+#     cluster_assignment: np.ndarray,
+#     samples: np.ndarray,
[…]

### 2024-01-23 scripts/lymixture/utils.py

@@ -199,10 +199,10 @@ def create_prev_vectors(
         if ax is None:
             plt.show()
     if len(states_all) > 4:
-        print(f"Prev Vector: {X_inv_list}")
+        logger.info(f"Prev Vector: {X_inv_list}")
     else:
         prev_formatted = "".join(f"{l}: {p}" for l, p in zip(states_all, X_inv_list))
-        print(f"Prev Vector: {prev_formatted}")
+        logger.info(f"Prev Vector: {prev_formatted}")
     return X_inv_list
 
 
@@ -243,7 +243,7 @@ def convert_params(params, n_clusters, n_subsites):
 def emcee_sampling(llh_function, n_params, sample_name, llh_args=None):
     nwalkers, nstep, burnin = 20 * n_params, 1000, 1500
     thin_by = 1
-    print(f"Dimension: {n_params} with n walkers: {nwalkers}")
+    logger.info(f"Dimension: {n_params} with n walkers: {nwalkers}")
     output_name = sample_name
 
     if False:
@@ -252,7 +252,7 @@ def emcee_sampling(llh_function, n_params, sample_name, llh_args=None):
         created_pool = mp.Pool(os.cpu_count())
         with created_pool as pool:
             starting_points = np.random.uniform(size=(nwalkers, n_params))
-            print(
+            logger.info(
                 f"Start Burning (steps = {burnin}) with {created_pool._processes} cores"
             )
             burnin_sampler = emcee.EnsembleSampler(
@@ -267,11 +267,11 @@ def emcee_sampling(llh_function, n_params, sample_name, llh_args=None):
             )
 
             ar = np.mean(burnin_sampler.acceptance_fraction)
-            print(
+            logger.info(
                 f"the HMM sampler for model 01 accepted {ar * 100 :.2f} % of samples."
             )
             last_sample = burnin_sampler.get_last_sample()[0]
-            print(f"The shape of the last sample is {last_sample.shape}")
+            logger.info(f"The shape of the last sample is {last_sample.shape}")
             starting_points = np.random.uniform(size=(nwalkers, n_params))
             original_sampler_mp = emcee.EnsembleSampler(
                 nwalkers,
@@ -286,7 +286,7 @@ def emcee_sampling(llh_function, n_params, sample_name, llh_args=None):
             )
 
         ar = np.mean(original_sampler_mp.acceptance_fraction)
-        print(f"the HMM sampler for model accepted {ar * 100 :.2f} % of samples.")
[…]

### 2024-01-23 scripts/mixture_model_sampling.py

@@ -1,0 +1,76 @@
+"""
+Perform an EM-like sampling round to infer parameters and mixture components of a
+n-component mixture of lymphatic progression models for k different tumor subsites.
+"""
+import argparse
+from pathlib import Path
+import numpy as np
+import yaml
+
+from lyscripts.utils import (
+    load_patient_data,
+    create_model_from_config,
+)
+
+from helpers import simplify_subsite
+from lymixture.mixture_model import LymphMixtureModel
+
+
+SUBSITE = ("tumor", "1", "subsite")
+SIMPLE_SUBSITE = ("tumor", "1", "simple_subsite")
+
+
+def create_parser() -> argparse.ArgumentParser:
+    """Assemble the parser for the command line arguments."""
+    parser = argparse.ArgumentParser(description=__doc__)
+    parser.add_argument(
+        "-i", "--input", type=Path, default="data/enhanced.csv",
+        help="Path to the patient data file.",
+    )
+    parser.add_argument(
+        "-m", "--models", type=Path, default="models/",
+        help="Directory for sampled models.",
+    )
+    parser.add_argument(
+        "-p", "--params", type=Path, default="_variables.yml",
+        help="Path to the parameter file. Looks for a key `barplot_kwargs`.",
+    )
+    return parser
+
+
+def main():
+    """Execute the main routine."""
+    args = create_parser().parse_args()
+
+    with open(args.params, mode="r", encoding="utf-8") as f:
+        params = yaml.safe_load(f)
+
+    patient_data = load_patient_data(args.input)
+    patient_data[SIMPLE_SUBSITE] = patient_data[SUBSITE].apply(simplify_subsite)
+    num_subsites = len(patient_data[SIMPLE_SUBSITE].unique())
+    lymph_model = create_model_from_config(params)
+    lymph_model.modalities = {"max_llh": [1., 1.]}
+    lymph_model.diag_time_dists = {"all": lymph_model.diag_time_dists["early"]}
+
+    mixture_model = LymphMixtureModel(
+        lymph_model=lymph_model,
+        n_clusters=params["mixture_model"]["num_clusters"],
+        n_subpopulation=num_subsites,
+        name="test",
+    )
+    mixture_model.load_data(
+        patient_data=patient_data,
+        split_by=SIMPLE_SUBSITE,
+        mapping=lambda x: "all",
+    )
+    mixture_model.cluster_assignments = np.zeros(shape=(
[…]

### 2024-01-23 scripts/prevalence_by_subsite.py

@@ -14,6 +14,8 @@ from tueplots import figsizes, fontsizes
 from lyscripts.utils import load_patient_data
 from lyscripts.plot.utils import COLORS as USZ
 
+from scripts.helpers import simplify_subsite
+
 
 SUBSITE = ("tumor", "1", "subsite")
 LOCATION = ("tumor", "1", "location")
@@ -40,11 +42,6 @@ def create_parser() -> argparse.ArgumentParser:
         help="Path to the parameter file. Looks for a key `barplot_kwargs`.",
     )
     return parser
-
-
-def simplify_subsite(icd_code: str) -> str:
-    """Only use the part of the ICD code before the decimal point."""
-    return icd_code.split(".")[0]
 
 
 def generate_location_colors(

### 2024-01-23 scripts/lymixture/em_sampling.py
add modified lymixture em run to pipeline

@@ -190,9 +190,6 @@ class ExpectationMaximization:
             self.current_cluster_parameters, self.current_cluster_assignments
         )
 
-        self.em_dir = lmm.base_dir / "EM"
-        self.em_dir.mkdir(parents=True, exist_ok=True)
-
 
     def mcmc_sampling(self, log_prob_fn, ndim, llh_args=None, initial_guess=None):
         """
@@ -308,9 +305,6 @@ class ExpectationMaximization:
                 self.current_likelihood,
                 self.convergence.current_convergence_values,
             )
-
-            # Store the history object in each round
-            self.history.save(save_dir=self.em_dir)
 
             if converged:
                 logger.info(f"Step {self.current_step} / {self.max_steps}: Converged!")
@@ -562,8 +556,8 @@ class History:
 
         for i, label in enumerate(label_ts):
             ax.plot(
-                x=range(len(self.convergence_value)),
-                y=[v[i] for v in self.convergence_value],
+                range(len(self.convergence_value)),
+                [v[i] for v in self.convergence_value],
                 label=label,
             )
         ax.set_xlabel("Steps")

### 2024-01-23 scripts/lymixture/mixture_model.py

@@ -90,8 +90,7 @@ class LymphMixtureModel:
         lymph_model: lymph.models.Unilateral,
         n_clusters: int,
         n_subpopulation: int,
-        base_dir: Optional[Path] = None,
-        name: Optional[str] = None,
+        hdf5_output: str | Path | None = None,
         **_kwargs,
     ):
         # Initialize model configurations
@@ -104,32 +103,17 @@ class LymphMixtureModel:
         self.n_clusters = n_clusters
         self.n_subpopulation = n_subpopulation
 
+        if hdf5_output is None:
+            hdf5_output = Path("models/mixture.hdf5")
+        self.hdf5_output = Path(hdf5_output)
+
         # Compute number of parameters expected by the model
         self.compute_expected_n()
 
-        # Set up directories
-        self.name = name if name else "LMM"
-        if base_dir is None:
-            base_dir = Path(os.getcwd())
-        self.base_dir = base_dir.joinpath(self.name)
-        self._create_directories()
-
         logger.info(
             f"Create LymphMixtureModel of type {type(self.lymph_model)} with "
-            f"{self.n_clusters} clusters in {self.base_dir}"
+            f"{self.n_clusters} clusters."
         )
-
-    def _create_directories(self):
-        """
-        Creates necessary directories for saving outputs.
-        """
-        self.samples_dir = self.base_dir.joinpath("samples/")
-        self.figures_dir = self.base_dir.joinpath("figures/")
-        self.predictions_dir = self.base_dir.joinpath("predictions/")
-        self.base_dir.mkdir(parents=True, exist_ok=True)
-        self.samples_dir.mkdir(parents=True, exist_ok=True)
-        self.figures_dir.mkdir(parents=True, exist_ok=True)
-        self.predictions_dir.mkdir(parents=True, exist_ok=True)
 
 
     def delete_cached_property(self, property: str):
@@ -355,72 +339,46 @@ class LymphMixtureModel:
         return estimated_cluster_assignments, em_history
 
 
-    def _mcmc_sampling(
-        self,
-        mcmc_config: dict,
-        save_samples: bool = True,
-        force_resampling: bool = False,
-    ):
+    def _mcmc_sampling(self, mcmc_config: dict):
         """
         Performs MCMC sampling to determine model parameters for the current cluster
         assignments.
         """
         global LMM_GLOBAL
-
         LMM_GLOBAL = self
+
[…]

### 2024-01-23 scripts/mixture_model_sampling.py

@@ -7,6 +7,8 @@ from pathlib import Path
 import numpy as np
 import yaml
 
+import matplotlib.pyplot as plt
+import h5py
 from lyscripts.utils import (
     load_patient_data,
     create_model_from_config,
@@ -28,8 +30,12 @@ def create_parser() -> argparse.ArgumentParser:
         help="Path to the patient data file.",
     )
     parser.add_argument(
-        "-m", "--models", type=Path, default="models/",
-        help="Directory for sampled models.",
+        "-m", "--model", type=Path, default="models/mixture.hdf5",
+        help="Path for sampled mixture model.",
+    )
+    parser.add_argument(
+        "-f", "--figure", type=Path, default="figures/mixture_history.png",
+        help="Path for figure over EM history.",
     )
     parser.add_argument(
         "-p", "--params", type=Path, default="_variables.yml",
@@ -44,6 +50,9 @@ def main():
 
     with open(args.params, mode="r", encoding="utf-8") as f:
         params = yaml.safe_load(f)
+        params["mixture_model"]["em_config"]["m_step"].update({
+            "imputation_function": lambda x: int(10 / 6 * x**1.1 +1)
+        })
 
     patient_data = load_patient_data(args.input)
     patient_data[SIMPLE_SUBSITE] = patient_data[SUBSITE].apply(simplify_subsite)
@@ -56,7 +65,7 @@ def main():
         lymph_model=lymph_model,
         n_clusters=params["mixture_model"]["num_clusters"],
         n_subpopulation=num_subsites,
-        name="test",
+        hdf5_output=args.model,
     )
     mixture_model.load_data(
         patient_data=patient_data,
@@ -69,7 +78,22 @@ def main():
     mixture_model.cluster_parameters = np.random.uniform(
         size=mixture_model.n_cluster_parameters
     )
-    print(mixture_model.mm_hmm_likelihood())
+
+    _chain, cluster_components, history = mixture_model.fit(
+        em_config=params["mixture_model"]["em_config"],
+        mcmc_config=params["mixture_model"]["mcmc_config"],
+    )
+    with h5py.File(args.model, mode="a") as h5_file:
+        if "em/cluster_components" in h5_file:
+            del h5_file["em/cluster_components"]
+        h5_file.create_dataset("em/cluster_components", data=cluster_components)
+    history.plot_history(
+        mixture_model.subpopulation_labels,
[…]

### 2024-01-23 scripts/prevalence_by_subsite.py

@@ -14,7 +14,7 @@ from tueplots import figsizes, fontsizes
 from lyscripts.utils import load_patient_data
 from lyscripts.plot.utils import COLORS as USZ
 
-from scripts.helpers import simplify_subsite
+from helpers import simplify_subsite
 
 
 SUBSITE = ("tumor", "1", "subsite")

### 2024-01-24 scripts/lymixture/em_sampling.py
use inverted sampling method

@@ -10,6 +10,7 @@ from matplotlib.ticker import MaxNLocator
 
 from lyscripts.plot.utils import save_figure, get_size
 from lymixture.types import EMConfigType
+from lymixture.utils import sample_from_global_model_and_configs
 # from lyscripts.sample import sample_from_global_model_and_configs
 
 global MODELS, N_CLUSTERS
@@ -212,9 +213,7 @@ class ExpectationMaximization:
                 ndim,
                 sampling_params=self.em_config["e_step"],
                 starting_point=initial_guess,
-                save_dir=self.em_dir,
                 models=LMM,
-                llh_args=llh_args,
                 verbose=self.em_config["e_step"]["show_progress"]
                 & self.em_config["verbose"],
             )

### 2024-01-24 scripts/lymixture/mixture_model.py

@@ -33,6 +33,7 @@ from lymixture.mm_plotting import (
     plot_cluster_assignments,
     plot_cluster_parameters,
 )
+from lymixture.utils import sample_from_global_model_and_configs
 # pylint: disable=logging-fstring-interpolation
 
 
@@ -374,10 +375,8 @@ class LymphMixtureModel:
                 ndim=self.n_cluster_parameters,
                 sampling_params=sampling_params,
                 starting_point=None,
-                hdf5_backend=hdf5_backend,
-                save_dir=self.samples_dir,
+                backend=hdf5_backend,
                 models=self,
-                verbose=True,
             )
 
         return sample_chain, end_point, log_probs

### 2024-01-24 scripts/lymixture/utils.py

@@ -1,9 +1,9 @@
+import logging
 from cycler import cycler
 import pandas as pd
 import lymph
 import numpy as np
 import emcee
-import multiprocess as mp
 import os
 from lyscripts.predict.prevalences import (
     compute_observed_prevalence,
@@ -11,9 +11,10 @@ from lyscripts.predict.prevalences import (
 import matplotlib.colors as mcolors
 from pathlib import Path
 from scipy.special import factorial
-import itertools, sys
+import itertools
 import scipy as sp
 import matplotlib.pyplot as plt
+from lyscripts.sample import run_mcmc_with_burnin, DummyPool
 
 # Plotting styles
 usz_blue = "#005ea8"
@@ -43,6 +44,7 @@ COLORS = {
 }
 
 PLOT_PATH = Path("./figures/")
+logger = logging.getLogger(__name__)
 
 
 # Define a function to set plot size based on desired width, unit, and ratio
@@ -311,54 +313,51 @@ def emcee_sampling_ext(
     logger.info(f"Dimension: {n_params} with n walkers: {nwalkers}")
     output_name = sample_name
 
-    if False:
-        samples = np.load("samples/" + output_name + ".npy")
-    else:
-        created_pool = mp.Pool(os.cpu_count())
-        with created_pool as pool:
-            if start_with is None:
-                starting_points = np.random.uniform(size=(nwalkers, n_params))
+    created_pool = DummyPool()
+    with created_pool as pool:
+        if start_with is None:
+            starting_points = np.random.uniform(size=(nwalkers, n_params))
+        else:
+            if np.shape(start_with) != np.shape(
+                np.random.uniform(size=(nwalkers, n_params))
+            ):
+                starting_points = np.tile(start_with, (nwalkers, 1))
             else:
-                if np.shape(start_with) != np.shape(
-                    np.random.uniform(size=(nwalkers, n_params))
-                ):
-                    starting_points = np.tile(start_with, (nwalkers, 1))
-                else:
-                    starting_points = start_with
-            logger.info(
-                f"Start Burning (steps = {burnin}) with {created_pool._processes} cores"
-            )
-            burnin_sampler = emcee.EnsembleSampler(
-                nwalkers,
-                n_params,
-                llh_function,
-                args=llh_args,
-                pool=pool,
-            )
-            burnin_results = burnin_sampler.run_mcmc(
-                initial_state=starting_points, nsteps=burnin, progress=True
-            )
[…]

### 2024-01-24 scripts/mixture_model_sampling.py

@@ -72,12 +72,6 @@ def main():
         split_by=SIMPLE_SUBSITE,
         mapping=lambda x: "all",
     )
-    mixture_model.cluster_assignments = np.zeros(shape=(
-        mixture_model.n_subpopulation * (mixture_model.n_clusters - 1)
-    )) + 0.2
-    mixture_model.cluster_parameters = np.random.uniform(
-        size=mixture_model.n_cluster_parameters
-    )
 
     _chain, cluster_components, history = mixture_model.fit(
         em_config=params["mixture_model"]["em_config"],

### 2024-01-24 scripts/helpers.py
add 1D plot of cluster assignments

@@ -1,6 +1,7 @@
 """
 Helper functions for the scripts.
 """
+OROPHARYNX_ICDS = ["C01", "C09", "C10"]
 
 def simplify_subsite(icd_code: str) -> str:
     """Only use the part of the ICD code before the decimal point."""

### 2024-01-24 scripts/plot_cluster_assignments.py

@@ -1,0 +1,74 @@
+"""
+Visualize the cluster assignments of the trained mixture model.
+"""
+import argparse
+from pathlib import Path
+import yaml
+
+import h5py
+import matplotlib.pyplot as plt
+from matplotlib.colors import to_rgba
+from tueplots import figsizes, fontsizes
+from lyscripts.plot.utils import COLORS as USZ
+
+from helpers import OROPHARYNX_ICDS
+
+
+def create_parser() -> argparse.ArgumentParser:
+    """Assemble the parser for the command line arguments."""
+    parser = argparse.ArgumentParser(description=__doc__)
+    parser.add_argument(
+        "-m", "--model", type=Path, default="models/mixture.hdf5",
+        help=(
+            "Path to the model HDF5 file. Needs to contain a dataset called "
+            "``em/cluster_assignments``."
+        )
+    )
+    parser.add_argument(
+        "-o", "--output", type=Path, default="figures/cluster_assignments.png",
+        help="Path to the output file.",
+    )
+    parser.add_argument(
+        "-p", "--params", type=Path, default="_variables.yml",
+        help="Path to the parameter file..",
+    )
+    return parser
+
+
+def main():
+    """Execute the main routine."""
+    args = create_parser().parse_args()
+    with open(args.params) as file:
+        params = yaml.safe_load(file)
+        num_patients = params["num_patients"]
+        num_patients.pop("total")
+
+    with h5py.File(args.model, mode="r") as h5_file:
+        cluster_components = h5_file["em/cluster_components"][...]
+
+    plt.rcParams.update(figsizes.icml2022_half())
+    plt.rcParams.update(fontsizes.icml2022())
+
+    _, bottom_ax = plt.subplots()
+    bottom_ax.scatter(
+        [cluster_components[i] for i, _ in enumerate(num_patients.keys())],
+        [0. for _ in num_patients.keys()],
+        s=[num for num in num_patients.values()],
+        c=[
+            to_rgba(USZ["orange"] if i in OROPHARYNX_ICDS else USZ["blue"], alpha=0.7)
+            for i in num_patients.keys()
+        ],
+        alpha=0.7,
+        linewidths=0.,
+    )
+    bottom_ax.set_xlabel("assignment to cluster A")
+    top_ax = bottom_ax.secondary_xaxis(
+        location="top",
+        functions=(lambda x: 1. - x, lambda x: 1. - x),
+    )
[…]

### 2024-01-24 scripts/prevalence_by_subsite.py

@@ -14,7 +14,7 @@ from tueplots import figsizes, fontsizes
 from lyscripts.utils import load_patient_data
 from lyscripts.plot.utils import COLORS as USZ
 
-from helpers import simplify_subsite
+from helpers import simplify_subsite, OROPHARYNX_ICDS
 
 
 SUBSITE = ("tumor", "1", "subsite")
@@ -23,7 +23,6 @@ LNL_I = ("max_llh", "ipsi", "I")
 LNL_II = ("max_llh", "ipsi", "II")
 LNL_III = ("max_llh", "ipsi", "III")
 LNL_IV = ("max_llh", "ipsi", "IV")
-OROPHARYNX_ICDS = ["C01", "C09", "C10"]
 
 
 def create_parser() -> argparse.ArgumentParser:

### 2024-01-24 scripts/plot_cluster_assignments.py
add cluster assignment plot and gh-pages publish

@@ -61,12 +61,13 @@ def main():
         alpha=0.7,
         linewidths=0.,
     )
-    bottom_ax.set_xlabel("assignment to cluster A")
+    bottom_ax.set_xlabel("assignment to cluster A [%]")
+    bottom_ax.set_xticklabels([f"{int(tick):.2%}" for tick in bottom_ax.get_xticks()])
     top_ax = bottom_ax.secondary_xaxis(
         location="top",
         functions=(lambda x: 1. - x, lambda x: 1. - x),
     )
-    top_ax.set_xlabel("assignment to cluster B")
+    top_ax.set_xlabel("assignment to cluster B [%]")
     plt.savefig(args.output, bbox_inches="tight", dpi=300)
 
 

### 2024-01-25 scripts/helpers.py
finalize cluster assignment plot

@@ -1,8 +1,35 @@
 """
 Helper functions for the scripts.
 """
+from typing import Generator
+
+from lyscripts.plot.utils import COLORS as USZ
+from matplotlib.colors import to_rgba
+
+
 OROPHARYNX_ICDS = ["C01", "C09", "C10"]
 
 def simplify_subsite(icd_code: str) -> str:
     """Only use the part of the ICD code before the decimal point."""
     return icd_code.split(".")[0]
+
+
+def generate_location_colors(
+    icd_codes: list[str],
+    delta_alpha: float = 0.15,
+) -> Generator[tuple[float, float, float, float], None, None]:
+    """Make a list of colors for each location."""
+    oropharynx_alpha = 1.0
+    oral_cavity_alpha = 1.0
+    colors, alphas = [], []
+    for icd_code in icd_codes:
+        if icd_code in OROPHARYNX_ICDS:
+            colors.append(USZ["orange"])
+            alphas.append(oropharynx_alpha)
+            oropharynx_alpha -= delta_alpha
+            yield to_rgba(USZ["orange"], oropharynx_alpha)
+        else:
+            colors.append(USZ["blue"])
+            alphas.append(oral_cavity_alpha)
+            oral_cavity_alpha -= delta_alpha
+            yield to_rgba(USZ["blue"], oral_cavity_alpha)

### 2024-01-25 scripts/plot_cluster_assignments.py

@@ -3,15 +3,16 @@ Visualize the cluster assignments of the trained mixture model.
 """
 import argparse
 from pathlib import Path
+from matplotlib.ticker import StrMethodFormatter
+import numpy as np
 import yaml
 
 import h5py
 import matplotlib.pyplot as plt
-from matplotlib.colors import to_rgba
 from tueplots import figsizes, fontsizes
 from lyscripts.plot.utils import COLORS as USZ
 
-from helpers import OROPHARYNX_ICDS
+from helpers import generate_location_colors
 
 
 def create_parser() -> argparse.ArgumentParser:
@@ -50,24 +51,48 @@ def main():
     plt.rcParams.update(fontsizes.icml2022())
 
     _, bottom_ax = plt.subplots()
+    cluster_x = [cluster_components[i] for i, _ in enumerate(num_patients.keys())]
+    cluster_y = [0. for _ in num_patients.keys()]
+    annotations = [f"{label}\n({num})" for label, num in num_patients.items()]
     bottom_ax.scatter(
-        [cluster_components[i] for i, _ in enumerate(num_patients.keys())],
-        [0. for _ in num_patients.keys()],
+        cluster_x, cluster_y,
         s=[num for num in num_patients.values()],
-        c=[
-            to_rgba(USZ["orange"] if i in OROPHARYNX_ICDS else USZ["blue"], alpha=0.7)
-            for i in num_patients.keys()
-        ],
+        c=list(generate_location_colors(num_patients.keys())),
         alpha=0.7,
         linewidths=0.,
+        zorder=10,
     )
-    bottom_ax.set_xlabel("assignment to cluster A [%]")
-    bottom_ax.set_xticklabels([f"{int(tick):.2%}" for tick in bottom_ax.get_xticks()])
+
+    sorted_idx = cluster_components.argsort()
+    sorted_x = cluster_components[sorted_idx]
+    sorted_annotations = [annotations[i] for i in sorted_idx]
+    sorted_num = [list(num_patients.values())[i] for i in sorted_idx]
+    for i, (x, num, annotation) in enumerate(zip(sorted_x, sorted_num, sorted_annotations)):
+        bottom_ax.annotate(
+            annotation,
+            # sqrt, because marker's area grows linearly with patient num, not radius
+            xy=(x, np.sqrt(0.0000003 * num) * (- 1)**i),
+            xytext=(x, 0.025 * (- 1)**i),
[…]

### 2024-01-25 scripts/prevalence_by_subsite.py

@@ -5,16 +5,13 @@ location (different colors).
 """
 import argparse
 from pathlib import Path
-from typing import Generator
 import yaml
 
 import matplotlib.pyplot as plt
-from matplotlib.colors import to_rgba
 from tueplots import figsizes, fontsizes
 from lyscripts.utils import load_patient_data
-from lyscripts.plot.utils import COLORS as USZ
 
-from helpers import simplify_subsite, OROPHARYNX_ICDS
+from helpers import simplify_subsite, generate_location_colors
 
 
 SUBSITE = ("tumor", "1", "subsite")
@@ -41,27 +38,6 @@ def create_parser() -> argparse.ArgumentParser:
         help="Path to the parameter file. Looks for a key `barplot_kwargs`.",
     )
     return parser
-
-
-def generate_location_colors(
-    icd_codes: list[str],
-    delta_alpha: float = 0.15,
-) -> Generator[tuple[float, float, float, float], None, None]:
-    """Make a list of colors for each location."""
-    oropharynx_alpha = 1.0
-    oral_cavity_alpha = 1.0
-    colors, alphas = [], []
-    for icd_code in icd_codes:
-        if icd_code in OROPHARYNX_ICDS:
-            colors.append(USZ["orange"])
-            alphas.append(oropharynx_alpha)
-            oropharynx_alpha -= delta_alpha
-            yield to_rgba(USZ["orange"], oropharynx_alpha)
-        else:
-            colors.append(USZ["blue"])
-            alphas.append(oral_cavity_alpha)
-            oral_cavity_alpha -= delta_alpha
-            yield to_rgba(USZ["blue"], oral_cavity_alpha)
 
 
 def main():

### 2024-01-25 content/manuscript.qmd
add cluster assignment plot and gh-pages publish
add schematic & fix subsite labels in plots

@@
 
-We model the process of tumor progression via lymphatic drainage by connecting the RVs in a graph, as shown in @fig-graph-with-obs.
-The arcs in [this graph](#fig-graph-with-obs) represent conditional probabilities.
+We model the process of tumor progression via lymphatic drainage by connecting the RVs in a graph, as shown in @fig-schematic-and-graph.
+The arcs in [this graph](#fig-schematic-and-graph) represent conditional probabilities.
 The orange arcs correspond to observing a diagnosis $Y_v$, given the true state $X_v$.
@@
 
-The red arcs in the graph of @fig-graph-with-obs depict the probability that the primary tumor spreads within one abstract time-step.
+The red arcs in the graph of @fig-schematic-and-graph depict the probability that the primary tumor spreads within one abstract time-step.
 While the blue arcs symbolize the spread from an upstream LNL -- given it is already metastatic -- to a downstream level.
@@
 : Conditional probability $P\left( X_3[t+1] \mid X_2[t] \right)$ for a spread from LNL II to III during the transition from $t$ to $t+1$.
-This corresponds to one of the blue arcs in @fig-graph-with-obs.
+This corresponds to one of the blue arcs in @fig-schematic-and-graph.
 Note that the values in the row with $X_3[t+1]=0$ is all zeros, and the row with $X_3[t+1]=1$ all ones if $X_3[t]=1$.
@@
 Here, when we use $\boldsymbol{\xi}$ instead of $\mathbf{x}$ for the values the random vector $\mathbf{X}$ can take on, the $i$ and $j$ enumerate all $2^V$ combinations of the $V$ binary RVs.
-In the graph shown in @fig-graph-with-obs, this amounts to $2^V=8$ distinct $\boldsymbol{\xi}_i$.
+In the graph shown in @fig-schematic-and-graph, this amounts to $2^V=8$ distinct $\boldsymbol{\xi}_i$.
 Because these terms are essentially products of terms like those in @tbl-spread-2-to-3.
@@
 
-![Parametrized graph representation of the lymphatic network considering four LNLs.
-Blue, round nodes represent the hidden RVs, while orange square nodes show the observed RVs.
-Arcs represent a conditional probability parametrized with the quantity noted next to it.](../static/graph_with_obs.svg){#fig-graph-with-obs width="25%" fig-pos="th"}
+![On the left: Rough anatomical sketch of the tumor subsites and correspsonding ICD-10 codes that are present in the used data.
+The subsite "other parts of mouth" (C06) was not drawn.
+On the right: Parametrized graph representation of the lymphatic network considering four LNLs.
+Blue nodes represent the hidden RVs, while the red one is the tumor.
+Arcs represent a conditional probability parametrized with the quantity noted next to it](../static/schematic_and_graph.svg){#fig-schematic-and-graph fig-pos="th"}
 
@@
 Here, the $\mathbf{c} \in [0,1]^g$ is the vector of mixing proportions with $\sum_{j=0}^g c_j = 1$, while the $\boldsymbol{\Psi} = \left( \boldsymbol{\theta}_1, \ldots, \boldsymbol{\theta}_g\right)$ is the vector of all $g$ models' parameters.
-Note that we will implement our model such that some of the parameters in each $\boldsymbol{\theta}_j$ are shared across the $g$ components -- namely the $t_{r \rightarrow v}$ corresponding to the blue arcs in @fig-graph-with-obs.
 
@@
 The subsites are sorted in ascending order by their prevalence of involvement in LNL II.
-Oral cavity subsites are plotted in shaed of blue, oropharynx subsites in shades of orange.](../figures/prevalence_by_subsite.svg){#fig-prevalence-by-subsite fig-pos="b" width="49%"}
+Oral cavity subsites are plotted in shaed of blue, oropharynx subsites in shades of orange.](../figures/prevalence_by_subsite.svg){#fig-prevalence-by-subsite fig-pos="b"}
 
@@
 
+![Assignment of each subsite to each of the two clusters.
+The fruther left a subsite, the more it is assigned to cluster A, the further right, the more to cluster B.
+The size of the marker (area) corresponds to the number of patients in the subsite.](../figures/cluster_assignments.svg){#fig-cluster-assignments}
+
 # Discussion {#sec-discussion}

### 2024-01-25 scripts/prevalence_by_subsite.py

@@ -34,7 +34,7 @@ def create_parser() -> argparse.ArgumentParser:
         help="Path to the output file.",
     )
     parser.add_argument(
-        "-p", "--params", type=Path, default="params.yaml",
+        "-p", "--params", type=Path, default="_variables.yml",
         help="Path to the parameter file. Looks for a key `barplot_kwargs`.",
     )
     return parser
@@ -49,6 +49,7 @@ def main():
             params = yaml.safe_load(f)
         barplot_kwargs.update(params.get("barplot_kwargs", {}))
         icd_code_map = params.get("icd_code_map", {})
+        icd_code_map = {k: f"{v} ({k})" for k, v in icd_code_map.items()}
 
     patient_data = load_patient_data(args.input)
     patient_data[SUBSITE] = patient_data[SUBSITE].apply(simplify_subsite)

### 2024-01-25 content/manuscript.qmd
work in Jan's edits

@@
 Head and neck squamous cell carcinomas (HNSCC) frequently spread through the lymphatic system [@lindberg_distribution_1972;@woolgar_histological_1999].
-Current diagnostic imaging modalities are unable to detect microscopic nodal metastases, which requires pathological examination of extracted tissue [@snyder_petct_2021;@strohl_petct_2021].
-Since the recurrence of nodal disease is detrimental to a patient's outcome [@ho_decision_2014], large volumes in the head and neck region are often irradiated electively to minimize the risk of missing occult disease.
-Decision guidelines about which nodal regions -- i.e., anatomically defined lymph node levels (LNLs) -- to irradiate [@biau_selection_2019] are currently not based on a patient's individual risk, but only on the overall prevalence of nodal disease as reported in the literature [@lindberg_distribution_1972;@woolgar_histological_1999].
+Current diagnostic imaging modalities are unable to detect microscopic lymph node metastases [@snyder_petct_2021;@strohl_petct_2021].
+To avoid nodal recurrences [@ho_decision_2014], large volumes in the neck are irradiated electively, which are at risk of harbouring occult disease.
+Guidelines about which lymph node levels (LNLs) to irradiate [@biau_selection_2019] are currently not based on a patient's individual risk, but only on the overall prevalence of nodal disease as reported in the literature [@lindberg_distribution_1972;@woolgar_histological_1999].
 
-To personalize this prediction of the risk for occult diease, given a patient's individual diagnosis, we published
+To personalize this prediction of the risk for occult disease, given a patient's individual diagnosis, we published
 
@@
 
-Such a personalized risk prediction may allow clinitians to safely reduce the elective clinical target volume (CTV-N) and thus reduce side-effects that degrade the patient's quality of life [@batth_practical_2014].
+Such a personalized risk prediction may allow clinicians to safely reduce the elective clinical target volume (CTV-N) and thus reduce side-effects that degrade the patient's quality of life [@batth_practical_2014].
 
-HNSCC patients with primary tumors at different subsites, e.g.
-in the oral cavity and in the oropharynx, also show different patterns of lymphatic spread [@lindberg_distribution_1972;@woolgar_histological_1999].
-Our model does so far not have the capability to naturally describe different tumor subsites.
-To that end, we present an approach using mixtures of these HMMs in this work.
-This makes intuitive sense, because if a tumor lies anatomically between two anatomically close subsites with slightly different spread patterns, we may be able to describe its lymphatic progression as a mixture of these two spread patterns.
+Here, we extend the previous work by incorporating the primary tumor location (specified as ICD-10 code) into the model of lymphatic tumour progression, focusing on tumours in the oropharynx and the oral cavity.
+HNSCC patients with primary tumors at different subsites show different patterns of lymphatic spread [@lindberg_distribution_1972;@woolgar_histological_1999].
+So far, this could be handled by training different models for broader categories of tumour locations, e.g.
+oropharynx and oral cavity tumours.
+However, this approach does not describe differences in lymphatic spread between different subsites within the oropharynx and oral cavity.
+To address this issue, we present an approach using mixtures of HMMs.
+The intuition is that the lymphatic spread of a tumor that lies anatomically at the boarder of oropharynx and oral cavity (e.g.
+tumours in the palate) may be described by a mixture of different models.
 
@@
 
-To compute the personalized risk of occult disease $\mathbf{X}$, given a diagnosis $\mathbf{Y}$, we can begin by stating Bayes' law:
+Each LNL $v \in \{ 1, 2, \ldots, V\}$ considered in our model is represented by a binary random variable (RV) $X_v$ representing the true state of that level (0 for "healthy" and 1 for "involved").
+A patient's state of lymph node involvement can be represented in a random vector $\mathbf{X} = \left( X_1, X_2, \ldots, X_V \right)$.
+When a patient is diagnosed with HNSCC, we only observe the clinical lymph node involvement based on imaging, which we denote as another binary random variable $Y_v$.
+To compute the personalized risk of occult disease $\mathbf{X}$, given a diagnosis $\mathbf{Y}$, we apply Bayes' law:
 
@@
 
-In the above equation, the term $P \left( \mathbf{Y} \mid \mathbf{X} \right)$ is given by the sensitivity and specificity of the diagnosis.
-The crucial term that our model attempts to compute, is the prior probability of involvement $P \left( \mathbf{X} \right)$ (or rather of the hidden state of involvement, see the [next section](#sec-hmm)).
+In the above equation, the term $P \left( \mathbf{Y} \mid \mathbf{X} \right)$ is given by the sensitivity and specificity of the diagnostic procedure.
+The term $P \left( \mathbf{X} \right)$ represents the prior probability of involvement, which depends on the probability of the tumour to spread through the lymphatic system.
+The main task of the HMM is to model $P \left( \mathbf{X} \right)$ and the main contribution of this paper is to incorporate the primary tutor subset into the model of $P \left( \mathbf{X} \right)$.
 
@@
 
-Each LNL $v \in \{ 1, 2, \ldots, V\}$ considered in our model is represented by a binary random variable (RV) $X_v[t]$ taking on the true state of that level at the abstract time-step $t$ (0 for "healthy" and 1 for "involved").
-Collected in a random vector $\mathbf{X}[t] = \left( X_1[t], X_2[t], \ldots, X_V[t] \right)$ they form the patient's state w.r.t.
-their lymphatic involvement at time $t$.
-
-We model the process of tumor progression via lymphatic drainage by connecting the RVs in a graph, as shown in @fig-schematic-and-graph.
-The arcs in [this graph](#fig-schematic-and-graph) represent conditional probabilities.
-The orange arcs correspond to observing a diagnosis $Y_v$, given the true state $X_v$.
-For the sake of brevity, we will not go into the details of describing how to infer the true -- but technically hidden -- state of LNL involvement from diagnoses with lower-than-one sensitivity and specificity.
-This description can be found in @zora231470.
-Throughout this work, we instead combine diagnostic and pathologic involvement information from the data into a "maximum likelihood" diagnosis and assume its sensitivity and specificity to be one, meaning the normally hidden state $X_v$ becomes the observed state.
-This simplification is reasonable for pathologic involvement and because we are at this stage more interested in testing whether our model is able to describe a realistic probability distribution over lymphatic involvement.
-
-The red arcs in the graph of @fig-schematic-and-graph depict the probability that the primary tumor spreads within one abstract time-step.
-While the blue arcs symbolize the spread from an upstream LNL -- given it is already metastatic -- to a downstream level.
-For example, the edge from $X_2$ to $X_3$ encodes the probability $P\left( X_3[t+1] \mid X_2[t] \right)$, which is parametrized with $b_3$, and $t_{2 \rightarrow 3}$ and tabulated in @tbl-spread-2-to-3.
-There is an additional restriction on any LNL's state $X_v[t+1]=0$ to be healthy: It requires that the level was also healthy in the time-step before, meaning $X_v[t]=0$.
-This is because we assume no spontaneous self-healing of metastatic levels.
-
-+--------------+------------+-----------------------------------------------------+
-|              | $X_2[t]=0$ | $X_2[t]=1$                                          |
-+==============+:==========:+:===================================================:+
-| $X_3[t+1]=0$ | $1-b_3$    | $(1-b_3)(1-t_{2 \rightarrow 3})$                    |
-+--------------+------------+-----------------------------------------------------+
-| $X_3[t+1]=1$ | $b_3$      | $1-b_3-t_{2 \rightarrow 3}+b_3 t_{2 \rightarrow 3}$ |
-+--------------+------------+-----------------------------------------------------+
-
-: Conditional probability $P\left( X_3[t+1] \mid X_2[t] \right)$ for a spread from LNL II to III during the transition from $t$ to $t+1$.
-This corresponds to one of the blue arcs in @fig-schematic-and-graph.
-Note that the values in the row with $X_3[t+1]=0$ is all zeros, and the row with $X_3[t+1]=1$ all ones if $X_3[t]=1$.
-{#tbl-spread-2-to-3 tbl-colwidths="[25,18,40]"}
-
-With the introduced conditional probabilities, we can now compute the joint probability of any complete state $\mathbf{X}[t]=\boldsymbol{\xi}_i$ transitioning to any other possible state $\mathbf{X}[t+1]=\boldsymbol{\xi}_j$.
-Here, when we use $\boldsymbol{\xi}$ instead of $\mathbf{x}$ for the values the random vector $\mathbf{X}$ can take on, the $i$ and $j$ enumerate all $2^V$ combinations of the $V$ binary RVs.
-In the graph shown in @fig-schematic-and-graph, this amounts to $2^V=8$ distinct $\boldsymbol{\xi}_i$.
-Because these terms are essentially products of terms like those in @tbl-spread-2-to-3.
-We can then collect these terms in a _transition matrix_ $\mathbf{A}$:
+A patient's state of lymph node involvement $\mathbf{X}[t]$ evolves over discrete time steps $t$.
+Let us enumerate all $2^V$ possible states, representing all combinations of LNLs.
+In this paper, we consider ipsilateral LNLs I, II, III and IV, which amounts to 16 possible states.
+The HMM is specified by _transition matrix_ $\mathbf{A}$:
 
@@
 
-Note that this matrix still depends on the $b_v$ and $t_{r \rightarrow v}$ parameters, although we have dropped the explicit dependcy to keep the equations brief.
-Now, assuming that every patient started their disease with all LNLs being healthy, we can define the _starting distribution_ $\boldsymbol{\pi}$:
+which contains the conditional probabilities that a state $\mathbf{X}[t]=\boldsymbol{\xi}_i$ transitions to $\mathbf{X}[t+1]=\boldsymbol{\xi}_j$ over one time step.
+The transition matrix is specified and parameterised via the graphical model shown in @fig-schematic-and-graph.
+The red arcs in the graph of @fig-schematic-and-graph are associated the probability that the primary tumor spreads directly to a LNL (parameters $b_v$).
+The blue arcs describe the spread from an upstream LNL -- given it is already metastatic -- to a downstream level (parameters $t_{v-1 \rightarrow v}$).
+
+Now, let $\boldsymbol{\pi}$ be the _starting distribution_
 
@@
 
-And set every entry of this starting distribution to zero, except the first one, which we set to one.
-This means at $t=0$ there is a probability of one to be in the completly healthy state $\boldsymbol{\xi}_0 = \begin{pmatrix} 0, 0, \ldots, 0 \end{pmatrix}$.
+denoting the probability to start in state $\boldsymbol{\xi}_i$ at time step 0.
+Assuming that every patient started their disease with all LNLs being healthy, we set $\pi_i$ to zero for all states  except the completly healthy state $\boldsymbol{\xi} = \begin{pmatrix} 0, 0, 0, 0 \end{pmatrix}$, which has probability 1.
 
-Using the quantities introduced so far, the probability distribution vector with elements $P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right)$ after $t$ time-steps can now be conveniently expressed as a matrix product:
+Using the quantities introduced so far, the probability $P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right)$ to be in state $i$ in time step $t$ can now be conveniently expressed as a matrix product:
 
@@
 This evolution implicitly marginalizes over all possible paths to arrive at state $\boldsymbol{\xi}_i$ after $t$ time-steps.
-Additionally, we also need to marginalize over the time of diagnosis -- which is unknown -- using a time-prior $P_T(t)$.
-Fortunately, the exact length and shape of this distribution on its own has little impact.
-But because we assume that early and advanced T-category patients are fundamentally the same, just on average diagnosed at different times $t$, we use the time-prior only to separate the respective patient's evolutions:
+Additionally, we must marginalize over the unknown time of diagnosis using a time-prior $P(t)$.
+This finally defines the probability distribution over all states of lymph node involvement used in @eq-bayes-law.
 
 $$
-P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid T, \boldsymbol{\theta} \right) = \sum_{t=0}^{t_\text{max}} P_T(t) \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
+P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta} \right) = \sum_{t=0}^{t_\text{max}} P_T(t) \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
 $$ {#eq-marginalized-evolution}
 
-where $T \in \{ \text{early}, \text{advanced} \}$ denotes the T-category.
-We fix $P_\text{early}(t)$ to a binomial distribution with parameters $n=t_\text{max}=$ {{< var model.max_t >}} and $p_\text{early}=$ {{< var model.first_binom_prob >}}, while the advanced T-category's time-prior is also a binomial distribution where the $p_\text{advanced}$ parameter is learned.
-
-Note that @eq-marginalized-evolution still depends on the parametrization of the transition matrix.
-We collect these parameters in a vector $\boldsymbol{\theta}=\{ b_v, t_{r \rightarrow v}, p_\text{advanced} \}$.
-
-To train our model, we need to compute the likelihood of a dataset $\mathbf{D}=\left( \mathbf{x}_1, \mathbf{x}_2, \ldots, \mathbf{x}_N \right)$ of $N$ patients, given the model's parameters $\boldsymbol{\theta}$.
-This is given by a product of the terms in @eq-marginalized-evolution:
-
-$$
-P \left(\mathbf{D} \mid \boldsymbol{\theta} \right) = \prod_{i=1}^N P \left( \mathbf{X} = \mathbf{x}_i \mid T_i, \boldsymbol{\theta} \right)
-$$ {#eq-likelihood}
-
-Typically, one would compute the logarithm of this quantity for computational reasons.
-There is a wide array of available methods to obtain maximum likelihood estimates from this function or sample from the posterior over $\boldsymbol{\theta}$.
-In this work, we use Markov chain Monte Carlo sampling (MCMC) via the `emcee` Python package [@foreman-mackey_emcee_2013].
+where $\boldsymbol{\theta}=\{ b_v, t_{r \rightarrow v} \}$ denotes the set of all model parameters (7 in our case).
+Fortunately, the exact length and shape of this distribution on its own has little impact as previously shown.
+We set $t_\text{max}=$ {{< var model.max_t >}} and $P_\text{early}(t)$ to a binomial distribution with parameter {{< var model.first_binom_prob >}}.
+Further details on the HMM can be found in ...
 
@@
 ## Mixture of HMMs {#sec-mixture-of-hmms}
+
+Let us now assume that primary tumors at different subsites have different patterns of lymphatic spread, corresponding to different model parameters $\boldsymbol{\theta}$.
+Training a separate model for every possible subsite (ICD-10 code) would require a sufficiently large dataset for every tumor site.
+However, anatomically nearby locations are expected to show very similar patterns of LNL involvement.
+Therefore, we consider a mixture model.
+
+Let us assume that we have a dataset $\mathbf{D}$ that is specified via the number of patients $N_{is}$ that were diagnosed in LNL involvement state $i$ and had a primary tutor in subsite $s$.
+Let us further assume that we want to describe this dataset using a mixture of $M$ HMMs, each with a different set of model parameters $\boldsymbol{\theta_m}$.
+As the generative model of the data, we assume that a patient with subsite $s$ is generated with probability $\pi_{sm}$ from model $m$.
+The likelihood of the dataset can then be written as
+
+$$
+P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_s \prod_i \left[ \sum_{m=1}^M \pi_{sm} P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m \right) \right]^{N_{is}}
+$$ {#eq-mixture-distribution}
+
+We now have two types of parameters, the probabilities of tutor spread for the different models $\boldsymbol{\theta_m}$, and the mixing coefficients $\pi_{sm}$.
+Assuming a uniform prior in the interval $[0,1]$ for all parameters, the posterior distribution over the parameters $P \left( \boldsymbol{\theta}, \boldsymbol{\pi} \mid \mathbf{D} \right)$ is given by the likelihood @eq-mixture-distribution except for a normalisation constant.
+In this work, we use Markov chain Monte Carlo sampling (MCMC) via the `emcee` Python package [@foreman-mackey_emcee_2013] to sample model parameters from the posterior distribution.
+However, $P \left( \boldsymbol{\theta}, \boldsymbol{\pi} \mid \mathbf{D} \right)$ itself is a multi-model distribution because one can permute the different models.
+To address this problem, we revert to an _expectation-maximization (EM)_ algorithm where we iteratively sample model parameters $\boldsymbol{\theta_m}$ using MCMC and then determine the most likely mixing coefficients.
+
+...
 
@@
 Here, the $\mathbf{c} \in [0,1]^g$ is the vector of mixing proportions with $\sum_{j=0}^g c_j = 1$, while the $\boldsymbol{\Psi} = \left( \boldsymbol{\theta}_1, \ldots, \boldsymbol{\theta}_g\right)$ is the vector of all $g$ models' parameters.
+Note that we will implement our model such that some of the parameters in each $\boldsymbol{\theta}_j$ are shared across the $g$ components -- namely the $t_{r \rightarrow v}$ corresponding to the blue arcs in @fig-schematic-and-graph.
 

### 2024-01-25 scripts/helpers.py
start with prevalence plots

@@ -2,6 +2,7 @@
 Helper functions for the scripts.
 """
 from typing import Generator
+import argparse
 
 from lyscripts.plot.utils import COLORS as USZ
 from matplotlib.colors import to_rgba
@@ -33,3 +34,17 @@ def generate_location_colors(
             alphas.append(oral_cavity_alpha)
             oral_cavity_alpha -= delta_alpha
             yield to_rgba(USZ["blue"], oral_cavity_alpha)
+
+
+def str2bool(v: str) -> bool | None:
+    """Transform a string to a boolean or ``None``."""
+    if isinstance(v, bool):
+        return v
+    if v.lower() in ('yes', 'true', 't', 'y', '1'):
+        return True
+    elif v.lower() in ('no', 'false', 'f', 'n', '0'):
+        return False
+    elif v.lower() in ('none', 'null', ''):
+        return None
+    else:
+        raise argparse.ArgumentTypeError('Boolean or None value expected.')

### 2024-01-25 scripts/mixture_model_sampling.py

@@ -59,7 +59,6 @@ def main():
     num_subsites = len(patient_data[SIMPLE_SUBSITE].unique())
     lymph_model = create_model_from_config(params)
     lymph_model.modalities = {"max_llh": [1., 1.]}
-    lymph_model.diag_time_dists = {"all": lymph_model.diag_time_dists["early"]}
 
     mixture_model = LymphMixtureModel(
         lymph_model=lymph_model,

### 2024-01-25 scripts/prevalence_comparison.py

@@ -1,0 +1,127 @@
+"""
+Plot the prevalence for involvement, as predicted by the trained mixture model, and
+compare it to the prevalence of involvement in the data.
+"""
+import argparse
+from pathlib import Path
+import yaml
+
+import numpy as np
+import matplotlib.pyplot as plt
+import h5py
+from emcee.backends import HDFBackend
+from lyscripts.utils import load_patient_data, create_model_from_config
+from lyscripts.predict.prevalences import (
+    compute_observed_prevalence,
+    generate_predicted_prevalences,
+)
+
+from helpers import str2bool, OROPHARYNX_ICDS
+
+
+T_STAGE = ("tumor", "1", "t_stage")
+LNLS = ["I", "II", "III", "IV"]
+
+
+def create_parser() -> argparse.ArgumentParser:
+    """Assemble the parser for the command line arguments."""
+    parser = argparse.ArgumentParser(description=__doc__)
+    parser.add_argument(
+        "-d", "--data", type=Path, default="data/enhanced.csv",
+        help="Path to the data file.",
+    )
+    parser.add_argument(
+        "--mixture-model", type=Path, default="models/mixture.hdf5",
+        help=(
+            "Path to the mixture model HDF5 file. Needs to contain a dataset called "
+            "``em/cluster_components``. Parent directory is assumed to also contain "
+            "the HDF5 file of the independent model (either `oropharynx.hdf5` or "
+            "`oral_cavity.hdf5`)."
+        )
+    )
+    parser.add_argument(
+        "-o", "--output", type=Path, default="figures/prevalence_comparison.png",
+        help="Path to the output file.",
+    )
+    parser.add_argument(
+        "-p", "--params", type=Path, default="_variables.yml",
+        help="Path to the parameter file..",
+    )
+    parser.add_argument(
+        "-i", "--involvement", nargs=4, type=str2bool, required=True,
+        help="Involvement pattern to compare the prevalences for.",
+    )
+    parser.add_argument(
+        "-s", "--subsite", type=str, default="C01",
+        help="ICD code of the subsite to plot the prevalences for.",
+    )
+    parser.add_argument(
+        "--thin-indie", type=int, default=10,
+        help="Thinning factor for the independent model's chain.",
+    )
+    parser.add_argument(
+        "--thin-mixture", type=int, default=1000,
[…]

### 2024-01-25 scripts/helpers.py
rerun indie sampling with custom T-stage mapping

@@ -9,6 +9,16 @@ from matplotlib.colors import to_rgba
 
 
 OROPHARYNX_ICDS = ["C01", "C09", "C10"]
+SUBSITE = ("tumor", "1", "subsite")
+SIMPLE_SUBSITE = ("tumor", "1", "simple_subsite")
+LOCATION = ("tumor", "1", "location")
+T_STAGE = ("tumor", "1", "t_stage")
+LNLS = ["I", "II", "III", "IV"]
+LNL_I = ("max_llh", "ipsi", "I")
+LNL_II = ("max_llh", "ipsi", "II")
+LNL_III = ("max_llh", "ipsi", "III")
+LNL_IV = ("max_llh", "ipsi", "IV")
+
 
 def simplify_subsite(icd_code: str) -> str:
     """Only use the part of the ICD code before the decimal point."""

### 2024-01-25 scripts/mixture_model_sampling.py

@@ -14,14 +14,11 @@ from lyscripts.utils import (
     create_model_from_config,
 )
 
-from helpers import simplify_subsite
+from helpers import simplify_subsite, SUBSITE, SIMPLE_SUBSITE
 from lymixture.mixture_model import LymphMixtureModel
 
 
 SUBSITE = ("tumor", "1", "subsite")
-SIMPLE_SUBSITE = ("tumor", "1", "simple_subsite")
-
-
 def create_parser() -> argparse.ArgumentParser:
     """Assemble the parser for the command line arguments."""
     parser = argparse.ArgumentParser(description=__doc__)

### 2024-01-25 scripts/prevalence_by_subsite.py

@@ -11,15 +11,15 @@ import matplotlib.pyplot as plt
 from tueplots import figsizes, fontsizes
 from lyscripts.utils import load_patient_data
 
-from helpers import simplify_subsite, generate_location_colors
-
-
-SUBSITE = ("tumor", "1", "subsite")
-LOCATION = ("tumor", "1", "location")
-LNL_I = ("max_llh", "ipsi", "I")
-LNL_II = ("max_llh", "ipsi", "II")
-LNL_III = ("max_llh", "ipsi", "III")
-LNL_IV = ("max_llh", "ipsi", "IV")
+from helpers import (
+    simplify_subsite,
+    generate_location_colors,
+    SUBSITE,
+    LNL_I,
+    LNL_II,
+    LNL_III,
+    LNL_IV,
+)
 
 
 def create_parser() -> argparse.ArgumentParser:

### 2024-01-25 scripts/prevalence_comparison.py

@@ -16,11 +16,8 @@ from lyscripts.predict.prevalences import (
     generate_predicted_prevalences,
 )
 
-from helpers import str2bool, OROPHARYNX_ICDS
+from helpers import str2bool, OROPHARYNX_ICDS, SUBSITE, T_STAGE, LNLS
 
-
-T_STAGE = ("tumor", "1", "t_stage")
-LNLS = ["I", "II", "III", "IV"]
 
 
 def create_parser() -> argparse.ArgumentParser:
@@ -80,11 +77,13 @@ def main():
     patient_data[T_STAGE] = ["all"] * len(patient_data)
 
     if args.subsite in OROPHARYNX_ICDS:
+        print("using oropharynx model")
         indie_chain = HDFBackend(
             args.mixture_model.parent / "oropharynx.hdf5",
             read_only=True,
         ).get_chain(flat=True, thin=args.thin_indie)
     else:
+        print("using oral cavity model")
         indie_chain = HDFBackend(
             args.mixture_model.parent / "oral_cavity.hdf5",
             read_only=True,
@@ -95,10 +94,13 @@ def main():
     ).get_chain(flat=True, thin=args.thin_mixture)
     with h5py.File(args.mixture_model, mode="r") as h5_file:
         cluster_assignments = h5_file["em/cluster_components"][...]
+    idx = list(params["icd_code_map"].keys()).index(args.subsite)
+    cluster_A_prob = cluster_assignments[idx]
+    cluster_B_prob = 1 - cluster_A_prob
 
     matching, total = compute_observed_prevalence(
         pattern=pattern,
-        data=patient_data,
+        data=patient_data.loc[patient_data[SUBSITE].str.contains(args.subsite)],
         lnls=LNLS,
         t_stage="all",
         modality="max_llh",
@@ -114,14 +116,25 @@ def main():
     cluster_A_prevs = np.array(list(generate_predicted_prevalences(
         pattern=pattern,
         model=lymph_model,
-        samples=mixture_chain,
+        samples=mixture_chain[:,:7],
         t_stage="all",
         modality_spsn=[1., 1.],
     )))
+    cluster_B_prevs = np.array(list(generate_predicted_prevalences(
+        pattern=pattern,
+        model=lymph_model,
+        samples=mixture_chain[:,7:],
+        t_stage="all",
+        modality_spsn=[1., 1.],
+    )))
+    mixture_model_prevs = (
[…]

### 2024-01-25 scripts/write_num_patients.py

@@ -7,8 +7,7 @@ import yaml
 
 from lyscripts.utils import load_patient_data
 
-
-SUBSITE = ("tumor", "1", "subsite")
+from helpers import SUBSITE
 
 
 def create_parser() -> argparse.ArgumentParser:

### 2024-01-26 content/manuscript.qmd
run, produce prev plots & track SVGs

@@
   parameters in an iterative, EM-like algorithm from multi-centric data. The mixture
-  model manages to cluster anatomically close subsites and correctly infers the
+  model manages to group anatomically close subsites and correctly infers the
   clusters' parameters. Using this mixture model allows the prediction of individual
@@
 
-We did this in this and that fashion...
+In our implementation of the EM alorithm, we alternated between the following two steps:
+
+1.
+Given a set of $\boldsymbol{\pi}^\star$, we sampled new cluster parameters $\boldsymbol{\theta}^\star$ from the .
+we ran 7 times {{< var mixture_model.em_config.e_step.walkers_per_dim >}} parallel `emcee` walkers for  {{< var mixture_model.em_config.e_step.nburnin >}} steps and discarded the results as burn-in.
+Then, we completed another {{< var mixture_model.em_config.e_step.nsteps >}} steps and used
 

### 2024-01-26 scripts/helpers.py

@@ -50,11 +50,11 @@ def str2bool(v: str) -> bool | None:
     """Transform a string to a boolean or ``None``."""
     if isinstance(v, bool):
         return v
-    if v.lower() in ('yes', 'true', 't', 'y', '1'):
+    if v.lower() in ('yes', 'true', 't'):
         return True
-    elif v.lower() in ('no', 'false', 'f', 'n', '0'):
+    elif v.lower() in ('no', 'false', 'f'):
         return False
-    elif v.lower() in ('none', 'null', ''):
+    elif v.lower() in ('none', 'null', 'n'):
         return None
     else:
         raise argparse.ArgumentTypeError('Boolean or None value expected.')

### 2024-01-26 scripts/mixture_model_sampling.py

@@ -38,6 +38,10 @@ def create_parser() -> argparse.ArgumentParser:
         "-p", "--params", type=Path, default="_variables.yml",
         help="Path to the parameter file. Looks for a key `barplot_kwargs`.",
     )
+    parser.add_argument(
+        "-s", "--seed", type=int, default=42,
+        help="Seed for the random number generator.",
+    )
     return parser
 
 
@@ -69,6 +73,7 @@ def main():
         mapping=lambda x: "all",
     )
 
+    np.random.seed(args.seed)
     _chain, cluster_components, history = mixture_model.fit(
         em_config=params["mixture_model"]["em_config"],
         mcmc_config=params["mixture_model"]["mcmc_config"],

### 2024-01-26 scripts/plot_component_assignments.py

@@ -1,5 +1,5 @@
 """
-Visualize the cluster assignments of the trained mixture model.
+Visualize the component assignments of the trained mixture model.
 """
 import argparse
 from pathlib import Path
@@ -83,13 +83,13 @@ def main():
             }
         )
 
-    bottom_ax.set_xlabel("assignment to cluster B")
+    bottom_ax.set_xlabel("assignment to component B")
     bottom_ax.xaxis.set_major_formatter(StrMethodFormatter("{x:.0%}"))
     top_ax = bottom_ax.secondary_xaxis(
         location="top",
         functions=(lambda x: 1. - x, lambda x: 1. - x),
     )
-    top_ax.set_xlabel("assignment to cluster A")
+    top_ax.set_xlabel("assignment to component A")
     top_ax.xaxis.set_major_formatter(StrMethodFormatter("{x:.0%}"))
     bottom_ax.set_yticks([])
     bottom_ax.grid(axis="x", alpha=0.5, color=USZ["gray"], linestyle=":")

### 2024-01-26 scripts/prevalence_comparison.py

@@ -5,9 +5,12 @@ compare it to the prevalence of involvement in the data.
 import argparse
 from pathlib import Path
 import yaml
+import math
 
 import numpy as np
 import matplotlib.pyplot as plt
+from matplotlib.ticker import StrMethodFormatter
+from tueplots import figsizes, fontsizes
 import h5py
 from emcee.backends import HDFBackend
 from lyscripts.utils import load_patient_data, create_model_from_config
@@ -15,6 +18,7 @@ from lyscripts.predict.prevalences import (
     compute_observed_prevalence,
     generate_predicted_prevalences,
 )
+from lyscripts.plot.utils import COLORS as USZ
 
 from helpers import str2bool, OROPHARYNX_ICDS, SUBSITE, T_STAGE, LNLS
 
@@ -45,19 +49,19 @@ def create_parser() -> argparse.ArgumentParser:
         help="Path to the parameter file..",
     )
     parser.add_argument(
-        "-i", "--involvement", nargs=4, type=str2bool, required=True,
-        help="Involvement pattern to compare the prevalences for.",
+        "-l", "--lnl", type=str, choices=LNLS, default="II",
+        help="LNL to plot involvement prevalence for.",
     )
     parser.add_argument(
         "-s", "--subsite", type=str, default="C01",
         help="ICD code of the subsite to plot the prevalences for.",
     )
     parser.add_argument(
-        "--thin-indie", type=int, default=10,
+        "--thin-indie", type=int, default=2,
         help="Thinning factor for the independent model's chain.",
     )
     parser.add_argument(
-        "--thin-mixture", type=int, default=1000,
+        "--thin-mixture", type=int, default=200,
         help="Thinning factor for the mixture model's chain.",
     )
     return parser
@@ -72,18 +76,22 @@ def main():
     lymph_model = create_model_from_config(params)
     lymph_model.modalities = {"max_llh": [1., 1.]}
 
-    pattern = {"ipsi": {lnl: val for lnl, val in zip(LNLS, args.involvement)}}
+    pattern = {"ipsi": {}}
+    for lnl in LNLS:
+        pattern["ipsi"][lnl] = None
+        if lnl == args.lnl:
+            pattern["ipsi"][lnl] = True
+
     patient_data = load_patient_data(args.data)
     patient_data[T_STAGE] = ["all"] * len(patient_data)
 
-    if args.subsite in OROPHARYNX_ICDS:
[…]

### 2024-01-26 content/manuscript.qmd
add prev superplot and rerun

--- 2024-01-25 content/manuscript.qmd
@@
 ---
-title: A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer
+title: A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer depending on primary tumor location
 author:
@@
 
-We now have two types of parameters, the probabilities of tutor spread for the different models $\boldsymbol{\theta_m}$, and the mixing coefficients $\pi_{sm}$.
+We now have two types of parameters, the probabilities of tumor spread for the different models $\boldsymbol{\theta_m}$, and the mixing coefficients $\pi_{sm}$.
 Assuming a uniform prior in the interval $[0,1]$ for all parameters, the posterior distribution over the parameters $P \left( \boldsymbol{\theta}, \boldsymbol{\pi} \mid \mathbf{D} \right)$ is given by the likelihood @eq-mixture-distribution except for a normalisation constant.
@@
 
-...
-
-The just introduced model is capable of learning one set of or distribution over parameters $\boldsymbol{\theta}$ from a cohort of patients with a primary tumor in a given subsite.
-If we tried to train it with a cohort consisting of patients with tumors in two very different subsites, the model would likely learn parameters that represent a compromise between the two subcohort's true parameters.
-This compromise might describe neither of the subcohorts' lymphatic spread patterns sufficiently well.
-
-In such cases, mixture models are often considered. They assume the data to come from a finite mixture distribution, which -- in our particular case -- can be written as follows:
-
-$$
-P \left( \mathbf{D} \mid \boldsymbol{\Psi} \right) = \sum_{j=0}^g c_j P \left( \mathbf{D} \mid \boldsymbol{\theta}_j \right)
-$$ {#eq-mixture-distribution}
-
-Here, the $\mathbf{c} \in [0,1]^g$ is the vector of mixing proportions with $\sum_{j=0}^g c_j = 1$, while the $\boldsymbol{\Psi} = \left( \boldsymbol{\theta}_1, \ldots, \boldsymbol{\theta}_g\right)$ is the vector of all $g$ models' parameters.
-Note that we will implement our model such that some of the parameters in each $\boldsymbol{\theta}_j$ are shared across the $g$ components -- namely the $t_{r \rightarrow v}$ corresponding to the blue arcs in @fig-schematic-and-graph.
-
-Let now $\mathfrak{D} = \left( \mathbf{D}_1, \ldots, \mathbf{D}_s \right)$ be a dataset consisting of $s$ subcohorts of patients.
-Within a subcohort $i$ we find $N_i$ patients with tumors in the same subsite.
-We can then introduce a _latent variable_ $\mathbf{Z}$ with a one-hot-encoding: Basically, it can take on values $\mathbf{z}_i \in \{ 0,1 \}^g$ with $z_{ij}=1$ if subcohort $i$ belongs to component $j$ and $z_{ij}=0$ else.
-
-The latent variables are helpful in resolving the invariance of the likelihood w.r.t.
-permutations of the component labels, which may introduce problems, e.g.
-for common MCMC sampling methods.
-The $\mathbf{Z}$ allows us to derive two sets of interdependent equations that we may solve in an iterative fashion (see e.g.
-@bishop_pattern_2006 for a detailed derivation) that is commonly referred to as _expectation-maximization (EM)_ algorithm:
-
-The first set are the probabilities of subcohort $i$ to belong to component $j$, given a set of parameters $\boldsymbol{\Psi}^\star$. These are often called the _responsibilities_:
-$$
-\gamma (z_{ij}) = P \left( z_{ij}=1 \mid \mathbf{D}_i , \boldsymbol{\Psi}^\star, \mathbf{c} \right) = \frac{ c_j P \left( \mathbf{D}_i \mid \boldsymbol{\theta}_j^\star \right) }{\sum_{k=0}^g c_k P \left( \mathbf{D}_i \mid \boldsymbol{\theta}_k^\star \right)}
-$$
-
-From this, we can compute new mixing proportions $c_j^\star=\sum_{i=1}^s \gamma (z_{ij}) / s$ and then infer new parameters $\boldsymbol{\Psi}^\star$ -- e.g.
-via MCMC sampling -- from the resulting likelihood, which is the second set:
-
-$$
-P \left( \mathbf{D} \mid \boldsymbol{\Psi}, \mathbf{c}^\star \right) = \sum_{j=0}^g c_j^\star P \left( \mathbf{D} \mid \boldsymbol{\theta}_j \right)
-$$
-
-## Implementation {#sec-implementation}
-
-We did this in this and that fashion...
-
 ## Multi-Centric Data
@@
 
-1. 287 oropharyngeal patients from the University of Zurich in Swizerland
-2. 263 oropharyngeal patients from the Centre Léon Bérard in Fance
-3. 289 oropharyngeal and oral cavity patients from the Inselspital Bern in Swizerland
-4. 239 oropharyngeal and oral cavity patients from the Centre Léon Bérard in Fance
+1. 287 oropharyngeal patients from the University of Zurich in Switzerland
+2. 263 oropharyngeal patients from the Centre Léon Bérard in France
+3. 289 oropharyngeal and oral cavity patients from the Inselspital Bern in Switzerland
+4. 239 oropharyngeal and oral cavity patients from the Centre Léon Bérard in France
 5. 162 oropharyngeal patients from the Hospital Vall d'Hebron in Spain
@@
 The subsites are sorted in ascending order by their prevalence of involvement in LNL II.
-Oral cavity subsites are plotted in shaed of blue, oropharynx subsites in shades of orange.](../figures/prevalence_by_subsite.svg){#fig-prevalence-by-subsite fig-pos="b"}
+Oral cavity subsites are plotted in shades of blue, oropharynx subsites in shades of orange.](../figures/prevalence_by_subsite.svg){#fig-prevalence-by-subsite fig-pos="b"}
 
-The data comes in the form of CSV tables and are -- except for the last and most recent addition -- publicly available [@ludwig_detailed_2022;@ludwig_multi-centric_2023] and may be interactively explored in our **Ly**mphatic **Pro**gression e**X**plorer [LyProX](https://lyprox.org).
-Each row of these tables corresponds to one patient and details in which LNL metastatic involvement was found or not, according to different diagnostic and pathologic modalities.
+The data sets 1-4 are publicly available in the form of CSV tables [@ludwig_detailed_2022;@ludwig_multi-centric_2023] and may be interactively explored in our **Ly**mphatic **Pro**gression e**X**plorer [LyProX](https://lyprox.org).
+Data set 5 is not yet public but of similar format.
+For each patient, the primary tumor subsite is reported (among other patient and tumor characteristics) and each individual LNL is reported as metastatic or healthy, according to the available diagnostic modalities (in part pathology after neck dissection, otherwise clinical involvement).
 
-In @fig-prevalence-by-subsite, we have plotted the prevalence of involvement in the four LNLs I, II, III, and IV stratified by the primary tumor's subsite.
-We only included patients with tumors in the gum, floor of mouth, other/unspecified parts of the mouth, palate, oropharynx, tosil, base of tongue, and other/unspecified parts of tongue, resulting in {{< var num_patients.total >}} patients.
+In @fig-prevalence-by-subsite, we have plotted the prevalence of involvement in the four ipsilateral LNLs I, II, III, and IV stratified by the primary tumor's subsite.
+We included patients with tumors in the oral cavity, gum (C03), floor of mouth (C04), other/unspecified parts of the mouth (C06), other/unspecified parts of tongue (C02), palate (C05), and tumors in the oropharynx (C10), tonsil (C09), base of tongue (C01), resulting in {{< var num_patients.total >}} patients.
+The figure illustrates the variations in LNL involvement between subsites within the oral cavity and oropharynx categories.
 
@@
 
-Works super well, of course!
+![Assignment of each subsite to each of the two components.
+The further left a subsite, the more it is assigned to component A, the further right, the more to component B.
+The size of the marker (area) corresponds to the number of patients in the subsite.](../figures/cluster_assignments.svg){#fig-cluster-assignments}
 
-![Assignment of each subsite to each of the two clusters.
-The fruther left a subsite, the more it is assigned to cluster A, the further right, the more to cluster B.
-The size of the marker (area) corresponds to the number of patients in the subsite.](../figures/cluster_assignments.svg){#fig-cluster-assignments}
+We demonstrate the methodology for a mixture model with $M=2$ components, considering the ipsilateral involvement of LNLs I, II, III, and IV and the primary tumor subsites shown in @fig-prevalence-by-subsite.
+In @fig-cluster-assignments we show the resulting mixture coefficients $\pi_{sm}$.
+The interpretation of this result is as follows: Tumors of the base of tongue (C01) are fully described by component A, and tumors of the gum (C03) are fully described by component B.
+These two subsites are the most distinct regarding the involvement of LNLs I and II, and the result is thus intuitive.
+Component A may be interpreted as a model for oropharynx-like tumor spread, and component B as a model for oral cavity-like tumor spread.
+All other subsites are described as mixtures.
+Tumors in the tonsil (C09) have LNL involvement similar to base of tongue tumors and are mostly assigned to component A.
+Instead, tumors of the palate (C05) are to similar degree assigned to components A and B, which is consistent with the anatomical location and the observation that the LNL involvement is in between oropharynx and oral cavity-type patterns.
+
+![The prevalence of involvement as seen in the data (vertical dashed lines), predicted by an independent model for the oropharyngeal or oral cavity patients (outlined histograms), and predicted by the mixture model (filled histograms).
+Each row correpsonds to one subsite and each column to the predicted or observed prevalence in one LNL.](../figures/prevalence_comparison.svg){#fig-prevalence-comparison fig-env="figure*"}
 
@@
 
-All that's left is for other people to bring this into clinical practice.
+We have previously developed a model of lymphatic progression of HNSCC using HMM, which allows us to estimate the probability of occult lymph node metastases in clinically negative LNLs.
+Mixture models are a suitable method to incorporate the primary tumor location into the model, which allows us to account for differences in lymph node involvement for different subsites.
+Future work will extend the work to tumors in the hypopharynx and larynx and optimize the number of model components.

### 2024-01-26 scripts/helpers.py

@@ -1,9 +1,12 @@
 """
 Helper functions for the scripts.
 """
+from pathlib import Path
 from typing import Generator
 import argparse
+from emcee.backends import HDFBackend
 
+import h5py
 from lyscripts.plot.utils import COLORS as USZ
 from matplotlib.colors import to_rgba
 
@@ -58,3 +61,53 @@ def str2bool(v: str) -> bool | None:
         return None
     else:
         raise argparse.ArgumentTypeError('Boolean or None value expected.')
+
+
+def get_location(for_subsite):
+    """Return the location of the given subsite."""
+    if for_subsite in OROPHARYNX_ICDS:
+        return "oropharynx"
+    else:
+        return "oral cavity"
+
+
+def get_indie_chain(from_dir: Path, for_location: str, thin_by: int):
+    """Get the independent model's chain from the given directory."""
+    if for_location == "oropharynx":
+        indie_chain = HDFBackend(
+            from_dir / "oropharynx.hdf5",
+            read_only=True,
+        ).get_chain(flat=True, thin=thin_by)
+    else:
+        indie_chain = HDFBackend(
+            from_dir / "oral_cavity.hdf5",
+            read_only=True,
+        ).get_chain(flat=True, thin=thin_by)
+
+    return indie_chain
+
+
+def get_mixture_components(
+    from_dir: Path,
+    for_subsite: str,
+    icd_code_map: dict,
+) -> tuple[float, float]:
+    """Get the mixture model's components from the given directory."""
+    with h5py.File(from_dir / "mixture.hdf5", mode="r") as h5_file:
+        component_assignments = h5_file["em/cluster_components"][...]
+
+    idx = list(icd_code_map.keys()).index(for_subsite)
+    comp_A_prob = component_assignments[idx]
+    comp_B_prob = 1 - comp_A_prob
+
+    return comp_A_prob, comp_B_prob
+
+
+def get_prevalence_pattern(for_lnl: str) -> dict[str, dict[str, bool | None]]:
+    """Create a pattern correpsonding to the prevalence of one LNL's involvement."""
+    pattern = {"ipsi": {}}
+    for lnl in LNLS:
+        pattern["ipsi"][lnl] = None
+        if lnl == for_lnl:
+            pattern["ipsi"][lnl] = True
+    return pattern

### 2024-01-26 scripts/prevalence_comparison.py

@@ -11,7 +11,6 @@ import numpy as np
 import matplotlib.pyplot as plt
 from matplotlib.ticker import StrMethodFormatter
 from tueplots import figsizes, fontsizes
-import h5py
 from emcee.backends import HDFBackend
 from lyscripts.utils import load_patient_data, create_model_from_config
 from lyscripts.predict.prevalences import (
@@ -20,8 +19,15 @@ from lyscripts.predict.prevalences import (
 )
 from lyscripts.plot.utils import COLORS as USZ
 
-from helpers import str2bool, OROPHARYNX_ICDS, SUBSITE, T_STAGE, LNLS
-
+from helpers import (
+    SUBSITE,
+    T_STAGE,
+    LNLS,
+    get_indie_chain,
+    get_location,
+    get_mixture_components,
+    get_prevalence_pattern,
+)
 
 
 def create_parser() -> argparse.ArgumentParser:
@@ -76,35 +82,26 @@ def main():
     lymph_model = create_model_from_config(params)
     lymph_model.modalities = {"max_llh": [1., 1.]}
 
-    pattern = {"ipsi": {}}
-    for lnl in LNLS:
-        pattern["ipsi"][lnl] = None
-        if lnl == args.lnl:
-            pattern["ipsi"][lnl] = True
+    pattern = get_prevalence_pattern(for_lnl=args.lnl)
+    location = get_location(for_subsite=args.subsite)
 
     patient_data = load_patient_data(args.data)
     patient_data[T_STAGE] = ["all"] * len(patient_data)
 
-    location = "oropharynx" if args.subsite in OROPHARYNX_ICDS else "oral cavity"
-    if location == "oropharynx":
-        indie_chain = HDFBackend(
-            args.mixture_model.parent / "oropharynx.hdf5",
-            read_only=True,
-        ).get_chain(flat=True, thin=args.thin_indie)
-    else:
-        indie_chain = HDFBackend(
-            args.mixture_model.parent / "oral_cavity.hdf5",
-            read_only=True,
-        ).get_chain(flat=True, thin=args.thin_indie)
-
+    indie_chain = get_indie_chain(
+        from_dir=args.mixture_model.parent,
+        for_location=location,
+        thin_by=args.thin_indie,
+    )
     mixture_chain = HDFBackend(
         args.mixture_model, read_only=True,
     ).get_chain(flat=True, thin=args.thin_mixture, discard=1000)
[…]

### 2024-01-26 scripts/prevalence_superplot.py

@@ -1,0 +1,207 @@
+"""
+Plot selected prevalence predictions into one, nice figure.
+"""
+import argparse
+from pathlib import Path
+import yaml
+from cycler import cycler
+
+import numpy as np
+import matplotlib.pyplot as plt
+from tueplots import figsizes, fontsizes
+from emcee.backends import HDFBackend
+from lyscripts.utils import load_patient_data, create_model_from_config
+from lyscripts.predict.prevalences import (
+    compute_observed_prevalence,
+    generate_predicted_prevalences,
+)
+from lyscripts.plot.utils import COLORS as USZ
+from lyscripts.plot.utils import Histogram, draw
+
+from helpers import (
+    SUBSITE,
+    T_STAGE,
+    LNLS,
+    get_indie_chain,
+    get_location,
+    get_mixture_components,
+    get_prevalence_pattern,
+)
+
+
+COLOR_CYCLER = cycler(color=[USZ["green"], USZ["blue"], USZ["orange"], USZ["red"]])
+
+
+def create_parser() -> argparse.ArgumentParser:
+    """Assemble the parser for the command line arguments."""
+    parser = argparse.ArgumentParser(description=__doc__)
+    parser.add_argument(
+        "-d", "--data", type=Path, default="data/enhanced.csv",
+        help="Path to the data file.",
+    )
+    parser.add_argument(
+        "--models", type=Path, default="models",
+        help=(
+            "Path to the directory containing the HDF5 files of the trained models. "
+            "The directory is assumed to contain `mixture.hdf5`, `oropharynx.hdf5`, "
+            "and `oral_cavity.hdf5`."
+        )
+    )
+    parser.add_argument(
+        "-o", "--output", type=Path, default="figures/prevalence_superplot.png",
+        help="Path to the output file.",
+    )
+    parser.add_argument(
+        "-p", "--params", type=Path, default="_variables.yml",
+        help="Path to the parameter file..",
+    )
+    parser.add_argument(
+        "-l", "--lnls", nargs="+", type=str, default=["I", "II", "III"],
+        help="LNLs to plot involvement prevalences for.",
+    )
+    parser.add_argument(
+        "-s", "--subsites", nargs="+", type=str, default=["C01", "C09", "C05"],
+        help=(
+            "ICD codes of the subsites to plot the prevalences for. Each subsite will "
+            "be plotted in a separate row."
+        ),
+    )
+    parser.add_argument(
+        "--thin-indie", type=int, default=2,
[…]

### 2024-01-26 content/manuscript.tex
add compiled tex to git

\documentclass[11pt,twocolumn,twoside]{article}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{float}
\makeatletter
\let\oldlt\longtable
\let\endoldlt\endlongtable
\def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
\def\longtable@i[#1]{\begin{figure}[H]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldlt[#1]
}
\def\longtable@ii{\begin{figure}[H]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldlt
}
\def\endlongtable{\endoldlt
\end{minipage}
\twocolumn
\end{figure}}
\makeatother
\usepackage{iccr2024}
\usepackage{newtxtext,newtxmath}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\addbibresource{../references.bib}
\title{A mixture of hidden Markov models to predict the lymphatic spread
in head and neck cancer depending on primary tumor location}

\author[%
1,2%
]{Roman Ludwig}
\author[%
1,2%
]{Julian Brönnimann}
\author[%
1,2%
]{Yoel Perez Haas}
\author[%
1,2%
]{Esmée Lauren Looman}
\author[%
11%
]{Sergi Benavente}
\author[%
3,4,7%
]{Adrian Schubert}
\author[%
8%
]{Dorothea Barbatei}
\author[%
8%
]{Laurence Bauwens}
\author[%
2%
]{Jean-Marc Hoffmann}
\author[%
4,5%
]{Sandrine Werlen}
\author[%
3%
]{Olgun Elicin}
\author[%
6,10%
]{Matthias Dettmer}
\author[%
9%
]{Philippe Zrounba}
\author[%
2%
]{Bertrand Poumayou}
\author[%
2%
]{Panagiotis Balermpas}
\author[%
8%
]{Vincent Grégoire}
\author[%
4,5%
]{Roland Giger}
\author[%
1,2%
]{Jan Unkelbach}

\affil[1]{Department of Physics, %
University of Zurich%
, Zurich%
%
%
, Switzerland}
\affil[2]{Department of Radiation Oncology, %
University Hospital Zurich%
, Zurich%
%
%
, Switzerland}
\affil[3]{Department of Radiation Oncology, %
Bern University Hospital%
, Bern%
%
%
, Switzerland}
\affil[4]{Department of ENT, Head \& Neck Surgery, %
Bern University Hospital%
, Bern%
%
%
, Switzerland}
\affil[5]{Head and Neck Anticancer Center, %
Bern University Hospital%
, Bern%
%
%
, Switzerland}
\affil[6]{Institute of Tissue Medicine and Pathology, %
Bern University Hospital%
, Bern%
%
%
, Switzerland}
\affil[7]{Department of ENT, Head \& Neck Surgery, %
Réseau Hospitalier Neuchâtelois%
, Neuchâtelois%
%
%
, Switzerland}
\affil[8]{Department of Radiation Oncology, %
Centre Léon Bérard%
, Lyon%
%
%
, France}
\affil[9]{Department of Head and Neck Surgery, %
Centre Léon Bérard%
, Lyon%
%
%
, France}
\affil[10]{Institute of Pathology, %
Klinikum Stuttgart%
, Stuttgart%
%
%
, Germany}
\affil[11]{Departement of Radiation Oncology, %
Hospital Vall d'Hebron%
, Barcelona%
%
%
, Spain}

\date{}
\begin{document}

\maketitle
\thispagestyle{fancy}

\begin{customabstract}
We previously developed a mechanistic hidden Markov model (HMM) to
predict the lymphatic tumor progression in oropharyngeal squamous cell
carcinomas (OPSCCs). To extend the model to other tumor subsites,
defined by ICD-10 codes, in the head and neck, we employ a mixture of
these HMMs and learn the cluster assignments and model parameters in an
iterative, EM-like algorithm from multi-centric data. The mixture model
manages to group anatomically close subsites and correctly infers the
clusters' parameters. Using this mixture model allows the prediction of
individual risks of occult nodal disease, given a diagnosis that
includes tumor subsite.
\end{customabstract}

\section{Introduction}\label{introduction}

Head and neck squamous cell carcinomas (HNSCC) frequently spread through
the lymphatic system
\autocite{lindberg_distribution_1972,woolgar_histological_1999}. Current
diagnostic imaging modalities are unable to detect microscopic lymph
node metastases \autocite{snyder_petct_2021,strohl_petct_2021}. To avoid
nodal recurrences \autocite{ho_decision_2014}, large volumes in the neck
are irradiated electively, which are at risk of harbouring occult
disease. Guidelines about which lymph node levels (LNLs) to irradiate
\autocite{biau_selection_2019} are currently not based on a patient's
individual risk, but only on the overall prevalence of nodal disease as
reported in the literature
\autocite{lindberg_distribution_1972,woolgar_histological_1999}.

To personalize this prediction of the risk for occult disease, given a
patient's individual diagnosis, we published

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  large, multi-centric data that reports per patient which LNLs where
  clinically and/or pathologically involved
  \autocite{ludwig_dataset_2022,ludwig_multi-centric_2023}.
\end{enumerate}

And, building on this work,

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  an interpretable hidden Markov model (HMM), trained with this data, to
  predict the risk for occult nodal disease
  \autocite{ludwig_hidden_2021}, given an individual patient's
  diagnosis.
\end{enumerate}

Such a personalized risk prediction may allow clinicians to safely
reduce the elective clinical target volume (CTV-N) and thus reduce
side-effects that degrade the patient's quality of life
\autocite{batth_practical_2014}.

Here, we extend the previous work by incorporating the primary tumor
location (specified as ICD-10 code) into the model of lymphatic tumour
progression, focusing on tumours in the oropharynx and the oral cavity.
HNSCC patients with primary tumors at different subsites show different
patterns of lymphatic spread
\autocite{lindberg_distribution_1972,woolgar_histological_1999}. So far,
this could be handled by training different models for broader
categories of tumour locations, e.g.~oropharynx and oral cavity tumours.
However, this approach does not describe differences in lymphatic spread
between different subsites within the oropharynx and oral cavity. To
address this issue, we present an approach using mixtures of HMMs. The
intuition is that the lymphatic spread of a tumor that lies anatomically
at the boarder of oropharynx and oral cavity (e.g.~tumours in the
palate) may be described by a mixture of different models.

\section{Materials and Methods}\label{sec-materials-and-methods}

Each LNL \(v \in \{ 1, 2, \ldots, V\}\) considered in our model is
represented by a binary random variable (RV) \(X_v\) representing the
true state of that level (0 for ``healthy'' and 1 for ``involved''). A
patient's state of lymph node involvement can be represented in a random
vector \(\mathbf{X} = \left( X_1, X_2, \ldots, X_V \right)\). When a
patient is diagnosed with HNSCC, we only observe the clinical lymph node
involvement based on imaging, which we denote as another binary random
variable \(Y_v\). To compute the personalized risk of occult disease
\(\mathbf{X}\), given a diagnosis \(\mathbf{Y}\), we apply Bayes' law:

\begin{equation}\phantomsection\label{eq-bayes-law}{
P \left( \mathbf{X} \mid \mathbf{Y} \right) = \frac{P \left( \mathbf{Y} \mid \mathbf{X} \right) P \left( \mathbf{X} \right)}{\sum_{\mathbf{X}^\star} P \left( \mathbf{Y} \mid \mathbf{X}^\star \right) P \left( \mathbf{X}^\star \right)}
}\end{equation}

In the above equation, the term
\(P \left( \mathbf{Y} \mid \mathbf{X} \right)\) is given by the
sensitivity and specificity of the diagnostic procedure. The term
\(P \left( \mathbf{X} \right)\) represents the prior probability of
involvement, which depends on the probability of the tumour to spread
through the lymphatic system. The main task of the HMM is to model
\(P \left( \mathbf{X} \right)\) and the main contribution of this paper
is to incorporate the primary tutor subset into the model of
\(P \left( \mathbf{X} \right)\).

\subsection{Hidden Markov Model for Lymphatic
Progression}\label{sec-hmm}

A patient's state of lymph node involvement \(\mathbf{X}[t]\) evolves
over discrete time steps \(t\). Let us enumerate all \(2^V\) possible
states, representing all combinations of LNLs. In this paper, we
consider ipsilateral LNLs I, II, III and IV, which amounts to 16
possible states. The HMM is specified by \emph{transition matrix}
\(\mathbf{A}\):

\begin{equation}\phantomsection\label{eq-transition-matrix}{
\mathbf{A} = \begin{pmatrix} A_{ij} \end{pmatrix} = P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right)
}\end{equation}

which contains the conditional probabilities that a state
\(\mathbf{X}[t]=\boldsymbol{\xi}_i\) transitions to
\(\mathbf{X}[t+1]=\boldsymbol{\xi}_j\) over one time step. The
transition matrix is specified and parameterised via the graphical model
shown in figure~\ref{fig-schematic-and-graph}. The red arcs in the graph
of figure~\ref{fig-schematic-and-graph} are associated the probability
that the primary tumor spreads directly to a LNL (parameters \(b_v\)).
The blue arcs describe the spread from an upstream LNL -- given it is
already metastatic -- to a downstream level (parameters
\(t_{v-1 \rightarrow v}\)).

Now, let \(\boldsymbol{\pi}\) be the \emph{starting distribution}

\begin{equation}\phantomsection\label{eq-starting-distribution}{
\boldsymbol{\pi} = \begin{pmatrix} \pi_i \end{pmatrix} = P \left( \mathbf{X}[0] = \boldsymbol{\xi}_i \right)
}\end{equation}

denoting the probability to start in state \(\boldsymbol{\xi}_i\) at
time step 0. Assuming that every patient started their disease with all
LNLs being healthy, we set \(\pi_i\) to zero for all states except the
completly healthy state
\(\boldsymbol{\xi} = \begin{pmatrix} 0, 0, 0, 0 \end{pmatrix}\), which
has probability 1.

Using the quantities introduced so far, the probability
\(P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right)\) to be in state
\(i\) in time step \(t\) can now be conveniently expressed as a matrix
product:

\begin{equation}\phantomsection\label{eq-evolution}{
P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right) = \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
}\end{equation}

This evolution implicitly marginalizes over all possible paths to arrive
at state \(\boldsymbol{\xi}_i\) after \(t\) time-steps. Additionally, we
must marginalize over the unknown time of diagnosis using a time-prior
\(P(t)\). This finally defines the probability distribution over all
states of lymph node involvement used in equation~\ref{eq-bayes-law}.

\begin{equation}\phantomsection\label{eq-marginalized-evolution}{
P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta} \right) = \sum_{t=0}^{t_\text{max}} P_T(t) \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
}\end{equation}

where \(\boldsymbol{\theta}=\{ b_v, t_{r \rightarrow v} \}\) denotes the
set of all model parameters (7 in our case). Fortunately, the exact
length and shape of this distribution on its own has little impact as
previously shown. We set \(t_\text{max}=\) 10 and \(P_\text{early}(t)\)
to a binomial distribution with parameter 0.3. Further details on the
HMM can be found in \ldots{}

\begin{figure}[th]

\centering{

\includegraphics{manuscript_files/mediabag/../static/schematic_and_graph.pdf}

}

\caption{\label{fig-schematic-and-graph}On the left: Rough anatomical
sketch of the tumor subsites and correspsonding ICD-10 codes that are
present in the used data. The subsite ``other parts of mouth'' (C06) was
not drawn. On the right: Parametrized graph representation of the
lymphatic network considering four LNLs. Blue nodes represent the hidden
RVs, while the red one is the tumor. Arcs represent a conditional
probability parametrized with the quantity noted next to it}

\end{figure}%

\subsection{Mixture of HMMs}\label{sec-mixture-of-hmms}

Let us now assume that primary tumors at different subsites have
different patterns of lymphatic spread, corresponding to different model
parameters \(\boldsymbol{\theta}\). Training a separate model for every
possible subsite (ICD-10 code) would require a sufficiently large
dataset for every tumor site. However, anatomically nearby locations are
expected to show very similar patterns of LNL involvement. Therefore, we
consider a mixture model.

Let us assume that we have a dataset \(\mathbf{D}\) that is specified
via the number of patients \(N_{is}\) that were diagnosed in LNL
involvement state \(i\) and had a primary tutor in subsite \(s\). Let us
further assume that we want to describe this dataset using a mixture of
\(M\) HMMs, each with a different set of model parameters
\(\boldsymbol{\theta_m}\). As the generative model of the data, we
assume that a patient with subsite \(s\) is generated with probability
\(\pi_{sm}\) from model \(m\). The likelihood of the dataset can then be
written as

\begin{equation}\phantomsection\label{eq-mixture-distribution}{
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_s \prod_i \left[ \sum_{m=1}^M \pi_{sm} P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m \right) \right]^{N_{is}}
}\end{equation}

We now have two types of parameters, the probabilities of tumor spread
for the different models \(\boldsymbol{\theta_m}\), and the mixing
coefficients \(\pi_{sm}\). Assuming a uniform prior in the interval
\([0,1]\) for all parameters, the posterior distribution over the
parameters
\(P \left( \boldsymbol{\theta}, \boldsymbol{\pi} \mid \mathbf{D} \right)\)
is given by the likelihood equation~\ref{eq-mixture-distribution} except
for a normalisation constant. In this work, we use Markov chain Monte
Carlo sampling (MCMC) via the \texttt{emcee} Python package
\autocite{foreman-mackey_emcee_2013} to sample model parameters from the
posterior distribution. However,
\(P \left( \boldsymbol{\theta}, \boldsymbol{\pi} \mid \mathbf{D} \right)\)
itself is a multi-model distribution because one can permute the
different models. To address this problem, we revert to an
\emph{expectation-maximization (EM)} algorithm where we iteratively
sample model parameters \(\boldsymbol{\theta_m}\) using MCMC and then
determine the most likely mixing coefficients.

\subsection{Multi-Centric Data}\label{multi-centric-data}

For the analyses in this work, we used five datasets from four different
institutions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  287 oropharyngeal patients from the University of Zurich in
  Switzerland
\item
  263 oropharyngeal patients from the Centre Léon Bérard in France
\item
  289 oropharyngeal and oral cavity patients from the Inselspital Bern
  in Switzerland
\item
  239 oropharyngeal and oral cavity patients from the Centre Léon Bérard
  in France
\item
  162 oropharyngeal patients from the Hospital Vall d'Hebron in Spain
\end{enumerate}

\begin{figure}[b]

\centering{

\includegraphics{manuscript_files/mediabag/../figures/prevalence_by_subsite.pdf}

}

\caption{\label{fig-prevalence-by-subsite}Prevalence of LNL involvement
stratified by subsite. The subsites are sorted in ascending order by
their prevalence of involvement in LNL II. Oral cavity subsites are
plotted in shades of blue, oropharynx subsites in shades of orange.}

\end{figure}%

The data sets 1-4 are publicly available in the form of CSV tables
\autocite{ludwig_detailed_2022,ludwig_multi-centric_2023} and may be
interactively explored in our \textbf{Ly}mphatic \textbf{Pro}gression
e\textbf{X}plorer \href{https://lyprox.org}{LyProX}. Data set 5 is not
yet public but of similar format. For each patient, the primary tumor
subsite is reported (among other patient and tumor characteristics) and
each individual LNL is reported as metastatic or healthy, according to
the available diagnostic modalities (in part pathology after neck
dissection, otherwise clinical involvement).

In figure~\ref{fig-prevalence-by-subsite}, we have plotted the
prevalence of involvement in the four ipsilateral LNLs I, II, III, and
IV stratified by the primary tumor's subsite. We included patients with
tumors in the oral cavity, gum (C03), floor of mouth (C04),
other/unspecified parts of the mouth (C06), other/unspecified parts of
tongue (C02), palate (C05), and tumors in the oropharynx (C10), tonsil
(C09), base of tongue (C01), resulting in 1242 patients. The figure
illustrates the variations in LNL involvement between subsites within
the oral cavity and oropharynx categories.

\section{Results}\label{sec-results}

\begin{figure}

\centering{

\includegraphics{manuscript_files/mediabag/../figures/cluster_assignments.pdf}

}

\caption{\label{fig-cluster-assignments}Assignment of each subsite to
each of the two components. The further left a subsite, the more it is
assigned to component A, the further right, the more to component B. The
size of the marker (area) corresponds to the number of patients in the
subsite.}

\end{figure}%

We demonstrate the methodology for a mixture model with \(M=2\)
components, considering the ipsilateral involvement of LNLs I, II, III,
and IV and the primary tumor subsites shown in
figure~\ref{fig-prevalence-by-subsite}. In
figure~\ref{fig-cluster-assignments} we show the resulting mixture
coefficients \(\pi_{sm}\). The interpretation of this result is as
follows: Tumors of the base of tongue (C01) are fully described by
component A, and tumors of the gum (C03) are fully described by
component B. These two subsites are the most distinct regarding the
involvement of LNLs I and II, and the result is thus intuitive.
Component A may be interpreted as a model for oropharynx-like tumor
spread, and component B as a model for oral cavity-like tumor spread.
All other subsites are described as mixtures. Tumors in the tonsil (C09)
have LNL involvement similar to base of tongue tumors and are mostly
assigned to component A. Instead, tumors of the palate (C05) are to
similar degree assigned to components A and B, which is consistent with
the anatomical location and the observation that the LNL involvement is
in between oropharynx and oral cavity-type patterns.

\begin{figure*}

\centering{

\includegraphics{manuscript_files/mediabag/../figures/prevalence_comparison.pdf}

}

\caption{\label{fig-prevalence-comparison}The prevalence of involvement
as seen in the data (vertical dashed lines), predicted by an independent
model for the oropharyngeal or oral cavity patients (outlined
histograms), and predicted by the mixture model (filled histograms).
Each row correpsonds to one subsite and each column to the predicted or
observed prevalence in one LNL.}

\end{figure*}%

\section{Discussion}\label{sec-discussion}

We have previously developed a model of lymphatic progression of HNSCC
using HMM, which allows us to estimate the probability of occult lymph
node metastases in clinically negative LNLs. Mixture models are a
suitable method to incorporate the primary tumor location into the
model, which allows us to account for differences in lymph node
involvement for different subsites. Future work will extend the work to
tumors in the hypopharynx and larynx and optimize the number of model
components.

\printbibliography

\end{document}

### 2024-01-28 scripts/prevalence_by_subsite.py
fix schamtic and labels on figures

@@ -70,7 +70,7 @@ def main():
     )
 
     ax.grid(axis="y")
-    ax.set_xlabel("Lymph node level")
+    ax.set_xlabel("Lymph node level (ipsilateral)")
     ax.set_xticklabels(["I", "II", "III", "IV"], rotation=0)
     ax.set_ylabel("Prevalence [%]")
     ax.legend(fontsize="x-small", labelspacing=0.3)

### 2024-01-28 scripts/prevalence_superplot.py

@@ -195,7 +195,7 @@ def main():
             ax.legend(fontsize="x-small", loc="upper right", labelspacing=0.2)
 
     for ax, lnl in zip(axes[0], args.lnls):
-        ax.set_title(f"LNL {lnl}")
+        ax.set_title(f"ipsi LNL {lnl}")
 
     for ax in axes[-1]:
         ax.set_xlabel("prevalence [%]")

### 2024-01-28 content/manuscript.qmd
incorporate Jan's changes into source manuscript

@@
       - ref: 2
+  - name: Panagiotis Balermpas
+    orcid: 0000-0001-5261-6446
+    affiliations:
+      - ref: 2
   - name: Sergi Benavente
@@
       - ref: 2
-  - name: Sandrine Werlen
-    affiliations:
-      - ref: 4
-      - ref: 5
   - name: Olgun Elicin
@@
       - ref: 10
-  - name: Philippe Zrounba
-    orcid: 0000-0001-6223-6113
-    affiliations:
-      - ref: 9
   - name: Bertrand Poumayou
@@
       - ref: 2
-  - name: Panagiotis Balermpas
-    orcid: 0000-0001-5261-6446
-    affiliations:
-      - ref: 2
-  - name: Vincent Grégoire
-    affiliations:
-      - ref: 8
   - name: Roland Giger
@@
       - ref: 5
+  - name: Vincent Grégoire
+    affiliations:
+      - ref: 8
   - name: Jan Unkelbach
@@
 abstract: |
-  We previously developed a mechanistic hidden Markov model (HMM) to predict the
-  lymphatic tumor progression in oropharyngeal squamous cell carcinomas (OPSCCs). To
-  extend the model to other tumor subsites, defined by ICD-10 codes, in the head and
-  neck, we employ a mixture of these HMMs and learn the cluster assignments and model
-  parameters in an iterative, EM-like algorithm from multi-centric data. The mixture
-  model manages to group anatomically close subsites and correctly infers the
-  clusters' parameters. Using this mixture model allows the prediction of individual
-  risks of occult nodal disease, given a diagnosis that includes tumor subsite.
+  We previously developed a mechanistic hidden Markov model (HMM) to predict the lymphatic tumor progression in oropharyngeal squamous cell carcinomas.
+To extend the model to other tumor subsites in the head and neck defined by ICD-10 codes, we develop a mixture model combining multiple HMMs.
+The mixture coefficients and the model parameters are learned via an EM-like algorithm from a large multi-centric dataset on lymph node involvement.
+The methodology is demonstrated for tumors in the oropharynx and oral cavity.
+The mixture model groups anatomically nearly subsites and yields interpretable mixture coefficients consistent with anatomical location.
+It allows the prediction of differences in lymph node involvement depending on tumor subsite.
 format:
@@
 Current diagnostic imaging modalities are unable to detect microscopic lymph node metastases [@snyder_petct_2021;@strohl_petct_2021].
-To avoid nodal recurrences [@ho_decision_2014], large volumes in the neck are irradiated electively, which are at risk of harbouring occult disease.
-Guidelines about which lymph node levels (LNLs) to irradiate [@biau_selection_2019] are currently not based on a patient's individual risk, but only on the overall prevalence of nodal disease as reported in the literature [@lindberg_distribution_1972;@woolgar_histological_1999].
+To avoid nodal recurrences, large volumes in the neck, which are at risk of harbouring occult disease, are irradiated electively.
+Guidelines about which lymph node levels (LNLs) to irradiate [@biau_selection_2019] are currently not based on a patient's individual risk, but on the overall prevalence of nodal disease as reported in the literature [@lindberg_distribution_1972;@woolgar_histological_1999].
 
-To personalize this prediction of the risk for occult disease, given a patient's individual diagnosis, we published
+To personalize the prediction of the risk for occult disease, given a patient's individual diagnosis, we published
 
-1. large, multi-centric data that reports per patient which LNLs where clinically and/or pathologically involved [@ludwig_dataset_2022;@ludwig_multi-centric_2023].
+1. large, multi-centric data that reports per patient which LNLs were clinically/pathologically involved [@ludwig_dataset_2022;@ludwig_multi-centric_2023].
 
@@
 
-2. an interpretable hidden Markov model (HMM), trained with this data, to predict the risk for occult nodal disease [@ludwig_hidden_2021], given an individual patient's diagnosis.
+2. an interpretable hidden Markov model (HMM), trained with this data, to predict the risk for occult nodal disease, given an individual patient's diagnosis [@ludwig_hidden_2021].
 
@@
 
-Here, we extend the previous work by incorporating the primary tumor location (specified as ICD-10 code) into the model of lymphatic tumour progression, focusing on tumours in the oropharynx and the oral cavity.
+Here, we extend the previous work by incorporating the primary tumor location (specified as ICD-10 code) into the model of lymphatic tumor progression, focusing on tumors in the oropharynx and the oral cavity.
 HNSCC patients with primary tumors at different subsites show different patterns of lymphatic spread [@lindberg_distribution_1972;@woolgar_histological_1999].
-So far, this could be handled by training different models for broader categories of tumour locations, e.g.
-oropharynx and oral cavity tumours.
+So far, this could be handled by training different models for broader categories of tumor locations, e.g.
+oropharynx and oral cavity tumors.
 However, this approach does not describe differences in lymphatic spread between different subsites within the oropharynx and oral cavity.
@@
 The intuition is that the lymphatic spread of a tumor that lies anatomically at the boarder of oropharynx and oral cavity (e.g.
-tumours in the palate) may be described by a mixture of different models.
+tumors in the palate) may be described by a mixture of different models.
 
@@
 
-Each LNL $v \in \{ 1, 2, \ldots, V\}$ considered in our model is represented by a binary random variable (RV) $X_v$ representing the true state of that level (0 for "healthy" and 1 for "involved").
+Each LNL $v \in \{ 1, 2, \ldots, V\}$ considered in our model is represented by a binary random variable $X_v$ representing the true state of that level (0 for "healthy" and 1 for "involved").
 A patient's state of lymph node involvement can be represented in a random vector $\mathbf{X} = \left( X_1, X_2, \ldots, X_V \right)$.
@@
 In the above equation, the term $P \left( \mathbf{Y} \mid \mathbf{X} \right)$ is given by the sensitivity and specificity of the diagnostic procedure.
-The term $P \left( \mathbf{X} \right)$ represents the prior probability of involvement, which depends on the probability of the tumour to spread through the lymphatic system.
-The main task of the HMM is to model $P \left( \mathbf{X} \right)$ and the main contribution of this paper is to incorporate the primary tutor subset into the model of $P \left( \mathbf{X} \right)$.
+The term $P \left( \mathbf{X} \right)$ represents the prior probability of involvement, which depends on the probability of the tumor to spread through the lymphatic system.
+The main task of the HMM is to model $P \left( \mathbf{X} \right)$ and the main contribution of this paper is to incorporate the primary tumor subsite into the model of $P \left( \mathbf{X} \right)$.
 
@@
 In this paper, we consider ipsilateral LNLs I, II, III and IV, which amounts to 16 possible states.
-The HMM is specified by _transition matrix_ $\mathbf{A}$:
+The HMM is specified by a _transition matrix_:
 
@@
 
-which contains the conditional probabilities that a state $\mathbf{X}[t]=\boldsymbol{\xi}_i$ transitions to $\mathbf{X}[t+1]=\boldsymbol{\xi}_j$ over one time step.
+whose elements $A_{ij}$ contain the conditional probabilities that a state $\mathbf{X}[t]=\boldsymbol{\xi}_i$ transitions to $\mathbf{X}[t+1]=\boldsymbol{\xi}_j$ over one time step.
 The transition matrix is specified and parameterised via the graphical model shown in @fig-schematic-and-graph.
-The red arcs in the graph of @fig-schematic-and-graph are associated the probability that the primary tumor spreads directly to a LNL (parameters $b_v$).
-The blue arcs describe the spread from an upstream LNL -- given it is already metastatic -- to a downstream level (parameters $t_{v-1 \rightarrow v}$).
+The red arcs in the graph of @fig-schematic-and-graph (right panel) are associated with the probability that the primary tumor spreads directly to a LNL (parameters $b_v$).
+The blue arcs describe the spread from an upstream LNL -- given it is already metastatic -- to a downstream level (parameters $t_{v \rightarrow v+1}$).
 
@@
 denoting the probability to start in state $\boldsymbol{\xi}_i$ at time step 0.
-Assuming that every patient started their disease with all LNLs being healthy, we set $\pi_i$ to zero for all states  except the completly healthy state $\boldsymbol{\xi} = \begin{pmatrix} 0, 0, 0, 0 \end{pmatrix}$, which has probability 1.
+Assuming that every patient started with all LNLs being healthy, we set $\pi_i$ to zero for all states  except the completly healthy state $\boldsymbol{\xi} = \begin{pmatrix} 0, 0, 0, 0 \end{pmatrix}$, which has probability one.
 
-Using the quantities introduced so far, the probability $P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right)$ to be in state $i$ in time step $t$ can now be conveniently expressed as a matrix product:
+Using the quantities introduced so far, the probability $P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right)$ to be in state $\boldsymbol{\theta}_i$ in time step $t$ can now be conveniently expressed as a matrix product:
 
@@
 This evolution implicitly marginalizes over all possible paths to arrive at state $\boldsymbol{\xi}_i$ after $t$ time-steps.
-Additionally, we must marginalize over the unknown time of diagnosis using a time-prior $P(t)$.
+Additionally, we must marginalize over the unknown time of diagnosis using a time-prior $P_T(t)$.
 This finally defines the probability distribution over all states of lymph node involvement used in @eq-bayes-law.
@@
 
-where $\boldsymbol{\theta}=\{ b_v, t_{r \rightarrow v} \}$ denotes the set of all model parameters (7 in our case).
-Fortunately, the exact length and shape of this distribution on its own has little impact as previously shown.
+where $\boldsymbol{\theta}=\{ b_v, t_{v \rightarrow v+1} \}$ denotes the set of all model parameters (7 in our case).
+Fortunately, the exact length and shape of this distribution has little impact as previously shown [@ludwig_hidden_2021].
 We set $t_\text{max}=$ {{< var model.max_t >}} and $P_\text{early}(t)$ to a binomial distribution with parameter {{< var model.first_binom_prob >}}.
-Further details on the HMM can be found in ...
+Further details on the HMM can be found in @ludwig_hidden_2021.
 
-![On the left: Rough anatomical sketch of the tumor subsites and correspsonding ICD-10 codes that are present in the used data.
+![On the left: Anatomical sketch of the tumor subsites and correspsonding ICD-10 codes considered in this work.
 The subsite "other parts of mouth" (C06) was not drawn.
-On the right: Parametrized graph representation of the lymphatic network considering four LNLs.
-Blue nodes represent the hidden RVs, while the red one is the tumor.
-Arcs represent a conditional probability parametrized with the quantity noted next to it](../static/schematic_and_graph.svg){#fig-schematic-and-graph fig-pos="th"}
+On the right: Parametrized graphical model of the lymphatic network considering four LNLs.
+Blue nodes represent the hidden states of LNLs $X_v$, while the red one is the tumor.
+Arcs represent possible routes of metastatic spread, associated with a probability.](../static/schematic_and_graph.svg){#fig-schematic-and-graph fig-pos="th"}
 
@@
 
-We now have two types of parameters, the probabilities of tumor spread for the different models $\boldsymbol{\theta_m}$, and the mixing coefficients $\pi_{sm}$.
-Assuming a uniform prior in the interval $[0,1]$ for all parameters, the posterior distribution over the parameters $P \left( \boldsymbol{\theta}, \boldsymbol{\pi} \mid \mathbf{D} \right)$ is given by the likelihood @eq-mixture-distribution except for a normalisation constant.
+We now have two types of parameters, the probabilities of tumor spread for the different models, $\boldsymbol{\theta_m}$, and the mixing coefficients $\pi_{sm}$.
+Assuming a uniform prior in the interval $[0,1]$ for all parameters, the posterior distribution over the parameters $P \left( \boldsymbol{\theta}, \boldsymbol{\pi} \mid \mathbf{D} \right)$ is given by the likelihood in @eq-mixture-distribution except for a normalisation constant.
 In this work, we use Markov chain Monte Carlo sampling (MCMC) via the `emcee` Python package [@foreman-mackey_emcee_2013] to sample model parameters from the posterior distribution.
 However, $P \left( \boldsymbol{\theta}, \boldsymbol{\pi} \mid \mathbf{D} \right)$ itself is a multi-model distribution because one can permute the different models.
-To address this problem, we revert to an _expectation-maximization (EM)_ algorithm where we iteratively sample model parameters $\boldsymbol{\theta_m}$ using MCMC and then determine the most likely mixing coefficients.
+To address this problem, we revert to an _expectation-maximization (EM)_ algorithm where we iterate two steps until convergence of the mixing coefficients.
+In the E-step, we sample model parameters $\boldsymbol{\theta_m}$ using MCMC for given mixing coefficients $\pi_{sm}$.
+In the M-step, we maximize the likelihood with respect to the mixing coefficients for given samples of $\boldsymbol{\theta_m}$
 
@@
 
-For the analyses in this work, we used five datasets from four different institutions:
+For the analyses in this work, we used five datasets from four different institutions, resulting in 1242 patients in total.
 
@@
 4. 239 oropharyngeal and oral cavity patients from the Centre Léon Bérard in France
-5. 162 oropharyngeal patients from the Hospital Vall d'Hebron in Spain
+5. 162 oropharyngeal patients from the Hospital Vall d'Hebron in Spain (not yet public)
 
-![Prevalence of LNL involvement stratified by subsite.
+![Prevalence of ipsilateral LNL involvement stratified by subsite.
 The subsites are sorted in ascending order by their prevalence of involvement in LNL II.
@@
 The data sets 1-4 are publicly available in the form of CSV tables [@ludwig_detailed_2022;@ludwig_multi-centric_2023] and may be interactively explored in our **Ly**mphatic **Pro**gression e**X**plorer [LyProX](https://lyprox.org).
-Data set 5 is not yet public but of similar format.
 For each patient, the primary tumor subsite is reported (among other patient and tumor characteristics) and each individual LNL is reported as metastatic or healthy, according to the available diagnostic modalities (in part pathology after neck dissection, otherwise clinical involvement).
 
-In @fig-prevalence-by-subsite, we have plotted the prevalence of involvement in the four ipsilateral LNLs I, II, III, and IV stratified by the primary tumor's subsite.
-We included patients with tumors in the oral cavity, gum (C03), floor of mouth (C04), other/unspecified parts of the mouth (C06), other/unspecified parts of tongue (C02), palate (C05), and tumors in the oropharynx (C10), tonsil (C09), base of tongue (C01), resulting in {{< var num_patients.total >}} patients.
+In @fig-prevalence-by-subsite, we plot the prevalence of involvement in the four ipsilateral LNLs I, II, III, and IV stratified by the primary tumor's subsite.
 The figure illustrates the variations in LNL involvement between subsites within the oral cavity and oropharynx categories.
+The number of patients for each subsite is indicated in @fig-cluster-assignments.
 
 # Results {#sec-results}
+
+We demonstrate the methodology for a mixture model with $M=2$ components, considering the ipsilateral involvement of LNLs I, II, III, and IV and the primary tumor subsites shown in @fig-prevalence-by-subsite.
+In @fig-cluster-assignments we show the resulting mixture coefficients $\pi_{sm}$.
+The interpretation of this result is as follows: tumors of the base of tongue (C01) are fully described by component A, and tumors of the gum (C03) are fully described by component B.
+These two subsites are the most distinct regarding the involvement of LNLs I and II, and the result is thus intuitive.
+Component A may be interpreted as a model for oropharynx-like tumor spread, and component B as a model for oral cavity-like tumor spread.
+All other subsites are described as mixtures.
+tumors in the tonsil (C09) have LNL involvement similar to base of tongue tumors and are mostly assigned to component A.
+Instead, tumors of the palate (C05) are to similar degree assigned to components A and B, which is consistent with the anatomical location and the observation that the LNL involvement is in between oropharynx and oral cavity-type patterns (@fig-prevalence-by-subsite).
 
@@
 
-We demonstrate the methodology for a mixture model with $M=2$ components, considering the ipsilateral involvement of LNLs I, II, III, and IV and the primary tumor subsites shown in @fig-prevalence-by-subsite.
-In @fig-cluster-assignments we show the resulting mixture coefficients $\pi_{sm}$.
-The interpretation of this result is as follows: Tumors of the base of tongue (C01) are fully described by component A, and tumors of the gum (C03) are fully described by component B.
-These two subsites are the most distinct regarding the involvement of LNLs I and II, and the result is thus intuitive.
-Component A may be interpreted as a model for oropharynx-like tumor spread, and component B as a model for oral cavity-like tumor spread.
-All other subsites are described as mixtures.
-Tumors in the tonsil (C09) have LNL involvement similar to base of tongue tumors and are mostly assigned to component A.
-Instead, tumors of the palate (C05) are to similar degree assigned to components A and B, which is consistent with the anatomical location and the observation that the LNL involvement is in between oropharynx and oral cavity-type patterns.
+The @fig-prevalence-comparison illustrates the model's predictions of the overall prevalence of lymph node involvement in LNLs I, II and III (filled histograms) for selected subsites, obtained by summing the probabilities of states where the respecive level is involved.
+The mixture model is compared to two independent HMM models trained for oral cavity and oropharynx (by pooling the respective subsites).
+The mixture model and the independent oropharynx model perform similarly for tonsil tumors (C09), which is the largest patient group, dominating the independent oropharynx model and the component B in the mixture model.
+However, the mixture model better predicts the higher prevalence in level II for base of tongue and palate tumors.
 
 ![The prevalence of involvement as seen in the data (vertical dashed lines), predicted by an independent model for the oropharyngeal or oral cavity patients (outlined histograms), and predicted by the mixture model (filled histograms).
-Each row correpsonds to one subsite and each column to the predicted or observed prevalence in one LNL.](../figures/prevalence_comparison.svg){#fig-prevalence-comparison fig-env="figure*"}
+Each row correpsonds to one subsite and each column to one LNL.](../figures/prevalence_comparison.svg){#fig-prevalence-comparison fig-env="figure*"}
 

### 2024-01-28 content/manuscript.tex
incorporate Jan's changes into source manuscript

@@
 \author[%
+2%
+]{Panagiotis Balermpas}
+\author[%
 11%
@@
 \author[%
-4,5%
-]{Sandrine Werlen}
-\author[%
 3%
@@
 \author[%
-9%
-]{Philippe Zrounba}
-\author[%
 2%
@@
 \author[%
-2%
-]{Panagiotis Balermpas}
+4,5%
+]{Roland Giger}
 \author[%
@@
 ]{Vincent Grégoire}
-\author[%
-4,5%
-]{Roland Giger}
 \author[%
@@
 predict the lymphatic tumor progression in oropharyngeal squamous cell
-carcinomas (OPSCCs). To extend the model to other tumor subsites,
-defined by ICD-10 codes, in the head and neck, we employ a mixture of
-these HMMs and learn the cluster assignments and model parameters in an
-iterative, EM-like algorithm from multi-centric data. The mixture model
-manages to group anatomically close subsites and correctly infers the
-clusters' parameters. Using this mixture model allows the prediction of
-individual risks of occult nodal disease, given a diagnosis that
-includes tumor subsite.
+carcinomas. To extend the model to other tumor subsites in the head and
+neck defined by ICD-10 codes, we develop a mixture model combining
+multiple HMMs. The mixture coefficients and the model parameters are
+learned via an EM-like algorithm from a large multi-centric dataset on
+lymph node involvement. The methodology is demonstrated for tumors in
+the oropharynx and oral cavity. The mixture model groups anatomically
+nearly subsites and yields interpretable mixture coefficients consistent
+with anatomical location. It allows the prediction of differences in
+lymph node involvement depending on tumor subsite.
 \end{customabstract}
@@
 node metastases \autocite{snyder_petct_2021,strohl_petct_2021}. To avoid
-nodal recurrences \autocite{ho_decision_2014}, large volumes in the neck
-are irradiated electively, which are at risk of harbouring occult
-disease. Guidelines about which lymph node levels (LNLs) to irradiate
+nodal recurrences, large volumes in the neck, which are at risk of
+harbouring occult disease, are irradiated electively. Guidelines about
+which lymph node levels (LNLs) to irradiate
 \autocite{biau_selection_2019} are currently not based on a patient's
-individual risk, but only on the overall prevalence of nodal disease as
+individual risk, but on the overall prevalence of nodal disease as
 reported in the literature
@@
 
-To personalize this prediction of the risk for occult disease, given a
+To personalize the prediction of the risk for occult disease, given a
 patient's individual diagnosis, we published
@@
 \item
-  large, multi-centric data that reports per patient which LNLs where
-  clinically and/or pathologically involved
+  large, multi-centric data that reports per patient which LNLs were
+  clinically/pathologically involved
   \autocite{ludwig_dataset_2022,ludwig_multi-centric_2023}.
@@
   an interpretable hidden Markov model (HMM), trained with this data, to
-  predict the risk for occult nodal disease
-  \autocite{ludwig_hidden_2021}, given an individual patient's
-  diagnosis.
+  predict the risk for occult nodal disease, given an individual
+  patient's diagnosis \autocite{ludwig_hidden_2021}.
 \end{enumerate}
@@
 Here, we extend the previous work by incorporating the primary tumor
-location (specified as ICD-10 code) into the model of lymphatic tumour
-progression, focusing on tumours in the oropharynx and the oral cavity.
+location (specified as ICD-10 code) into the model of lymphatic tumor
+progression, focusing on tumors in the oropharynx and the oral cavity.
 HNSCC patients with primary tumors at different subsites show different
@@
 this could be handled by training different models for broader
-categories of tumour locations, e.g.~oropharynx and oral cavity tumours.
+categories of tumor locations, e.g.~oropharynx and oral cavity tumors.
 However, this approach does not describe differences in lymphatic spread
@@
 intuition is that the lymphatic spread of a tumor that lies anatomically
-at the boarder of oropharynx and oral cavity (e.g.~tumours in the
-palate) may be described by a mixture of different models.
+at the boarder of oropharynx and oral cavity (e.g.~tumors in the palate)
+may be described by a mixture of different models.
 
@@
 Each LNL \(v \in \{ 1, 2, \ldots, V\}\) considered in our model is
-represented by a binary random variable (RV) \(X_v\) representing the
-true state of that level (0 for ``healthy'' and 1 for ``involved''). A
+represented by a binary random variable \(X_v\) representing the true
+state of that level (0 for ``healthy'' and 1 for ``involved''). A
 patient's state of lymph node involvement can be represented in a random
@@
 \(P \left( \mathbf{X} \right)\) represents the prior probability of
-involvement, which depends on the probability of the tumour to spread
+involvement, which depends on the probability of the tumor to spread
 through the lymphatic system. The main task of the HMM is to model
 \(P \left( \mathbf{X} \right)\) and the main contribution of this paper
-is to incorporate the primary tutor subset into the model of
+is to incorporate the primary tumor subsite into the model of
 \(P \left( \mathbf{X} \right)\).
@@
 consider ipsilateral LNLs I, II, III and IV, which amounts to 16
-possible states. The HMM is specified by \emph{transition matrix}
-\(\mathbf{A}\):
+possible states. The HMM is specified by a \emph{transition matrix}:
 
@@
 
-which contains the conditional probabilities that a state
-\(\mathbf{X}[t]=\boldsymbol{\xi}_i\) transitions to
+whose elements \(A_{ij}\) contain the conditional probabilities that a
+state \(\mathbf{X}[t]=\boldsymbol{\xi}_i\) transitions to
 \(\mathbf{X}[t+1]=\boldsymbol{\xi}_j\) over one time step. The
@@
 shown in figure~\ref{fig-schematic-and-graph}. The red arcs in the graph
-of figure~\ref{fig-schematic-and-graph} are associated the probability
-that the primary tumor spreads directly to a LNL (parameters \(b_v\)).
-The blue arcs describe the spread from an upstream LNL -- given it is
-already metastatic -- to a downstream level (parameters
-\(t_{v-1 \rightarrow v}\)).
+of figure~\ref{fig-schematic-and-graph} (right panel) are associated
+with the probability that the primary tumor spreads directly to a LNL
+(parameters \(b_v\)). The blue arcs describe the spread from an upstream
+LNL -- given it is already metastatic -- to a downstream level
+(parameters \(t_{v \rightarrow v+1}\)).
 
@@
 denoting the probability to start in state \(\boldsymbol{\xi}_i\) at
-time step 0. Assuming that every patient started their disease with all
-LNLs being healthy, we set \(\pi_i\) to zero for all states except the
-completly healthy state
+time step 0. Assuming that every patient started with all LNLs being
+healthy, we set \(\pi_i\) to zero for all states except the completly
+healthy state
 \(\boldsymbol{\xi} = \begin{pmatrix} 0, 0, 0, 0 \end{pmatrix}\), which
-has probability 1.
+has probability one.
 
@@
 \(P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right)\) to be in state
-\(i\) in time step \(t\) can now be conveniently expressed as a matrix
-product:
+\(\boldsymbol{\theta}_i\) in time step \(t\) can now be conveniently
+expressed as a matrix product:
 
@@
 must marginalize over the unknown time of diagnosis using a time-prior
-\(P(t)\). This finally defines the probability distribution over all
+\(P_T(t)\). This finally defines the probability distribution over all
 states of lymph node involvement used in equation~\ref{eq-bayes-law}.
@@
 
-where \(\boldsymbol{\theta}=\{ b_v, t_{r \rightarrow v} \}\) denotes the
-set of all model parameters (7 in our case). Fortunately, the exact
-length and shape of this distribution on its own has little impact as
-previously shown. We set \(t_\text{max}=\) 10 and \(P_\text{early}(t)\)
-to a binomial distribution with parameter 0.3. Further details on the
-HMM can be found in \ldots{}
+where \(\boldsymbol{\theta}=\{ b_v, t_{v \rightarrow v+1} \}\) denotes
+the set of all model parameters (7 in our case). Fortunately, the exact
+length and shape of this distribution has little impact as previously
+shown \autocite{ludwig_hidden_2021}. We set \(t_\text{max}=\) 10 and
+\(P_\text{early}(t)\) to a binomial distribution with parameter 0.3.
+Further details on the HMM can be found in
+\textcite{ludwig_hidden_2021}.
 
@@
 
-\caption{\label{fig-schematic-and-graph}On the left: Rough anatomical
-sketch of the tumor subsites and correspsonding ICD-10 codes that are
-present in the used data. The subsite ``other parts of mouth'' (C06) was
-not drawn. On the right: Parametrized graph representation of the
-lymphatic network considering four LNLs. Blue nodes represent the hidden
-RVs, while the red one is the tumor. Arcs represent a conditional
-probability parametrized with the quantity noted next to it}
+\caption{\label{fig-schematic-and-graph}On the left: Anatomical sketch
+of the tumor subsites and correspsonding ICD-10 codes considered in this
+work. The subsite ``other parts of mouth'' (C06) was not drawn. On the
+right: Parametrized graphical model of the lymphatic network considering
+four LNLs. Blue nodes represent the hidden states of LNLs \(X_v\), while
+the red one is the tumor. Arcs represent possible routes of metastatic
+spread, associated with a probability.}
 
@@
 We now have two types of parameters, the probabilities of tumor spread
-for the different models \(\boldsymbol{\theta_m}\), and the mixing
+for the different models, \(\boldsymbol{\theta_m}\), and the mixing
 coefficients \(\pi_{sm}\). Assuming a uniform prior in the interval
@@
 \(P \left( \boldsymbol{\theta}, \boldsymbol{\pi} \mid \mathbf{D} \right)\)
-is given by the likelihood equation~\ref{eq-mixture-distribution} except
-for a normalisation constant. In this work, we use Markov chain Monte
-Carlo sampling (MCMC) via the \texttt{emcee} Python package
+is given by the likelihood in equation~\ref{eq-mixture-distribution}
+except for a normalisation constant. In this work, we use Markov chain
+Monte Carlo sampling (MCMC) via the \texttt{emcee} Python package
 \autocite{foreman-mackey_emcee_2013} to sample model parameters from the
@@
 different models. To address this problem, we revert to an
-\emph{expectation-maximization (EM)} algorithm where we iteratively
-sample model parameters \(\boldsymbol{\theta_m}\) using MCMC and then
-determine the most likely mixing coefficients.
+\emph{expectation-maximization (EM)} algorithm where we iterate two
+steps until convergence of the mixing coefficients. In the E-step, we
+sample model parameters \(\boldsymbol{\theta_m}\) using MCMC for given
+mixing coefficients \(\pi_{sm}\). In the M-step, we maximize the
+likelihood with respect to the mixing coefficients for given samples of
+\(\boldsymbol{\theta_m}\)
 
@@
 For the analyses in this work, we used five datasets from four different
-institutions:
+institutions, resulting in 1242 patients in total.
 
@@
   162 oropharyngeal patients from the Hospital Vall d'Hebron in Spain
+  (not yet public)
 \end{enumerate}
@@
 
-\caption{\label{fig-prevalence-by-subsite}Prevalence of LNL involvement
-stratified by subsite. The subsites are sorted in ascending order by
-their prevalence of involvement in LNL II. Oral cavity subsites are
-plotted in shades of blue, oropharynx subsites in shades of orange.}
+\caption{\label{fig-prevalence-by-subsite}Prevalence of ipsilateral LNL
+involvement stratified by subsite. The subsites are sorted in ascending
+order by their prevalence of involvement in LNL II. Oral cavity subsites
+are plotted in shades of blue, oropharynx subsites in shades of orange.}
 
@@
 interactively explored in our \textbf{Ly}mphatic \textbf{Pro}gression
-e\textbf{X}plorer \href{https://lyprox.org}{LyProX}. Data set 5 is not
-yet public but of similar format. For each patient, the primary tumor
-subsite is reported (among other patient and tumor characteristics) and
-each individual LNL is reported as metastatic or healthy, according to
-the available diagnostic modalities (in part pathology after neck
-dissection, otherwise clinical involvement).
+e\textbf{X}plorer \href{https://lyprox.org}{LyProX}. For each patient,
+the primary tumor subsite is reported (among other patient and tumor
+characteristics) and each individual LNL is reported as metastatic or
+healthy, according to the available diagnostic modalities (in part
+pathology after neck dissection, otherwise clinical involvement).
 
-In figure~\ref{fig-prevalence-by-subsite}, we have plotted the
-prevalence of involvement in the four ipsilateral LNLs I, II, III, and
-IV stratified by the primary tumor's subsite. We included patients with
-tumors in the oral cavity, gum (C03), floor of mouth (C04),
-other/unspecified parts of the mouth (C06), other/unspecified parts of
-tongue (C02), palate (C05), and tumors in the oropharynx (C10), tonsil
-(C09), base of tongue (C01), resulting in 1242 patients. The figure
-illustrates the variations in LNL involvement between subsites within
-the oral cavity and oropharynx categories.
+In figure~\ref{fig-prevalence-by-subsite}, we plot the prevalence of
+involvement in the four ipsilateral LNLs I, II, III, and IV stratified
+by the primary tumor's subsite. The figure illustrates the variations in
+LNL involvement between subsites within the oral cavity and oropharynx
+categories. The number of patients for each subsite is indicated in
+figure~\ref{fig-cluster-assignments}.
 
 \section{Results}\label{sec-results}
+
+We demonstrate the methodology for a mixture model with \(M=2\)
+components, considering the ipsilateral involvement of LNLs I, II, III,
+and IV and the primary tumor subsites shown in
+figure~\ref{fig-prevalence-by-subsite}. In
+figure~\ref{fig-cluster-assignments} we show the resulting mixture
+coefficients \(\pi_{sm}\). The interpretation of this result is as
+follows: tumors of the base of tongue (C01) are fully described by
+component A, and tumors of the gum (C03) are fully described by
+component B. These two subsites are the most distinct regarding the
+involvement of LNLs I and II, and the result is thus intuitive.
+Component A may be interpreted as a model for oropharynx-like tumor
+spread, and component B as a model for oral cavity-like tumor spread.
+All other subsites are described as mixtures. tumors in the tonsil (C09)
+have LNL involvement similar to base of tongue tumors and are mostly
+assigned to component A. Instead, tumors of the palate (C05) are to
+similar degree assigned to components A and B, which is consistent with
+the anatomical location and the observation that the LNL involvement is
+in between oropharynx and oral cavity-type patterns
+(figure~\ref{fig-prevalence-by-subsite}).
 
@@
 
-We demonstrate the methodology for a mixture model with \(M=2\)
-components, considering the ipsilateral involvement of LNLs I, II, III,
-and IV and the primary tumor subsites shown in
-figure~\ref{fig-prevalence-by-subsite}. In
-figure~\ref{fig-cluster-assignments} we show the resulting mixture
-coefficients \(\pi_{sm}\). The interpretation of this result is as
-follows: Tumors of the base of tongue (C01) are fully described by
-component A, and tumors of the gum (C03) are fully described by
-component B. These two subsites are the most distinct regarding the
-involvement of LNLs I and II, and the result is thus intuitive.
-Component A may be interpreted as a model for oropharynx-like tumor
-spread, and component B as a model for oral cavity-like tumor spread.
-All other subsites are described as mixtures. Tumors in the tonsil (C09)
-have LNL involvement similar to base of tongue tumors and are mostly
-assigned to component A. Instead, tumors of the palate (C05) are to
-similar degree assigned to components A and B, which is consistent with
-the anatomical location and the observation that the LNL involvement is
-in between oropharynx and oral cavity-type patterns.
+The figure~\ref{fig-prevalence-comparison} illustrates the model's
+predictions of the overall prevalence of lymph node involvement in LNLs
+I, II and III (filled histograms) for selected subsites, obtained by
+summing the probabilities of states where the respecive level is
+involved. The mixture model is compared to two independent HMM models
+trained for oral cavity and oropharynx (by pooling the respective
+subsites). The mixture model and the independent oropharynx model
+perform similarly for tonsil tumors (C09), which is the largest patient
+group, dominating the independent oropharynx model and the component B
+in the mixture model. However, the mixture model better predicts the
+higher prevalence in level II for base of tongue and palate tumors.
 
@@
 histograms), and predicted by the mixture model (filled histograms).
-Each row correpsonds to one subsite and each column to the predicted or
-observed prevalence in one LNL.}
+Each row correpsonds to one subsite and each column to one LNL.}
 

### 2024-01-28 scripts/plot_component_assignments.py

@@ -83,13 +83,13 @@ def main():
             }
         )
 
-    bottom_ax.set_xlabel("assignment to component B")
+    bottom_ax.set_xlabel("assignment to component A")
     bottom_ax.xaxis.set_major_formatter(StrMethodFormatter("{x:.0%}"))
     top_ax = bottom_ax.secondary_xaxis(
         location="top",
         functions=(lambda x: 1. - x, lambda x: 1. - x),
     )
-    top_ax.set_xlabel("assignment to component A")
+    top_ax.set_xlabel("assignment to component B")
     top_ax.xaxis.set_major_formatter(StrMethodFormatter("{x:.0%}"))
     bottom_ax.set_yticks([])
     bottom_ax.grid(axis="x", alpha=0.5, color=USZ["gray"], linestyle=":")

### 2024-01-28 content/manuscript.tex
sanitize graphics files in tex for overleaf

@@
 
-\includegraphics{manuscript_files/mediabag/../static/schematic_and_graph.pdf}
+\includegraphics{manuscript_files/static/schematic_and_graph.pdf}
 
@@
 
-\includegraphics{manuscript_files/mediabag/../figures/prevalence_by_subsite.pdf}
+\includegraphics{manuscript_files/figures/prevalence_by_subsite.pdf}
 
@@
 
-\includegraphics{manuscript_files/mediabag/../figures/cluster_assignments.pdf}
+\includegraphics{manuscript_files/figures/cluster_assignments.pdf}
 
@@
 
-\includegraphics{manuscript_files/mediabag/../figures/prevalence_comparison.pdf}
+\includegraphics{manuscript_files/figures/prevalence_comparison.pdf}
 

### 2024-01-28 content/manuscript.tex
Updates from Overleaf

@@
 ]{Jean-Marc Hoffmann}
+%\author[%
+%4,5%
+%]{Sandrine Werlen}
 \author[%
@@
 ]{Matthias Dettmer}
+%\author[%
+%9%
+%]{Philippe Zrounba}
 \author[%
@@
 predict the lymphatic tumor progression in oropharyngeal squamous cell
-carcinomas. To extend the model to other tumor subsites in the head and
-neck defined by ICD-10 codes, we develop a mixture model combining
-multiple HMMs. The mixture coefficients and the model parameters are
-learned via an EM-like algorithm from a large multi-centric dataset on
-lymph node involvement. The methodology is demonstrated for tumors in
-the oropharynx and oral cavity. The mixture model groups anatomically
-nearly subsites and yields interpretable mixture coefficients consistent
-with anatomical location. It allows the prediction of differences in
-lymph node involvement depending on tumor subsite.
+carcinomas. To extend the model to other tumor subsites in the head and neck
+defined by ICD-10 codes, we develop a mixture model combining multiple
+HMMs.
+The mixture coefficients and the model parameters are learned via an EM-like algorithm from a large multi-centric dataset on lymph node involvement.
+The methodology is demonstrated for tumors in the oropharynx and oral cavity.
+The mixture model groups anatomically nearly subsites and yields interpretable mixture coefficients consistent with anatomical location.
+It allows the prediction of
+differences in lymph node involvement depending on tumor subsite.
 \end{customabstract}
@@
 diagnostic imaging modalities are unable to detect microscopic lymph
-node metastases \autocite{snyder_petct_2021,strohl_petct_2021}. To avoid
-nodal recurrences, large volumes in the neck, which are at risk of
-harbouring occult disease, are irradiated electively. Guidelines about
-which lymph node levels (LNLs) to irradiate
+node metastases \autocite{snyder_petct_2021,strohl_petct_2021}. 
+To avoid nodal recurrences 
+%\autocite{ho_decision_2014}
+, large volumes in the neck, which are at risk of harbouring occult disease are irradiated electively. Guidelines about which lymph node levels (LNLs) to irradiate
 \autocite{biau_selection_2019} are currently not based on a patient's
@@
 \autocite{lindberg_distribution_1972,woolgar_histological_1999}.
-
 To personalize the prediction of the risk for occult disease, given a
@@
   an interpretable hidden Markov model (HMM), trained with this data, to
-  predict the risk for occult nodal disease, given an individual
-  patient's diagnosis \autocite{ludwig_hidden_2021}.
+  predict the risk for occult nodal disease, given an individual patient's
+  diagnosis \autocite{ludwig_hidden_2021}.
 \end{enumerate}
@@
 reduce the elective clinical target volume (CTV-N) and thus reduce
-side-effects that degrade the patient's quality of life
+side-effects that degrade the patient's quality of life without compromising treatment efficacy
 \autocite{batth_practical_2014}.
@@
 this could be handled by training different models for broader
-categories of tumor locations, e.g.~oropharynx and oral cavity tumors.
+categories of tumor locations, e.g.~oropharynx and oral cavity tumours.
 However, this approach does not describe differences in lymphatic spread
@@
 intuition is that the lymphatic spread of a tumor that lies anatomically
-at the boarder of oropharynx and oral cavity (e.g.~tumors in the palate)
-may be described by a mixture of different models.
+at the boarder of oropharynx and oral cavity (e.g.~tumours in the
+palate) may be described by a mixture of different models.
 
@@
 Each LNL \(v \in \{ 1, 2, \ldots, V\}\) considered in our model is
-represented by a binary random variable \(X_v\) representing the true
-state of that level (0 for ``healthy'' and 1 for ``involved''). A
+represented by a binary random variable \(X_v\) representing the
+true state of that level (0 for ``healthy'' and 1 for ``involved''). A
 patient's state of lymph node involvement can be represented in a random
@@
 In the above equation, the term
-\(P \left( \mathbf{Y} \mid \mathbf{X} \right)\) is given by the
-sensitivity and specificity of the diagnostic procedure. The term
-\(P \left( \mathbf{X} \right)\) represents the prior probability of
-involvement, which depends on the probability of the tumor to spread
-through the lymphatic system. The main task of the HMM is to model
-\(P \left( \mathbf{X} \right)\) and the main contribution of this paper
-is to incorporate the primary tumor subsite into the model of
+\(P \left( \mathbf{Y} \mid \mathbf{X} \right)\) is given by the sensitivity and specificity of the diagnostic procedure.
+The term \(P \left( \mathbf{X} \right)\) represents the prior probability of involvement, which depends on the probability of the tumor to spread through the lymphatic system.
+The main task of the HMM is to model
+\(P \left( \mathbf{X} \right)\) and the main contribution of this paper is to incorporate the primary tumor subsite into the model for
 \(P \left( \mathbf{X} \right)\).
@@
 consider ipsilateral LNLs I, II, III and IV, which amounts to 16
-possible states. The HMM is specified by a \emph{transition matrix}:
-
+possible states. The HMM is specified by a \emph{transition matrix}
 \begin{equation}\phantomsection\label{eq-transition-matrix}{
@@
 
-whose elements \(A_{ij}\) contain the conditional probabilities that a
-state \(\mathbf{X}[t]=\boldsymbol{\xi}_i\) transitions to
+whose elements \(A_{ij}\) contains the conditional probabilities that a state
+\(\mathbf{X}[t]=\boldsymbol{\xi}_i\) transitions to
 \(\mathbf{X}[t+1]=\boldsymbol{\xi}_j\) over one time step. The
@@
 shown in figure~\ref{fig-schematic-and-graph}. The red arcs in the graph
-of figure~\ref{fig-schematic-and-graph} (right panel) are associated
-with the probability that the primary tumor spreads directly to a LNL
-(parameters \(b_v\)). The blue arcs describe the spread from an upstream
-LNL -- given it is already metastatic -- to a downstream level
-(parameters \(t_{v \rightarrow v+1}\)).
+of figure~\ref{fig-schematic-and-graph}(right panel) are associated with the probability that the primary tumor spreads directly to a LNL (parameters \(b_v\)).
+The blue arcs describe the spread from an upstream LNL -- given it is
+already metastatic -- to a downstream level (parameters
+\(t_{v \rightarrow v+1}\)). \\
 
 Now, let \(\boldsymbol{\pi}\) be the \emph{starting distribution}
-
 \begin{equation}\phantomsection\label{eq-starting-distribution}{
@@
 denoting the probability to start in state \(\boldsymbol{\xi}_i\) at
-time step 0. Assuming that every patient started with all LNLs being
-healthy, we set \(\pi_i\) to zero for all states except the completly
-healthy state
+time step 0. Assuming that every patient started with all
+LNLs being healthy, we set \(\pi_i\) to zero for all states except the
+completly healthy state
 \(\boldsymbol{\xi} = \begin{pmatrix} 0, 0, 0, 0 \end{pmatrix}\), which
 has probability one.
-
 Using the quantities introduced so far, the probability
 \(P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right)\) to be in state
-\(\boldsymbol{\theta}_i\) in time step \(t\) can now be conveniently
-expressed as a matrix product:
-
+\(\boldsymbol{\xi}_i\) in time step \(t\) can now be conveniently expressed as a matrix product:
 \begin{equation}\phantomsection\label{eq-evolution}{
@@
 \(P_T(t)\). This finally defines the probability distribution over all
-states of lymph node involvement used in equation~\ref{eq-bayes-law}.
+states of lymph node involvement used in equation~(\ref{eq-bayes-law}).
 
@@
 
-where \(\boldsymbol{\theta}=\{ b_v, t_{v \rightarrow v+1} \}\) denotes
-the set of all model parameters (7 in our case). Fortunately, the exact
-length and shape of this distribution has little impact as previously
-shown \autocite{ludwig_hidden_2021}. We set \(t_\text{max}=\) 10 and
-\(P_\text{early}(t)\) to a binomial distribution with parameter 0.3.
-Further details on the HMM can be found in
-\textcite{ludwig_hidden_2021}.
+where \(\boldsymbol{\theta}=\{ b_v, t_{v \rightarrow v+1} \}\) denotes the
+set of all model parameters (7 in our case). Fortunately, the exact
+length and shape of this distribution has little impact as
+previously shown \autocite{ludwig_hidden_2021}. We set \(t_\text{max}=\) 10 and \(P_\text{early}(t)\)
+to a binomial distribution with parameter 0.3. Further details on the
+HMM can be found in \autocite{ludwig_hidden_2021}.
 
@@
 \centering{
-
 \includegraphics{manuscript_files/static/schematic_and_graph.pdf}
-
 }
 
-\caption{\label{fig-schematic-and-graph}On the left: Anatomical sketch
-of the tumor subsites and correspsonding ICD-10 codes considered in this
-work. The subsite ``other parts of mouth'' (C06) was not drawn. On the
-right: Parametrized graphical model of the lymphatic network considering
-four LNLs. Blue nodes represent the hidden states of LNLs \(X_v\), while
-the red one is the tumor. Arcs represent possible routes of metastatic
-spread, associated with a probability.}
-
+\caption{\label{fig-schematic-and-graph}On the left: Anatomical sketch of the tumor subsites and correspsonding ICD-10 codes considered in this work. The subsite ``other parts of mouth'' (C06) was
+not drawn. On the right: Parametrized graphical model of the
+lymphatic network considering four LNLs.
+Blue nodes represent the hidden states of LNLs \(X_v\), while the red one is the tumor.
+Arcs represent possible routes of metastatic spread, associated with a probability.}
 \end{figure}%
@@
 via the number of patients \(N_{is}\) that were diagnosed in LNL
-involvement state \(i\) and had a primary tutor in subsite \(s\). Let us
+involvement state \(i\) and had a primary tumor in subsite \(s\). Let us
 further assume that we want to describe this dataset using a mixture of
@@
 written as
-
 \begin{equation}\phantomsection\label{eq-mixture-distribution}{
@@
 \(P \left( \boldsymbol{\theta}, \boldsymbol{\pi} \mid \mathbf{D} \right)\)
-is given by the likelihood in equation~\ref{eq-mixture-distribution}
-except for a normalisation constant. In this work, we use Markov chain
-Monte Carlo sampling (MCMC) via the \texttt{emcee} Python package
+is given by the likelihood in equation~(\ref{eq-mixture-distribution}) except
+for a normalisation constant. In this work, we use Markov chain Monte
+Carlo sampling (MCMC) via the \texttt{emcee} Python package
 \autocite{foreman-mackey_emcee_2013} to sample model parameters from the
@@
 different models. To address this problem, we revert to an
-\emph{expectation-maximization (EM)} algorithm where we iterate two
-steps until convergence of the mixing coefficients. In the E-step, we
-sample model parameters \(\boldsymbol{\theta_m}\) using MCMC for given
-mixing coefficients \(\pi_{sm}\). In the M-step, we maximize the
-likelihood with respect to the mixing coefficients for given samples of
-\(\boldsymbol{\theta_m}\)
+\emph{expectation-maximization (EM)} algorithm where we iterate two steps until convergence of the mixing coefficients.
+In the E-step, we sample model parameters \(\boldsymbol{\theta_m}\) using MCMC for given mixing coefficients \(\pi_{sm}\).
+In the M-step, we maximize the likelihood with respect to the mixing coefficients for given samples of \(\boldsymbol{\theta_m}\).
 
@@
 
-For the analyses in this work, we used five datasets from four different
-institutions, resulting in 1242 patients in total.
+For the analyses in this work, we used five datasets from four different institutions, resulting in 1242 patients in total.
 
@@
 \item
-  162 oropharyngeal patients from the Hospital Vall d'Hebron in Spain
-  (not yet public)
+  162 oropharyngeal patients from the Hospital Vall d'Hebron in Spain (not yet public)
 \end{enumerate}
@@
 \centering{
-
 \includegraphics{manuscript_files/figures/prevalence_by_subsite.pdf}
-
 }
 
-\caption{\label{fig-prevalence-by-subsite}Prevalence of ipsilateral LNL
-involvement stratified by subsite. The subsites are sorted in ascending
-order by their prevalence of involvement in LNL II. Oral cavity subsites
-are plotted in shades of blue, oropharynx subsites in shades of orange.}
+\caption{\label{fig-prevalence-by-subsite}Prevalence of ipsilateral LNL involvement stratified by subsite.
+The subsites are sorted in ascending order by their prevalence of involvement in LNL II.
+Oral cavity subsites are
+plotted in shades of blue, oropharynx subsites in shades of orange.}
 
@@
 interactively explored in our \textbf{Ly}mphatic \textbf{Pro}gression
-e\textbf{X}plorer \href{https://lyprox.org}{LyProX}. For each patient,
-the primary tumor subsite is reported (among other patient and tumor
-characteristics) and each individual LNL is reported as metastatic or
-healthy, according to the available diagnostic modalities (in part
-pathology after neck dissection, otherwise clinical involvement).
+e\textbf{X}plorer \href{https://lyprox.org}{LyProX}. For each patient, the primary tumor
+subsite is reported (among other patient and tumor characteristics) and
+each individual LNL is reported as metastatic or healthy, according to
+the available diagnostic modalities (in part pathology after neck
+dissection, otherwise clinical involvement).
 
-In figure~\ref{fig-prevalence-by-subsite}, we plot the prevalence of
-involvement in the four ipsilateral LNLs I, II, III, and IV stratified
-by the primary tumor's subsite. The figure illustrates the variations in
-LNL involvement between subsites within the oral cavity and oropharynx
-categories. The number of patients for each subsite is indicated in
-figure~\ref{fig-cluster-assignments}.
+In figure~\ref{fig-prevalence-by-subsite}, we plot the
+prevalence of involvement in the four ipsilateral LNLs I, II, III, and
+IV stratified by primary tumor subsite. 
+%We included patients with tumors in the oral cavity, gum (C03), floor of mouth (C04), other/unspecified parts of the mouth (C06), other/unspecified parts of tongue (C02), palate (C05), and tumors in the oropharynx (C10), tonsil (C09), base of tongue (C01),
+The figure illustrates the variations in LNL involvement between subsites within
+the oral cavity and oropharynx categories. The number of patients for each subsite is indicated in figure \ref{fig-cluster-assignments}.
 
@@
 coefficients \(\pi_{sm}\). The interpretation of this result is as
-follows: tumors of the base of tongue (C01) are fully described by
-component A, and tumors of the gum (C03) are fully described by
-component B. These two subsites are the most distinct regarding the
+follows: Tumors of the base of tongue (C01) are fully described by
+component B, and tumors of the gum (C03) are fully described by
+component A. These two subsites are the most distinct regarding the
 involvement of LNLs I and II, and the result is thus intuitive.
-Component A may be interpreted as a model for oropharynx-like tumor
-spread, and component B as a model for oral cavity-like tumor spread.
-All other subsites are described as mixtures. tumors in the tonsil (C09)
+Component B may be interpreted as a model for oropharynx-like tumor
+spread, and component A as a model for oral cavity-like tumor spread.
+All other subsites are described as mixtures. Tumors in the tonsil (C09)
 have LNL involvement similar to base of tongue tumors and are mostly
-assigned to component A. Instead, tumors of the palate (C05) are to
+assigned to component B. Instead, tumors of the palate (C05) are to
 similar degree assigned to components A and B, which is consistent with
 the anatomical location and the observation that the LNL involvement is
-in between oropharynx and oral cavity-type patterns
-(figure~\ref{fig-prevalence-by-subsite}).
+in between oropharynx and oral cavity-type patterns (figure~\ref{fig-prevalence-by-subsite}).
 
@@
 \centering{
-
 \includegraphics{manuscript_files/figures/cluster_assignments.pdf}
-
 }
@@
 size of the marker (area) corresponds to the number of patients in the
-subsite.}
+subsite indicated in brackets. The approximate locations of subsites are skeched in Figure \ref{fig-schematic-and-graph}.}
 
@@
 
-The figure~\ref{fig-prevalence-comparison} illustrates the model's
-predictions of the overall prevalence of lymph node involvement in LNLs
-I, II and III (filled histograms) for selected subsites, obtained by
-summing the probabilities of states where the respecive level is
-involved. The mixture model is compared to two independent HMM models
-trained for oral cavity and oropharynx (by pooling the respective
-subsites). The mixture model and the independent oropharynx model
-perform similarly for tonsil tumors (C09), which is the largest patient
-group, dominating the independent oropharynx model and the component B
-in the mixture model. However, the mixture model better predicts the
-higher prevalence in level II for base of tongue and palate tumors.
+Figure \ref{fig-prevalence-comparison} illustrates the model's predictions of the overall prevalence of lymph node involvement in LNLs I, II and III (filled histograms) for selected subsites, obtained by summing the probabilities of states where the respecive level is involved.
+The mixture model is compared to two independent HMM models trained for oral cavity and oropharynx (by pooling the respective subsites).
+The mixture model and the independent oropharynx model perform similarly for tonsil tumors (C09), which is the largest patient group, dominating the independent oropharynx model and the component B in the mixture model.
+However, the mixture model better predicts the higher prevalence in level II for base of tongue and palate tumors.
 
@@
 \centering{
-
 \includegraphics{manuscript_files/figures/prevalence_comparison.pdf}
-
 }

### 2024-04-11 content/manuscript.qmd
incorporate minor changes & add acknowledgements

@@
       - ref: 2
-  - name: Yoel Perez Haas
+  - name: Yoel Pérez Haas
     affiliations:
@@
       - ref: 10
-  - name: Bertrand Poumayou
+  - name: Bertrand Pouymayou
     orcid: 0000-0002-8060-7827
@@
       - text: "\\usepackage{newtxtext,newtxmath}"
+    link-citations: true
   html:
@@
 
-Such a personalized risk prediction may allow clinicians to safely reduce the elective clinical target volume (CTV-N) and thus reduce side-effects that degrade the patient's quality of life [@batth_practical_2014].
+Such a personalized risk prediction may allow clinicians to safely reduce the elective clinical target volume (CTV-N) and thus reduce side-effects that degrade the patient's quality of life without compromising treatment efficacy [@batth_practical_2014].
 
@@
 tumors in the palate) may be described by a mixture of different models.
+Tumor subsites used in this work are sketched in @fig-schematic-and-graph.
 
@@
 We set $t_\text{max}=$ {{< var model.max_t >}} and $P_\text{early}(t)$ to a binomial distribution with parameter {{< var model.first_binom_prob >}}.
-Further details on the HMM can be found in @ludwig_hidden_2021.
+Further details on the HMM can be found in @ludwig_hidden_2021 and @zora231470.
 
-![On the left: Anatomical sketch of the tumor subsites and correspsonding ICD-10 codes considered in this work.
+![On the left: Anatomical sketch of the tumor subsites and corresponding ICD-10 codes considered in this work.
 The subsite "other parts of mouth" (C06) was not drawn.
@@
 Future work will extend the work to tumors in the hypopharynx and larynx and optimize the number of model components.
+
+# Acknowledgements {.unnumbered}
+
+This work was supported by the clinical research priority program (CRPP) "Artificial intelligence in oncological imaging" of the University of Zurich and by the Swiss cancer research foundation under 
+grant KFS 5645–08–2022.

### 2024-04-11 content/manuscript.tex
incorporate minor changes & add acknowledgements

--- 2024-01-28 content/manuscript.tex
@@
 1,2%
-]{Yoel Perez Haas}
+]{Yoel Pérez Haas}
 \author[%
@@
 2%
-]{Bertrand Poumayou}
+]{Bertrand Pouymayou}
 \author[%
@@
 reduce the elective clinical target volume (CTV-N) and thus reduce
-side-effects that degrade the patient's quality of life
-\autocite{batth_practical_2014}.
+side-effects that degrade the patient's quality of life without
+compromising treatment efficacy \autocite{batth_practical_2014}.
 
@@
 at the boarder of oropharynx and oral cavity (e.g.~tumors in the palate)
-may be described by a mixture of different models.
+may be described by a mixture of different models. Tumor subsites used
+in this work are sketched in figure~\ref{fig-schematic-and-graph}.
 
@@
 \(P_\text{early}(t)\) to a binomial distribution with parameter 0.3.
-Further details on the HMM can be found in
-\textcite{ludwig_hidden_2021}.
+Further details on the HMM can be found in \textcite{ludwig_hidden_2021}
+and \textcite{zora231470}.
 
@@
 \caption{\label{fig-schematic-and-graph}On the left: Anatomical sketch
-of the tumor subsites and correspsonding ICD-10 codes considered in this
+of the tumor subsites and corresponding ICD-10 codes considered in this
 work. The subsite ``other parts of mouth'' (C06) was not drawn. On the
@@
 
+\section*{Acknowledgements}\label{acknowledgements}
+\addcontentsline{toc}{section}{Acknowledgements}
+
+This work was supported by the clinical research priority program (CRPP)
+``Artificial intelligence in oncological imaging'' of the University of
+Zurich and by the Swiss cancer research foundation under grant KFS
+5645--08--2022.
+
 \printbibliography

### 2024-06-27 content/manuscript.qmd
use png over svg in schematic

@@
 Blue nodes represent the hidden states of LNLs $X_v$, while the red one is the tumor.
-Arcs represent possible routes of metastatic spread, associated with a probability.](../static/schematic_and_graph.svg){#fig-schematic-and-graph fig-pos="th"}
+Arcs represent possible routes of metastatic spread, associated with a probability.](../static/schematic_and_graph.png){#fig-schematic-and-graph fig-pos="th"}
 
